---
title: "Figure S4: Hydrology — Drainage, Texture, and N Leaching"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 10
    fig-height: 5
execute:
  warning: false
  message: false
  echo: true
---

# Caption

**Figure S4.** Relationships between soil texture, annual drainage, and
cumulative N leaching across treatment plots in 2023 and 2024. **(a)** Annual
drainage (mm) vs. cumulative leached N (kg N ha⁻¹); **(b)** sand content (%)
vs. annual drainage (mm). Lines show per-treatment OLS regression fits with
95% confidence bands. Point shapes distinguish treatments. Equation and R²
annotations are shown per treatment.

# Setup

```{r setup}
library(tidyverse)
library(readxl)
library(patchwork)
library(lme4)
library(lmerTest)   # p-values for lmer
library(broom.mixed)
conflicted::conflicts_prefer(lmerTest::lmer)

# ── Publication theme (matches figure_1_weather.qmd) ──────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(2, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "top",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}

# ── Colors & shapes ────────────────────────────────────────────────────────────
supp_colors <- c(
  "Maize"         = "#B0B0B0",
  "Soybean"       = "#A8C8E8",
  "Sorghum"       = "#D8D97AFF",
  "Sorghum + WCC" = "#95C36EFF"
)
supp_shapes <- c("Maize" = 22, "Soybean" = 24,
                 "Sorghum" = 21, "Sorghum + WCC" = 23)
```

# Data

```{r data}
plot_area_ha <- (120 * 160) / 107639.1041671

# ── Leaching & drainage ────────────────────────────────────────────────────────
annual_leaching <- read_csv("data/clean_leaching.csv", show_col_types = FALSE) |>
  janitor::clean_names() |>
  mutate(date = as.Date(date), year = lubridate::year(date)) |>
  filter(year %in% c(2023, 2024)) |>
  mutate(plot = as.numeric(plot)) |>
  group_by(plot, treatment, year) |>
  summarise(
    annual_cumulative_leaching = (max(cumulative_n_loss_mg, na.rm = TRUE) / 1e6) / plot_area_ha,
    annual_drainage_l          = max(cumulative_flow_l, na.rm = TRUE),
    annual_drainage_mm         = (annual_drainage_l / plot_area_ha) / 10000,
    .groups = "drop"
  ) |>
  filter(is.finite(annual_cumulative_leaching), annual_cumulative_leaching >= 0)

# ── Soil texture ───────────────────────────────────────────────────────────────
texture <- read_excel(
  "data/PSA Report - AGRON-McDaniel-Potter-2024.xlsm",
  sheet = "Main Data File format", n_max = 22
) |>
  select(Sample, Sand = S, Silt = T, Clay = C) |>
  separate(Sample, into = c("ID", "Plot"), sep = "_") |>
  mutate(across(c(Plot, Sand, Silt, Clay), as.numeric)) |>
  drop_na() |>
  select(Plot, Sand, Silt, Clay)

# ── Join & relabel ─────────────────────────────────────────────────────────────
plot_hydrology <- annual_leaching |>
  select(year, plot, treatment,
         drainage_mm    = annual_drainage_mm,
         leaching_kg_ha = annual_cumulative_leaching) |>
  left_join(texture, by = c("plot" = "Plot")) |>
  mutate(
    year      = as.factor(year),
    Treatment = case_when(
      treatment == "Corn"          ~ "Maize",
      treatment == "Soy"           ~ "Soybean",
      treatment == "Sorghum + Rye" ~ "Sorghum + WCC",
      TRUE                         ~ treatment
    ),
    Treatment = factor(Treatment,
                       levels = c("Maize", "Soybean", "Sorghum", "Sorghum + WCC"))
  )
```

# Sample count diagnostic

```{r sample-counts}
# Note: Maize and Soybean plots rotate between years (crop rotation design).
# Sorghum and Sorghum + WCC plots are stable across years.
plot_hydrology |>
  count(year, Treatment, name = "n_plots") |>
  arrange(year, Treatment) |>
  print()
```

# Figure S4

```{r lm-labels-helper}
# Per-treatment OLS label: slope, intercept, R²
lm_labels <- function(df, xvar, yvar) {
  df |>
    group_by(Treatment) |>
    summarise(
      lbl = {
        mod <- lm(get(yvar) ~ get(xvar))
        b0  <- round(coef(mod)[1], 2)
        b1  <- round(coef(mod)[2], 3)
        r2  <- round(summary(mod)$r.squared, 2)
        paste0("β = ", b1, ",  R² = ", r2)
      },
      .groups = "drop"
    )
}
```

```{r figure-s4}
eq_loss <- lm_labels(plot_hydrology, "drainage_mm", "leaching_kg_ha")
eq_tex  <- lm_labels(plot_hydrology, "Sand", "drainage_mm")

# Panel (a): Drainage vs. N Leaching ──────────────────────────────────────────
p_loss <- ggplot(plot_hydrology,
                 aes(x = drainage_mm, y = leaching_kg_ha)) +
  geom_smooth(aes(color = Treatment), method = "lm", se = FALSE,
              linewidth = 0.8) +
  geom_point(aes(fill = Treatment, shape = Treatment),
             size = 3.5, alpha = 0.9, color = "black", stroke = 0.4) +
  geom_text(
    data = eq_loss,
    aes(label = lbl, color = Treatment),
    x = -Inf, y = Inf,
    hjust = -0.05,
    vjust = seq(1.6, by = 1.4, length.out = nrow(eq_loss)),
    size = 2.8, fontface = "italic", show.legend = FALSE
  ) +
  scale_fill_manual(values  = supp_colors) +
  scale_color_manual(values = supp_colors) +
  scale_shape_manual(values = supp_shapes) +
  labs(
    tag   = "(a)",
    x     = "Annual Drainage (mm)",
    y     = expression("Cumulative Leached N (kg N ha"^{-1}*")"),
    fill  = NULL, shape = NULL, color = NULL
  ) +
  theme_publication(base_size = 10) +
  theme(
    plot.tag          = element_text(face = "bold", size = 11),
    plot.tag.position = "topleft"
  )

# Panel (b): Sand content vs. Drainage ────────────────────────────────────────
p_texture <- ggplot(plot_hydrology,
                    aes(x = Sand, y = drainage_mm)) +
  geom_smooth(aes(color = Treatment), method = "lm", se = FALSE,
              linewidth = 0.8) +
  geom_point(aes(fill = Treatment, shape = Treatment),
             size = 3.5, alpha = 0.9, color = "black", stroke = 0.4) +
  geom_text(
    data = eq_tex,
    aes(label = lbl, color = Treatment),
    x = -Inf, y = Inf,
    hjust = -0.05,
    vjust = seq(1.6, by = 1.4, length.out = nrow(eq_tex)),
    size = 2.8, fontface = "italic", show.legend = FALSE
  ) +
  scale_fill_manual(values  = supp_colors) +
  scale_color_manual(values = supp_colors) +
  scale_shape_manual(values = supp_shapes) +
  labs(
    tag   = "(b)",
    x     = "Sand Content (%)",
    y     = "Annual Drainage (mm)",
    fill  = NULL, shape = NULL, color = NULL
  ) +
  theme_publication(base_size = 10) +
  theme(
    plot.tag          = element_text(face = "bold", size = 11),
    plot.tag.position = "topleft"
  )

figure_s4 <- (p_loss + p_texture) +
  plot_layout(widths = c(1, 1), guides = "collect") &
  theme(
    legend.position      = "top",
    legend.justification = "center",
    plot.margin          = margin(t = 2, r = 4, b = 2, l = 4, unit = "mm")
  )
figure_s4
```

```{r save}
ggsave(
  "SubmissionFiguresTables/FigureS4_Hydrology.tiff",
  plot        = figure_s4,
  width       = 174, height = 100, units = "mm", dpi = 600, bg = "white",
  compression = "lzw"
)
```

# Per-treatment OLS: Drainage → N Leaching

Simple linear models fit within each treatment group. With n = 7–8 per
treatment, these are exploratory; interpret slopes and R² directionally.

```{r per-treatment-ols}
ols_results <- plot_hydrology |>
  group_by(Treatment) |>
  summarise(
    n         = n(),
    slope     = coef(lm(leaching_kg_ha ~ drainage_mm))[2],
    intercept = coef(lm(leaching_kg_ha ~ drainage_mm))[1],
    r_squared = summary(lm(leaching_kg_ha ~ drainage_mm))$r.squared,
    p_slope   = summary(lm(leaching_kg_ha ~ drainage_mm))$coefficients[2, 4],
    .groups   = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

print(ols_results)
```

# Linear Mixed Effects Model: Do slopes differ by treatment?

**Design note:** Maize and Soybean plots rotate between years (crop rotation);
Sorghum/Sorghum+WCC plots are fixed. The `(1 | plot)` random intercept
captures plot-level spatial effects (soil, drainage position) that persist
across years regardless of treatment.

**Model:** `leaching_kg_ha ~ drainage_mm × Treatment + (1 | plot)`

The `drainage_mm:Treatment` interaction tests whether the slope of N leaching
on drainage differs among treatments (i.e., does each treatment have a
distinct drainage-to-leaching efficiency?).

```{r lmer-model}
# Fit full model with treatment × drainage interaction
m_full <- lmer(
  leaching_kg_ha ~ drainage_mm * Treatment + (1 | plot),
  data = plot_hydrology,
  REML = FALSE   # use ML for model comparison
)

# Reduced model (parallel slopes — no interaction)
m_reduced <- lmer(
  leaching_kg_ha ~ drainage_mm + Treatment + (1 | plot),
  data = plot_hydrology,
  REML = FALSE
)

cat("=== Likelihood ratio test: interaction vs. no interaction ===\n")
print(anova(m_reduced, m_full))

cat("\n=== Full model fixed effects (lmerTest p-values) ===\n")
m_full_reml <- lmer(
  leaching_kg_ha ~ drainage_mm * Treatment + (1 | plot),
  data = plot_hydrology,
  REML = TRUE   # REML for final estimates
)
print(summary(m_full_reml)$coefficients |> round(4))
```

```{r lmer-slopes}
# Extract per-treatment slopes from the full model
# Slope for reference (Maize) = drainage_mm coefficient
# Slope for others = drainage_mm + drainage_mm:TreatmentX interaction
fe <- fixef(m_full_reml)
ci <- confint(m_full_reml, parm = "beta_", method = "Wald")

trt_levels <- levels(plot_hydrology$Treatment)
ref_trt    <- trt_levels[1]   # Maize is reference

slopes <- tibble(
  Treatment  = trt_levels,
  slope      = c(
    fe["drainage_mm"],
    fe["drainage_mm"] + fe[paste0("drainage_mm:Treatment", trt_levels[-1])]
  ),
  ci_lo = c(
    ci["drainage_mm", 1],
    ci["drainage_mm", 1] + ci[paste0("drainage_mm:Treatment", trt_levels[-1]), 1]
  ),
  ci_hi = c(
    ci["drainage_mm", 2],
    ci["drainage_mm", 2] + ci[paste0("drainage_mm:Treatment", trt_levels[-1]), 2]
  )
)

cat("\n=== Model-estimated slopes (kg N ha⁻¹ per mm drainage) by treatment ===\n")
print(slopes |> mutate(across(where(is.numeric), \(x) round(x, 4))))
```

```{r slope-plot, fig.height=4}
# Coefficient plot: per-treatment slopes with 95% Wald CI
ggplot(slopes, aes(x = slope, y = Treatment, color = Treatment)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                 height = 0.25, linewidth = 0.8) +
  geom_point(aes(fill = Treatment, shape = Treatment),
             size = 4, color = "black", stroke = 0.5) +
  scale_color_manual(values = supp_colors) +
  scale_fill_manual(values  = supp_colors) +
  scale_shape_manual(values = supp_shapes) +
  labs(
    x     = expression("Slope (kg N ha"^{-1}*" per mm drainage)"),
    y     = "",
    title = "Mixed model drainage → leaching slopes by treatment",
    subtitle = "Points = REML estimate; bars = 95% Wald CI"
  ) +
  theme_publication(base_size = 11) +
  theme(legend.position = "none")
```
