---
title: "Figure 4: Nitrous Oxide Emissions"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
  echo: true
---

# Scientific Context

Nitrous oxide (N~2~O) is a potent greenhouse gas produced during soil nitrogen
transformations. We measured N~2~O fluxes using static chambers throughout the
growing season to quantify emissions from sorghum systems with and without
winter cover crops. Fluxes were generally low and episodic, with peak emissions
often following fertilization or precipitation events.

The figure has six panels arranged in a 2 x 3 grid:

- **(a--b)** Daily N~2~O flux with precipitation overlay (2023, 2024)
- **(c)** Daily treatment difference boxplot (WCC minus Sorghum) by year
- **(d--e)** Cumulative N~2~O flux (2023, 2024)
- **(f)** Annual cumulative N~2~O by treatment (years combined)

# Setup

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(patchwork)
library(lubridate)
library(emmeans)
library(multcomp)

# ── Publication theme ──────────────────────────────────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(3, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 0, hjust = 0.5, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "none",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}

treatment_colors <- c(
  "Sorghum"       = "#D8D97AFF",
  "Sorghum + Rye" = "#95C36EFF",
  "Sorghum + WCC" = "#95C36EFF"
)
```

# Statistical Approach

**Current methods:**

1. **Per-date Wilcoxon rank-sum tests** (*p* < 0.05) for asterisks on dates
   with significant treatment differences in daily N~2~O flux. Plot-level
   means are used as replicates (averaging row/interrow within a plot first).
2. **One-sample t-test** of daily differences (WCC minus Sorghum) against zero
   for the boxplot panel (c).
3. **One-way ANOVA** with Tukey-adjusted comparisons on annual cumulative
   N~2~O (pooling years) for panel (f).

**Assessment:**

- **Per-date Wilcoxon tests:** N~2~O fluxes are notoriously zero-inflated and
  right-skewed (many near-zero measurements, occasional spikes). The Wilcoxon
  test handles non-normality but with n = 4 per treatment the test has
  extremely low power; only perfect separation can achieve *p* < 0.05. The
  asterisks should be interpreted cautiously.

- **Cumulative calculation via linear interpolation:** Interpolating between
  measurement days is standard for chamber-based N~2~O studies, but it can
  miss short-lived emission pulses between sampling events. This is a known
  limitation of manual chamber methods.

- **Annual ANOVA (panel f):** Same concern as Figure 3 — pooling years
  without accounting for year structure.

**Suggested alternatives:**

- **Generalized linear mixed model:** For daily fluxes, a GLMM with Gamma or
  Tweedie family and a log link
  (`glmmTMB(flux ~ Treatment + (1|plot) + (1|date), family = Tweedie)`) would
  better handle the skewed, zero-inflated distribution while accounting for
  repeated measures.
- **Trapezoidal integration:** Replace linear interpolation with the
  trapezoidal rule for cumulative N~2~O calculation. In practice the two
  methods give very similar results for irregularly spaced data, but
  trapezoidal is the more standard approach in the soil science literature.
- **Panel f:** Same recommendation as Figure 3 — include year as a term.

# Data Processing

```{r load-data}
flux_data <- read_csv("data/cleaned_flux_data.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date), year = year(date))

rain <- read_xlsx("data/rain/historical_rain.xlsx") |>
  select(day, doy, precipmm) |>
  mutate(date = as.Date(day), year = year(date))

fert_events <- tibble(
  year            = c(2023, 2024),
  fertilizer_date = as.Date(c("2023-05-05", "2024-07-17"))
)

# Filter to sorghum treatments
flux_subset <- flux_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    Treatment = factor(Treatment, levels = c("Sorghum", "Sorghum + WCC"))
  ) |>
  filter(year %in% c(2023, 2024))
```

```{r daily-flux}
# Daily means in g N ha⁻¹ d⁻¹
daily_flux <- flux_subset |>
  summarise(
    mean_flux_g = mean(gnha_day_no_negative, na.rm = TRUE),
    se_flux_g   = sd(gnha_day_no_negative, na.rm = TRUE) /
                  sqrt(sum(!is.na(gnha_day_no_negative))),
    .by = c(Treatment, year, date)
  )
```

```{r significance-tests}
# Per-date Wilcoxon tests (plot-level means as replicates)
daily_significance <- flux_subset |>
  summarise(plot_mean_flux = mean(gnha_day_no_negative, na.rm = TRUE),
            .by = c(year, date, Treatment, plot)) |>
  summarise(
    p_value = {
      s <- plot_mean_flux[Treatment == "Sorghum"]       |> na.omit()
      w <- plot_mean_flux[Treatment == "Sorghum + WCC"] |> na.omit()
      if (length(s) >= 3 && length(w) >= 3)
        tryCatch(wilcox.test(s, w, exact = FALSE)$p.value, error = \(e) NA_real_)
      else NA_real_
    },
    .by = c(year, date)
  ) |>
  mutate(significant = !is.na(p_value) & p_value < 0.05)

star_data <- daily_flux |>
  summarise(y_top = max(mean_flux_g + se_flux_g, na.rm = TRUE), .by = c(year, date)) |>
  left_join(daily_significance, by = c("year", "date")) |>
  filter(significant) |>
  mutate(star_y = y_top * 1.1, star_label = "*")
```

```{r cumulative-flux}
# Plot-level cumulative N2O via linear interpolation
daily_flux_plot_level <- flux_subset |>
  summarise(daily_flux_kg = mean(gnha_day_no_negative, na.rm = TRUE) / 1000,
            .by = c(Treatment, year, plot, date))

cum_flux_plot_level <- daily_flux_plot_level |>
  arrange(Treatment, year, plot, date) |>
  group_by(Treatment, year, plot) |>
  group_modify(\(df, keys) {
    df <- df |>
      complete(date = seq(min(date), max(date), by = "day")) |>
      arrange(date)
    non_na <- !is.na(df$daily_flux_kg)
    if (sum(non_na) == 0) return(tibble())

    flux_interp <- if (sum(non_na) >= 2) {
      approx(df$date[non_na], df$daily_flux_kg[non_na],
             xout = df$date, rule = 1, ties = "ordered")$y
    } else {
      rep(df$daily_flux_kg[non_na][1], nrow(df))
    }

    df |>
      mutate(flux_interp = flux_interp, cum_flux_kg = cumsum(flux_interp)) |>
      filter(!is.na(flux_interp))
  }) |>
  ungroup()

cum_flux_summary <- cum_flux_plot_level |>
  summarise(
    cum_flux_mean = mean(cum_flux_kg, na.rm = TRUE),
    cum_flux_sd   = sd(cum_flux_kg, na.rm = TRUE),
    n_plots       = n(),
    cum_flux_se   = cum_flux_sd / sqrt(n_plots),
    t_crit        = if_else(n_plots > 1, qt(0.975, df = n_plots - 1), NA_real_),
    cum_flux_ci   = t_crit * cum_flux_se,
    .by = c(Treatment, year, date)
  )
```

```{r diff-and-annual}
# ── Daily difference (WCC - Sorghum) ──────────────────────────────────────
daily_n2o_diff <- daily_flux |>
  mutate(trt_key = if_else(Treatment == "Sorghum + WCC", "WCC", "Sorghum")) |>
  select(year, date, trt_key, mean_flux_g) |>
  pivot_wider(names_from = trt_key, values_from = mean_flux_g) |>
  mutate(diff_wcc_minus_sorghum = WCC - Sorghum)

one_sample_vs_zero <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, mu = 0)$p.value, error = \(e) NA_real_)
}

n2o_diff_stats <- daily_n2o_diff |>
  summarise(
    n          = sum(!is.na(diff_wcc_minus_sorghum)),
    mean_diff  = mean(diff_wcc_minus_sorghum, na.rm = TRUE),
    p_value    = one_sample_vs_zero(diff_wcc_minus_sorghum),
    significant = !is.na(p_value) & p_value < 0.05,
    letter     = if_else(significant, "a", NA_character_),
    .by = year
  )

# ── Annual cumulative N2O + one-way ANOVA ─────────────────────────────────
annual_n2o <- cum_flux_plot_level |>
  summarise(annual_cum_n2o = max(cum_flux_kg, na.rm = TRUE),
            .by = c(year, plot, Treatment)) |>
  filter(is.finite(annual_cum_n2o), annual_cum_n2o >= 0)

anova_n2o   <- aov(annual_cum_n2o ~ Treatment, data = annual_n2o)
anova_n2o_p <- summary(anova_n2o)[[1]]["Treatment", "Pr(>F)"]

n2o_letters <- tryCatch({
  emmeans(anova_n2o, ~ Treatment) |>
    cld(adjust = "tukey", Letters = letters) |>
    as_tibble() |>
    transmute(Treatment = as.character(Treatment),
              letter    = gsub("\\s+", "", .group))
}, error = \(e) {
  annual_n2o |> distinct(Treatment) |> mutate(letter = "a")
})
```

# Figure 4

```{r figure4}
#| fig-width: 12
#| fig-height: 8
#| fig-dpi: 300

# ── Precipitation trimmed to flux windows ──────────────────────────────────
flux_date_ranges <- daily_flux |>
  summarise(min_date = min(date), max_date = max(date), .by = year)

daily_precip_trimmed <- rain |>
  select(date, year, precipmm) |>
  filter(year %in% c(2023, 2024)) |>
  left_join(flux_date_ranges, by = "year") |>
  filter(date >= min_date, date <= max_date)

# ── Scaling constants ─────────────────────────────────────────────────────
global_max_flux_g     <- max(daily_flux$mean_flux_g + daily_flux$se_flux_g, na.rm = TRUE)
global_max_precip     <- max(daily_precip_trimmed$precipmm, na.rm = TRUE)
global_max_cum_flux   <- max(cum_flux_summary$cum_flux_mean + cum_flux_summary$cum_flux_ci, na.rm = TRUE)
precip_scale_factor   <- global_max_flux_g * 0.4 / global_max_precip
plot_top_flux         <- global_max_flux_g * 1.12

# ── Panel builder: daily N2O ──────────────────────────────────────────────
create_n2o_daily_panel <- function(year_val, is_right, tag_label) {
  flux_year   <- filter(daily_flux, year == year_val)
  precip_year <- filter(daily_precip_trimmed, year == year_val)
  star_year   <- filter(star_data, year == year_val)
  fert_year   <- filter(fert_events, year == year_val)
  x_lims      <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot() +
    geom_rect(
      data = precip_year,
      aes(xmin = date - 0.5, xmax = date + 0.5,
          ymin = plot_top_flux - precipmm * precip_scale_factor,
          ymax = plot_top_flux),
      fill = "#377EB8", alpha = 0.7
    ) +
    geom_point(data = flux_year,
               aes(x = date, y = mean_flux_g, color = Treatment), size = 1.4) +
    geom_errorbar(
      data = flux_year,
      aes(x = date,
          ymin = pmax(mean_flux_g - se_flux_g, 0),
          ymax = mean_flux_g + se_flux_g,
          color = Treatment),
      width = 0.2, alpha = 0.5
    ) +
    geom_vline(data = fert_year, aes(xintercept = fertilizer_date),
               linetype = "dashed", color = "black", alpha = 0.7) +
    geom_text(data = star_year,
              aes(x = date, y = star_y, label = star_label), size = 4.8) +
    scale_color_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right) expression("Daily N"[2] * "O Flux (g N ha"^-1 * " d"^-1 * ")") else "",
      limits = c(0, plot_top_flux),
      sec.axis = sec_axis(
        ~ (plot_top_flux - .) / precip_scale_factor,
        name = if (is_right) "Precipitation (mm)" else ""
      )
    ) +
    labs(x = "Month", tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5),
      axis.ticks.x     = element_line(),
      axis.line.x      = element_line(),
      legend.position  = "none",
      plot.tag          = element_text(face = "bold", size = 11),
      plot.tag.position = "topleft",
      plot.margin       = margin(t = 14, r = 4, b = 4, l = 4)
    )

  if (is_right) {
    p <- p + theme(
      axis.text.y = element_blank(), axis.ticks.y = element_blank(),
      axis.line.y = element_blank(), axis.title.y = element_blank(),
      axis.text.y.right  = element_text(color = "#377EB8"),
      axis.ticks.y.right = element_line(),
      axis.title.y.right = element_text(color = "#377EB8")
    )
  } else {
    p <- p + theme(
      axis.text.y.right = element_blank(), axis.ticks.y.right = element_blank(),
      axis.title.y.right = element_blank()
    )
  }
  p
}

# ── Panel builder: cumulative N2O ─────────────────────────────────────────
create_n2o_cum_panel <- function(year_val, is_right, tag_label) {
  cum_year  <- filter(cum_flux_summary, year == year_val)
  fert_year <- filter(fert_events, year == year_val)
  x_lims    <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot(cum_year, aes(x = date, y = cum_flux_mean,
                             color = Treatment, fill = Treatment)) +
    geom_ribbon(aes(ymin = pmax(cum_flux_mean - cum_flux_ci, 0),
                    ymax = cum_flux_mean + cum_flux_ci),
                alpha = 0.15, linewidth = 0) +
    geom_line(linewidth = 1) +
    geom_vline(data = fert_year, aes(xintercept = fertilizer_date),
               linetype = "dashed", color = "black", alpha = 0.7) +
    scale_color_manual(values = treatment_colors) +
    scale_fill_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right) expression("Cumulative N"[2] * "O Flux (kg N ha"^-1 * ")") else "",
      limits = c(0, global_max_cum_flux)
    ) +
    labs(x = "Month", tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(legend.position = "none",
          plot.tag = element_text(face = "bold", size = 11),
          plot.tag.position = "topleft",
          plot.margin = margin(t = 14, r = 4, b = 4, l = 4))

  if (is_right) {
    p <- p + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                   axis.line.y = element_blank(), axis.title.y = element_blank())
  }
  p
}

# ── Panel (c): daily difference boxplot ────────────────────────────────────
n2o_diff_span <- diff(range(daily_n2o_diff$diff_wcc_minus_sorghum, na.rm = TRUE))
if (!is.finite(n2o_diff_span) || n2o_diff_span == 0) n2o_diff_span <- 1

n2o_diff_anno <- daily_n2o_diff |>
  summarise(y = max(diff_wcc_minus_sorghum, na.rm = TRUE), .by = year) |>
  left_join(n2o_diff_stats, by = "year") |>
  filter(significant) |>
  mutate(y = y + n2o_diff_span * 0.08)

p4c <- ggplot(daily_n2o_diff, aes(x = factor(year), y = diff_wcc_minus_sorghum)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
  geom_jitter(aes(color = factor(year)), width = 0.12, alpha = 0.7,
              size = 1.4, show.legend = FALSE) +
  geom_text(data = n2o_diff_anno,
            aes(x = factor(year), y = y, label = letter),
            inherit.aes = FALSE, fontface = "bold", size = 3.8) +
  scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
  labs(
    x = "Year",
    y = expression(atop("Daily N"[2]*"O Difference",
                        "(g N ha"^{-1}~"d"^{-1}*", WCC \u2212 Sorghum)")),
    tag = "(c)"
  ) +
  theme_publication(base_size = 10) +
  theme(plot.tag = element_text(face = "bold", size = 11),
        plot.tag.position = "topleft",
        plot.margin = margin(t = 14, r = 4, b = 4, l = 4),
        legend.position = "none")

# ── Panel (f): annual cumulative by treatment ──────────────────────────────
n2o_letter_offset <- diff(range(annual_n2o$annual_cum_n2o, na.rm = TRUE))
if (!is.finite(n2o_letter_offset) || n2o_letter_offset == 0) n2o_letter_offset <- 0.05

n2o_letter_pos <- annual_n2o |>
  summarise(y = max(annual_cum_n2o, na.rm = TRUE), .by = Treatment) |>
  left_join(n2o_letters, by = "Treatment") |>
  mutate(y = y + n2o_letter_offset * 0.08)

p4f <- ggplot(annual_n2o, aes(x = Treatment, y = annual_cum_n2o, fill = Treatment)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.5, color = "black") +
  geom_jitter(aes(shape = factor(year)), width = 0.12, size = 1.8,
              alpha = 0.8, show.legend = FALSE) +
  geom_text(data = n2o_letter_pos,
            aes(x = Treatment, y = y, label = letter),
            inherit.aes = FALSE, fontface = "bold", size = 3.8) +
  scale_fill_manual(values = treatment_colors) +
  labs(
    x = "",
    y = expression("Cumulative N"[2] * "O Flux (kg N ha"^-1 * ")"),
    subtitle = paste0("One-way ANOVA p = ",
                      ifelse(anova_n2o_p < 0.001, "< 0.001",
                             sprintf("%.3f", anova_n2o_p))),
    tag = "(f)"
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "none",
        plot.tag = element_text(face = "bold", size = 11),
        plot.tag.position = "topleft",
        plot.margin = margin(t = 14, r = 4, b = 4, l = 4))

# ── Build panels ──────────────────────────────────────────────────────────
p4a <- create_n2o_daily_panel(2023, is_right = FALSE, "(a)")
p4b <- create_n2o_daily_panel(2024, is_right = TRUE,  "(b)")
p4d <- create_n2o_cum_panel(2023, is_right = FALSE, "(d)")
p4e <- create_n2o_cum_panel(2024, is_right = TRUE,  "(e)")

# p4a keeps its legend; all others are suppressed
p4b_nl <- p4b + theme(legend.position = "none")

# Style p4a legend
p4a <- p4a + theme(
  legend.position   = "top",
  legend.direction  = "horizontal",
  legend.text       = element_text(size = 10),
  legend.title      = element_text(size = 10),
  legend.key.width  = unit(1.2, "cm"),
  legend.key.height = unit(0.4, "cm")
)

# ── Assemble ──────────────────────────────────────────────────────────────
figure4 <- guide_area() /
  ((p4a | p4b_nl | p4c) / (p4d | p4e | p4f) +
     plot_layout(widths = c(1, 1, 0.95))) +
  plot_layout(heights = c(0.05, 1), guides = "collect")

figure4

ggsave(filename = "figure_4_n2o_spacing.tiff", plot = figure4, width = 174, height = 116, units = "mm", dpi = 600, device = "tiff", compression = "lzw")
```

```{r print-stats}
#| echo: false
cat("Daily difference stats:\n")
n2o_diff_stats
cat("\nANOVA p-value:", anova_n2o_p, "\n")
```

# Alternative Statistical Approaches

## Approach B: Log-Transform + lmer

```{r approach-b-model}
#| message: false
#| warning: false

library(lme4)

# Ensure year is a factor for model fitting (avoids column-name issues in emmeans)
annual_n2o <- annual_n2o |>
  mutate(year_fct = factor(year))

# Fit mixed model on log-transformed annual cumulative N2O (sorghum pair)
mod_n2o_log <- lmer(
  log(annual_cum_n2o + 0.001) ~ Treatment * year_fct + (1 | plot),
  data = annual_n2o
)

summary(mod_n2o_log)
```

```{r approach-b-emmeans}
#| message: false
#| warning: false

# Back-transformed emmeans
emm_n2o_log <- emmeans(mod_n2o_log, ~ Treatment | year_fct, type = "response")
emm_n2o_log

# Pairwise contrasts
contrast_n2o_log <- pairs(emm_n2o_log)
contrast_n2o_log
```

```{r approach-b-diagnostics}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Residual diagnostic plots
par(mfrow = c(1, 2))
plot(mod_n2o_log, main = "Residuals vs Fitted (Log-lmer)")
qqnorm(resid(mod_n2o_log), main = "QQ Plot of Residuals (Log-lmer)")
qqline(resid(mod_n2o_log), col = "red")
par(mfrow = c(1, 1))
```

```{r approach-b-barchart}
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

# Extract means on the LOG scale for manual back-transformation
emm_b_log_scale <- emmeans(mod_n2o_log, ~ Treatment | year_fct)
emm_b_df <- as.data.frame(emm_b_log_scale)

# CLD letters from pairwise contrasts
cld_b <- tryCatch({
  cld(emm_b_log_scale, adjust = "tukey", Letters = letters) |>
    as.data.frame() |>
    mutate(letter = gsub("\\s+", "", .group))
}, error = \(e) {
  emm_b_df |> mutate(letter = "a")
})

# Merge letters with emmeans
emm_b_plot <- emm_b_df |>
  left_join(
    cld_b |> select(Treatment, year_fct, letter),
    by = c("Treatment", "year_fct")
  ) |>
  mutate(
    # Back-transform from log scale for bar heights
    bt_mean = exp(emmean),
    bt_lower = exp(lower.CL),
    bt_upper = exp(upper.CL)
  )

ggplot(emm_b_plot, aes(x = Treatment, y = bt_mean, fill = Treatment)) +
  geom_col(width = 0.6, alpha = 0.7, color = "black") +
  geom_errorbar(aes(ymin = bt_lower, ymax = bt_upper), width = 0.2) +
  geom_text(aes(y = bt_upper, label = letter),
            vjust = -0.5, fontface = "bold", size = 4) +
  facet_wrap(~ year_fct) +
  scale_fill_manual(values = treatment_colors) +
  labs(
    x = "", y = expression("Back-Transformed Cumulative N"[2]*"O (kg N ha"^-1*")"),
    title = "Approach B: Log-lmer Back-Transformed Means with CLD"
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "none")
```

## Approach C: glmmTMB with Gamma Family

```{r approach-c-model}
#| message: false
#| warning: false

has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)

if (has_glmmTMB) {
  library(glmmTMB)

  mod_n2o_gamma <- glmmTMB(
    annual_cum_n2o ~ Treatment * year_fct + (1 | plot),
    family = Gamma(link = "log"),
    data = annual_n2o
  )

  summary(mod_n2o_gamma)
} else {
  mod_n2o_gamma <- NULL
  cat("⚠ glmmTMB is not installed. Install with:\n")
  cat("  install.packages('glmmTMB')\n")
  cat("Skipping Approach C (Gamma family model).\n")
}
```

```{r approach-c-emmeans}
#| message: false
#| warning: false

if (has_glmmTMB && !is.null(mod_n2o_gamma)) {
  # Pairwise comparisons on the response scale
  emm_n2o_gamma <- emmeans(mod_n2o_gamma, ~ Treatment | year_fct, type = "response")
  emm_n2o_gamma

  contrast_n2o_gamma <- pairs(emm_n2o_gamma)
  contrast_n2o_gamma
} else {
  emm_n2o_gamma <- NULL
  contrast_n2o_gamma <- NULL
}
```

```{r approach-c-diagnostics}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

if (has_glmmTMB && !is.null(mod_n2o_gamma)) {
  # Residual diagnostics — prefer DHARMa if available, otherwise base plots
  if (requireNamespace("DHARMa", quietly = TRUE)) {
    sim_res_gamma <- DHARMa::simulateResiduals(mod_n2o_gamma, plot = FALSE)
    plot(sim_res_gamma, main = "DHARMa Residual Diagnostics (Gamma glmmTMB)")
  } else {
    res_gamma <- residuals(mod_n2o_gamma, type = "pearson")
    fit_gamma <- fitted(mod_n2o_gamma)
    par(mfrow = c(1, 2))
    plot(fit_gamma, res_gamma,
         xlab = "Fitted values", ylab = "Pearson residuals",
         main = "Residuals vs Fitted (Gamma glmmTMB)")
    abline(h = 0, lty = 2, col = "red")
    qqnorm(res_gamma, main = "QQ Plot of Residuals (Gamma glmmTMB)")
    qqline(res_gamma, col = "red")
    par(mfrow = c(1, 1))
  }
} else {
  cat("Skipped — glmmTMB not available.\n")
}
```

```{r approach-c-aic}
#| message: false
#| warning: false

if (has_glmmTMB && !is.null(mod_n2o_gamma)) {
  # Compare AIC/BIC between log-lmer (Approach B) and Gamma glmmTMB (Approach C)
  aic_comparison <- tibble(
    Model = c("Approach B: Log-lmer", "Approach C: Gamma glmmTMB"),
    AIC = c(AIC(mod_n2o_log), AIC(mod_n2o_gamma)),
    BIC = c(BIC(mod_n2o_log), BIC(mod_n2o_gamma))
  )
  aic_comparison

  cat("\nNote: AIC/BIC values are not directly comparable between log-lmer and\n")
  cat("Gamma glmmTMB because the response variables differ (log(y+0.001) vs y).\n")
  cat("Use residual diagnostics to judge which model fits better.\n")
} else {
  cat("Skipped — glmmTMB not available.\n")
}
```

## Approach D: Non-Parametric (Kruskal-Wallis + Dunn's)

```{r approach-d-kruskal}
#| message: false
#| warning: false

library(rstatix)

# Kruskal-Wallis test by year on annual cumulative N2O (sorghum pair)
kw_n2o_results <- annual_n2o |>
  group_by(year) |>
  kruskal_test(annual_cum_n2o ~ Treatment)

cat("Kruskal-Wallis results by year:\n")
kw_n2o_results
```

```{r approach-d-dunns}
#| message: false
#| warning: false

# Follow up with Dunn's test if any year is significant
kw_significant_years <- kw_n2o_results |> filter(p < 0.05)

if (nrow(kw_significant_years) > 0) {
  dunn_n2o_results <- annual_n2o |>
    filter(year %in% kw_significant_years$year) |>
    group_by(year) |>
    dunn_test(annual_cum_n2o ~ Treatment, p.adjust.method = "bonferroni")
  cat("Dunn's post-hoc test results (significant years only):\n")
  dunn_n2o_results
} else {
  dunn_n2o_results <- NULL
  cat("No years had significant Kruskal-Wallis results; Dunn's test not performed.\n")
}

cat("\nNote: Kruskal-Wallis / Dunn's is a non-parametric fallback appropriate\n")
cat("when parametric model residuals show severe departures from assumptions.\n")
```

## Approach E: All-Treatment ANOVA with CLD

```{r approach-e-data}
#| message: false
#| warning: false

# Compute plot-level daily flux for ALL treatments
all_trt_flux <- flux_data |>
  filter(year %in% c(2023, 2024)) |>
  mutate(Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment)) |>
  summarise(daily_flux_kg = mean(gnha_day_no_negative, na.rm = TRUE) / 1000,
            .by = c(Treatment, year, plot, date))

# Interpolation and cumulative sum (same method as existing code)
all_trt_cum <- all_trt_flux |>
  arrange(Treatment, year, plot, date) |>
  group_by(Treatment, year, plot) |>
  group_modify(\(df, keys) {
    df <- df |>
      complete(date = seq(min(date), max(date), by = "day")) |>
      arrange(date)
    non_na <- !is.na(df$daily_flux_kg)
    if (sum(non_na) == 0) return(tibble())

    flux_interp <- if (sum(non_na) >= 2) {
      approx(df$date[non_na], df$daily_flux_kg[non_na],
             xout = df$date, rule = 1, ties = "ordered")$y
    } else {
      rep(df$daily_flux_kg[non_na][1], nrow(df))
    }

    df |>
      mutate(flux_interp = flux_interp, cum_flux_kg = cumsum(flux_interp)) |>
      filter(!is.na(flux_interp))
  }) |>
  ungroup()

# Annual totals for all treatments
annual_n2o_all <- all_trt_cum |>
  summarise(annual_cum_n2o = max(cum_flux_kg, na.rm = TRUE),
            .by = c(year, plot, Treatment)) |>
  filter(is.finite(annual_cum_n2o), annual_cum_n2o >= 0)

cat("Annual cumulative N2O — all treatments:\n")
annual_n2o_all |>
  summarise(
    mean = mean(annual_cum_n2o),
    sd = sd(annual_cum_n2o),
    n = n(),
    .by = c(Treatment, year)
  ) |>
  arrange(year, Treatment)
```

```{r approach-e-model}
#| message: false
#| warning: false

# Fit ANOVA / lmer with all treatments
annual_n2o_all <- annual_n2o_all |>
  mutate(year_fct = factor(year))

mod_n2o_all <- lmer(
  annual_cum_n2o ~ Treatment * year_fct + (1 | plot),
  data = annual_n2o_all
)

summary(mod_n2o_all)
anova_n2o_all <- anova(mod_n2o_all)
anova_n2o_all
```

```{r approach-e-cld}
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 6

# CLD via emmeans for all treatments
emm_n2o_all <- emmeans(mod_n2o_all, ~ Treatment | year_fct)

cld_n2o_all <- tryCatch({
  cld(emm_n2o_all, adjust = "tukey", Letters = letters) |>
    as.data.frame() |>
    mutate(letter = gsub("\\s+", "", .group))
}, error = \(e) {
  as.data.frame(emm_n2o_all) |> mutate(letter = "a")
})

cat("CLD groups for all treatments by year:\n")
cld_n2o_all |> select(Treatment, year_fct, emmean, SE, letter)

# Bar chart with all treatments
all_trt_colors <- c("Sorghum" = "#D8D97AFF", "Sorghum + WCC" = "#95C36EFF",
                    "Corn" = "#74C8C3FF", "Soy" = "#0A2E57FF")

ggplot(cld_n2o_all, aes(x = Treatment, y = emmean, fill = Treatment)) +
  geom_col(width = 0.6, alpha = 0.7, color = "black") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_text(aes(y = upper.CL, label = letter),
            vjust = -0.5, fontface = "bold", size = 4) +
  geom_jitter(data = annual_n2o_all,
              aes(x = Treatment, y = annual_cum_n2o),
              width = 0.1, size = 1.5, alpha = 0.5, inherit.aes = FALSE) +
  facet_wrap(~ year_fct) +
  scale_fill_manual(values = all_trt_colors) +
  labs(
    x = "",
    y = expression("Annual Cumulative N"[2]*"O (kg N ha"^-1*")"),
    title = "Approach E: All-Treatment ANOVA with CLD by Year"
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "none")
```

## Approach F: BH FDR-Corrected Per-Date Wilcoxon

```{r approach-f-fdr}
#| message: false
#| warning: false

# Apply Benjamini-Hochberg FDR correction to per-date Wilcoxon p-values within each year
daily_sig_fdr <- daily_significance |>
  filter(!is.na(p_value)) |>
  group_by(year) |>
  mutate(
    p_value_bh = p.adjust(p_value, method = "BH"),
    significant_bh = p_value_bh < 0.05
  ) |>
  ungroup()

# Comparison: significant dates before vs after FDR correction
fdr_comparison <- daily_sig_fdr |>
  summarise(
    total_dates = n(),
    sig_uncorrected = sum(significant, na.rm = TRUE),
    sig_fdr = sum(significant_bh, na.rm = TRUE),
    .by = year
  )

cat("Per-date Wilcoxon significance: Uncorrected vs BH FDR-corrected\n")
fdr_comparison

# Show details of dates that change significance status
cat("\nDates significant before but NOT after FDR correction:\n")
daily_sig_fdr |>
  filter(significant & !significant_bh) |>
  select(year, date, p_value, p_value_bh) |>
  arrange(year, date)

cat("\nDates significant after FDR correction:\n")
daily_sig_fdr |>
  filter(significant_bh) |>
  select(year, date, p_value, p_value_bh) |>
  arrange(year, date)
```

## Statistical Comparison

```{r stat-comparison}
#| message: false
#| warning: false

# ── Gather p-values and summaries from each approach ─────────────────────

# Approach A: existing ANOVA + Wilcoxon
n_sig_wilcox_uncorrected <- daily_significance |>
  filter(significant) |>
  nrow()

# Approach B: log-lmer
anova_b <- anova(mod_n2o_log)
# lmerTest provides "Pr(>F)"; without it, column may be "Pr(>Chisq)" or absent
p_treatment_b <- tryCatch({
  ab <- anova_b["Treatment", ]
  p_col <- intersect(c("Pr(>F)", "Pr(>Chisq)"), names(ab))
  if (length(p_col) > 0) ab[[p_col[1]]] else NA_real_
}, error = \(e) NA_real_)
if (is.null(p_treatment_b) || length(p_treatment_b) == 0) p_treatment_b <- NA_real_
pairwise_b_df <- as.data.frame(contrast_n2o_log)

# Approach C: Gamma glmmTMB (may be NULL if package not installed)
p_treatment_c <- NA_real_
pairwise_c_df <- NULL
if (has_glmmTMB && !is.null(mod_n2o_gamma)) {
  summ_c <- summary(mod_n2o_gamma)
  coef_c <- summ_c$coefficients$cond
  p_treatment_c <- tryCatch(coef_c["TreatmentSorghum + WCC", "Pr(>|z|)"], error = \(e) NA_real_)
  pairwise_c_df <- as.data.frame(contrast_n2o_gamma)
}

# Approach D: Kruskal-Wallis
kw_pvals <- kw_n2o_results |> select(year, p)

# Approach E: all-treatment CLD
cld_summary_e <- cld_n2o_all |>
  select(Treatment, year_fct, letter) |>
  arrange(year_fct, Treatment)

# Approach F: FDR-corrected Wilcoxon
n_sig_fdr <- daily_sig_fdr |> filter(significant_bh) |> nrow()

# ── Build summary table ──────────────────────────────────────────────────
format_p <- function(p) {
  if (is.null(p) || length(p) == 0 || is.na(p)) return("NA")
  if (p < 0.001) return("< 0.001")
  sprintf("%.4f", p)
}

comparison_table <- tibble(
  Approach = c(
    "A: One-way ANOVA (existing)",
    "B: Log-lmer (Treatment * Year)",
    "C: Gamma glmmTMB (Treatment * Year)",
    "D: Kruskal-Wallis (by year)",
    "E: All-treatment lmer with CLD",
    "F: BH FDR-corrected Wilcoxon"
  ),
  Key_Metric = c(
    paste0("ANOVA p = ", format_p(anova_n2o_p),
           "; ", n_sig_wilcox_uncorrected, " sig. Wilcoxon dates"),
    paste0("Treatment p = ", format_p(p_treatment_b),
           "; pairwise: ", paste(sapply(pairwise_b_df$p.value, format_p), collapse = ", ")),
    paste0("Treatment p = ", format_p(p_treatment_c),
           if (has_glmmTMB && !is.null(mod_n2o_gamma))
             paste0("; AIC = ", round(AIC(mod_n2o_gamma), 1))
           else "; glmmTMB not installed"),
    paste0("KW p by year: ",
           paste(paste0(kw_pvals$year, "=", sapply(kw_pvals$p, format_p)), collapse = "; ")),
    paste0("CLD: ",
           paste(paste0(cld_summary_e$Treatment, "(", cld_summary_e$year_fct, ")=",
                        cld_summary_e$letter), collapse = ", ")),
    paste0(n_sig_fdr, " sig. dates after FDR (was ", n_sig_wilcox_uncorrected, " uncorrected)")
  )
)

cat("════════════════════════════════════════════════════════════════════\n")
cat("                   STATISTICAL COMPARISON TABLE                    \n")
cat("════════════════════════════════════════════════════════════════════\n\n")

comparison_table |>
  mutate(row = paste0(Approach, "\n  → ", Key_Metric)) |>
  pull(row) |>
  walk(~ cat(.x, "\n\n"))
```

```{r residual-comparison}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# ── Residual comparison: Approach B vs Approach C ────────────────────────
if (has_glmmTMB && !is.null(mod_n2o_gamma)) {
  par(mfrow = c(1, 2))

  # Approach B residuals
  qqnorm(resid(mod_n2o_log),
         main = "Approach B: Log-lmer QQ Plot",
         col = "#95C36EFF", pch = 16)
  qqline(resid(mod_n2o_log), col = "red", lwd = 2)

  # Approach C residuals
  res_c <- residuals(mod_n2o_gamma, type = "pearson")
  qqnorm(res_c,
         main = "Approach C: Gamma glmmTMB QQ Plot",
         col = "#74C8C3FF", pch = 16)
  qqline(res_c, col = "red", lwd = 2)

  par(mfrow = c(1, 1))

  # Shapiro-Wilk normality test on residuals
  cat("Shapiro-Wilk test on residuals:\n")
  cat("  Approach B (log-lmer):     p =",
      format_p(shapiro.test(resid(mod_n2o_log))$p.value), "\n")
  cat("  Approach C (Gamma glmmTMB): p =",
      format_p(shapiro.test(res_c)$p.value), "\n")
} else {
  # Only show Approach B residuals
  qqnorm(resid(mod_n2o_log),
         main = "Approach B: Log-lmer QQ Plot",
         col = "#95C36EFF", pch = 16)
  qqline(resid(mod_n2o_log), col = "red", lwd = 2)

  cat("Shapiro-Wilk test on Approach B residuals: p =",
      format_p(shapiro.test(resid(mod_n2o_log))$p.value), "\n")
  cat("Approach C residual comparison skipped — glmmTMB not installed.\n")
}
```
