---
title: "Figure 1: Weather Conditions"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 10
    fig-height: 4.5
execute:
  warning: false
  message: false
  echo: true
---

# Scientific Context

Weather conditions varied substantially between study years. The year 2023
was characterized by below-average precipitation, while 2024 experienced
conditions closer to the historical mean. These differences influenced soil
moisture dynamics, crop growth, and nitrogen cycling processes.

This figure shows cumulative precipitation and growing degree days (GDD) for
2023 and 2024 against the 34-year historical envelope (min--max range) and
mean.

# Setup

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(patchwork)
library(lubridate)

# ── Publication theme ──────────────────────────────────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(3, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "top",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}
```

# Statistical Approach

**Current method:** No formal statistics — this is a descriptive/contextual
figure showing cumulative weather trajectories against a historical envelope.

**Assessment:** This is the correct approach. Weather panels serve as context
for interpreting treatment effects; hypothesis testing (e.g., "was 2023
significantly drier?") is not the goal. The historical min--max ribbon with
the 34-year mean line provides an appropriate visual reference.

**Potential enhancements (not implemented):**

- *Percentile bands:* Replace the min--max envelope with 10th/90th percentile
  ribbons to reduce the influence of extreme single-year outliers and give a
  better sense of "typical" variability.
- *Anomaly framing:* Show 2023/2024 as departures from the long-term mean
  (i.e., cumulative anomaly curves) to more directly quantify how unusual each
  year was.

Neither alternative would change the manuscript conclusions; the current
approach clearly communicates the drought--normal contrast.

# Data Processing

```{r load-data}
# Load historical weather (precipitation and GDD share the same file)
weather_raw <- read_xlsx("data/rain/historical_rain.xlsx")

# ── Precipitation ──────────────────────────────────────────────────────────
rain <- weather_raw |>
  select(day, doy, precipmm) |>
  mutate(date = as.Date(day), year = year(date)) |>
  group_by(year) |>
  arrange(date) |>
  mutate(
    cumulative_precip = cumsum(precipmm),
    doy_date = as.Date(doy - 1, origin = "2000-01-01")
  ) |>
  ungroup()

# ── Growing degree days ───────────────────────────────────────────────────
gdd <- weather_raw |>
  select(day, doy, gdd_50_86) |>
  mutate(date = as.Date(day), year = year(date)) |>
  group_by(year) |>
  arrange(date) |>
  mutate(
    cumulative_gdd = cumsum(gdd_50_86),
    doy_date = as.Date(doy - 1, origin = "2000-01-01")
  ) |>
  ungroup()
```

```{r compute-envelopes}
# Historical envelopes (excluding study years)
hist_years <- . |> filter(!year %in% c(2023, 2024))

hist_env_rain <- rain |>
  hist_years() |>
  summarise(
    min_cum = min(cumulative_precip, na.rm = TRUE),
    max_cum = max(cumulative_precip, na.rm = TRUE),
    .by = doy_date
  )

hist_env_gdd <- gdd |>
  hist_years() |>
  summarise(
    min_cum = min(cumulative_gdd, na.rm = TRUE),
    max_cum = max(cumulative_gdd, na.rm = TRUE),
    .by = doy_date
  )

# 34-year historical means
mean_cum_rain <- rain |>
  hist_years() |>
  summarise(cumulative_precip = mean(cumulative_precip, na.rm = TRUE), .by = doy_date) |>
  mutate(year = "Mean") |>
  filter(doy_date != "2000-12-31")

mean_cum_gdd <- gdd |>
  hist_years() |>
  summarise(cumulative_gdd = mean(cumulative_gdd, na.rm = TRUE), .by = doy_date) |>
  mutate(year = "Mean") |>
  filter(doy_date != "2000-12-31")

# Combine highlighted years (2023, 2024) with the historical mean
rain_highlight <- rain |>
  filter(year %in% c(2023, 2024)) |>
  mutate(year = as.character(year)) |>
  bind_rows(mean_cum_rain)

gdd_highlight <- gdd |>
  filter(year %in% c(2023, 2024)) |>
  mutate(year = as.character(year)) |>
  bind_rows(mean_cum_gdd)
```

# Figure 1

```{r figure1}
#| fig-width: 10
#| fig-height: 4.5
#| fig-dpi: 300

# Shared scale definitions
year_colors    <- c("2023" = "#E41A1C", "2024" = "#377EB8", "Mean" = "black")
year_labels    <- c("2023" = "2023", "2024" = "2024", "Mean" = "34-year Mean")
year_linetypes <- c("2023" = "solid", "2024" = "dashed", "Mean" = "dotdash")

# ── Panel (a): Cumulative precipitation ────────────────────────────────────
rain_plot <- ggplot() +
  geom_ribbon(
    data = hist_env_rain,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  geom_line(
    data = rain_highlight,
    aes(x = doy_date, y = cumulative_precip, color = year, linetype = year),
    linewidth = 0.7
  ) +
  scale_color_manual(name = "Year", values = year_colors, labels = year_labels) +
  scale_linetype_manual(name = "Year", values = year_linetypes, labels = year_labels) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  labs(x = "", y = "Cumulative Precipitation (mm)", tag = "(a)") +
  theme_publication(base_size = 10)

# ── Panel (b): Cumulative GDD ─────────────────────────────────────────────
gdd_plot <- ggplot() +
  geom_ribbon(
    data = hist_env_gdd,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  geom_line(
    data = gdd_highlight,
    aes(x = doy_date, y = cumulative_gdd, color = year, linetype = year),
    linewidth = 0.7
  ) +
  scale_color_manual(name = "Year", values = year_colors, labels = year_labels) +
  scale_linetype_manual(name = "Year", values = year_linetypes, labels = year_labels) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  labs(x = "", y = "Cumulative Growing Degree Days (GDD)", tag = "(b)") +
  theme_publication(base_size = 10)

# ── Assemble ───────────────────────────────────────────────────────────────
figure1 <- (rain_plot | gdd_plot) +
  plot_layout(guides = "collect", widths = c(1.05, 1.05)) &
  theme_publication(base_size = 10) &
  theme(
    legend.position  = "top",
    legend.text      = element_text(size = 9),
    legend.key.width = unit(1.1, "cm"),
    plot.margin      = margin(t = 4, r = 6, b = 3, l = 6, unit = "pt"),
    plot.tag         = element_text(face = "bold", size = 11),
    plot.tag.position = "topleft"
  ) &
  guides(
    color    = guide_legend(nrow = 1, byrow = TRUE),
    linetype = guide_legend(nrow = 1, byrow = TRUE)
  )

figure1
```
