---
title: "Figure 3: Nitrate Leaching"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
  echo: true
---

# Scientific Context

Nitrate leaching via subsurface tile drainage represents a primary nitrogen
loss pathway in agricultural systems. This figure shows daily and cumulative
nitrogen losses through tile drainage, demonstrating the key finding that the
winter cover crop substantially reduced nitrogen leaching compared to the
no-cover control.

The figure has six panels arranged in a 2 x 3 grid:

- **(a--b)** Daily leached N with precipitation overlay (2023, 2024)
- **(c)** Daily treatment difference boxplot (WCC minus Sorghum) by year
- **(d--e)** Cumulative leached N (2023, 2024)
- **(f)** Annual cumulative leached N by treatment (years combined)

# Setup

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(patchwork)
library(lubridate)
library(janitor)
library(emmeans)
library(multcomp)

# ── Publication theme ──────────────────────────────────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(3, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "top",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}

treatment_colors <- c(
  "Sorghum"       = "#D8D97AFF",
  "Sorghum + Rye" = "#95C36EFF",
  "Sorghum + WCC" = "#95C36EFF",
  "Corn"          = "#74C8C3FF",
  "Soy"           = "#0A2E57FF"
)
```

# Statistical Approach

**Current methods:**

1. **Per-date Wilcoxon rank-sum tests** (*p* < 0.05) to place asterisks on
   dates with significant treatment differences in daily leaching.
2. **One-sample t-test** of daily differences (WCC minus Sorghum) against zero
   for the boxplot panel (c).
3. **One-way ANOVA** with Tukey-adjusted pairwise comparisons (via `emmeans`)
   on annual cumulative leaching, pooling both years, for panel (f).

**Assessment:**

- **Per-date Wilcoxon tests (panels a--b):** Same multiple-comparison concern
  as Figure 2. With only ~3--4 plots per treatment on any given date, the
  Wilcoxon test has very low power (max test statistic with n = 4 vs n = 4 can
  only achieve *p* ~ 0.03 under perfect separation). Consider whether the
  asterisks overstate certainty on individual dates.

- **Annual ANOVA (panel f):** Pooling years and ignoring the year factor treats
  2023 and 2024 observations as exchangeable replicates, which inflates the
  effective sample size. A mixed model with year as a random or fixed effect
  would be more defensible.

**Suggested alternatives:**

- **For panels a--b:** Wilcoxon with Benjamini--Hochberg FDR correction across
  dates, or replace per-date tests with a cumulative divergence test (e.g.,
  permutation test on the area between cumulative curves).
- **For panel f:** A two-way ANOVA or mixed model
  `lmer(annual_N ~ Treatment + (1|year))` that properly accounts for the
  year structure. Given only 2 years, treating year as fixed
  (`annual_N ~ Treatment + year`) may be more appropriate than random.
- **Log transformation:** Leaching data are right-skewed. A log or
  square-root transformation (or a generalized linear model with Gamma family)
  would better meet normality assumptions for ANOVA.

# Data Processing

```{r load-data}
# Constants
plot_area_ft2 <- 120 * 160
plot_area_ha  <- plot_area_ft2 / 107639.1041671

fert_events <- tibble(
  year            = c(2023, 2024),
  fertilizer_date = as.Date(c("2023-05-05", "2024-07-17"))
)

leaching_data <- read_csv("data/clean_leaching.csv", show_col_types = FALSE) |>
  clean_names() |>
  mutate(date = as.Date(date), year = year(date)) |>
  filter(year %in% c(2023, 2024))

# Precipitation (for overlay bars)
rain <- read_xlsx("data/rain/historical_rain.xlsx") |>
  select(day, doy, precipmm) |>
  mutate(date = as.Date(day), year = year(date))
```

```{r compute-summaries}
# ── Daily leaching summary ─────────────────────────────────────────────────
daily_leaching_summary <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    daily_leached_n_kgha = (total_n_loss_mg / 1e6) / plot_area_ha
  ) |>
  summarise(
    mean_leached = mean(daily_leached_n_kgha, na.rm = TRUE),
    se_leached   = sd(daily_leached_n_kgha, na.rm = TRUE) /
                   sqrt(sum(!is.na(daily_leached_n_kgha))),
    .by = c(Treatment, year, date)
  )

# ── Cumulative leaching summary ───────────────────────────────────────────
cum_leaching_summary <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    cumulative_leached_n_kgha = (cumulative_n_loss_mg / 1e6) / plot_area_ha
  ) |>
  summarise(
    cum_leached_mean = mean(cumulative_leached_n_kgha, na.rm = TRUE),
    cum_leached_sd   = sd(cumulative_leached_n_kgha, na.rm = TRUE),
    n                = n(),
    cum_leached_se   = cum_leached_sd / sqrt(n),
    .by = c(Treatment, year, date)
  )
```

```{r significance-tests}
# ── Per-date Wilcoxon tests (asterisks) ────────────────────────────────────
daily_leaching_sig <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    daily_leached_n_kgha = (total_n_loss_mg / 1e6) / plot_area_ha
  ) |>
  summarise(
    p_value = {
      s <- daily_leached_n_kgha[Treatment == "Sorghum"]       |> na.omit()
      w <- daily_leached_n_kgha[Treatment == "Sorghum + WCC"] |> na.omit()
      if (length(s) >= 3 && length(w) >= 3)
        tryCatch(wilcox.test(s, w)$p.value, error = \(e) NA_real_)
      else NA_real_
    },
    .by = c(year, date)
  ) |>
  mutate(significant = !is.na(p_value) & p_value < 0.05)

leaching_stars <- daily_leaching_summary |>
  summarise(y_top = max(mean_leached + se_leached, na.rm = TRUE), .by = c(year, date)) |>
  left_join(daily_leaching_sig, by = c("year", "date")) |>
  filter(significant) |>
  mutate(star_y = y_top * 1.1, star_label = "*")
```

```{r daily-diff}
# ── Daily difference (WCC - Sorghum) and one-sample t-test ────────────────
daily_leach_diff <- daily_leaching_summary |>
  mutate(trt_key = if_else(Treatment == "Sorghum + WCC", "WCC", "Sorghum")) |>
  select(year, date, trt_key, mean_leached) |>
  pivot_wider(names_from = trt_key, values_from = mean_leached) |>
  mutate(diff_wcc_minus_sorghum = WCC - Sorghum)

one_sample_vs_zero <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, mu = 0)$p.value, error = \(e) NA_real_)
}

leach_diff_stats <- daily_leach_diff |>
  summarise(
    n          = sum(!is.na(diff_wcc_minus_sorghum)),
    mean_diff  = mean(diff_wcc_minus_sorghum, na.rm = TRUE),
    p_value    = one_sample_vs_zero(diff_wcc_minus_sorghum),
    significant = !is.na(p_value) & p_value < 0.05,
    letter     = if_else(significant, "a", NA_character_),
    .by = year
  )
```

```{r annual-anova}
# ── Annual cumulative leaching + one-way ANOVA ────────────────────────────
annual_leached <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    plot = as.character(plot)
  ) |>
  summarise(
    annual_leached_n = (max(cumulative_n_loss_mg, na.rm = TRUE) / 1e6) / plot_area_ha,
    .by = c(year, plot, Treatment)
  ) |>
  filter(is.finite(annual_leached_n), annual_leached_n >= 0)

anova_leached   <- aov(annual_leached_n ~ Treatment, data = annual_leached)
anova_leached_p <- summary(anova_leached)[[1]]["Treatment", "Pr(>F)"]

leached_letters <- tryCatch({
  emmeans(anova_leached, ~ Treatment) |>
    cld(adjust = "tukey", Letters = letters) |>
    as_tibble() |>
    transmute(Treatment = as.character(Treatment),
              letter    = gsub("\\s+", "", .group))
}, error = \(e) {
  annual_leached |> distinct(Treatment) |> mutate(letter = "a")
})
```

# Figure 3

```{r figure3}
#| fig-width: 12
#| fig-height: 8
#| fig-dpi: 300

# ── Precipitation trimmed to leaching windows ─────────────────────────────
leach_date_ranges <- daily_leaching_summary |>
  summarise(min_date = min(date), max_date = max(date), .by = year)

daily_precip_leaching <- rain |>
  select(date, year, precipmm) |>
  filter(year %in% c(2023, 2024)) |>
  left_join(leach_date_ranges, by = "year") |>
  filter(date >= min_date, date <= max_date)

# ── Scaling constants ─────────────────────────────────────────────────────
max_leached_daily <- max(daily_leaching_summary$mean_leached +
                         daily_leaching_summary$se_leached, na.rm = TRUE)
max_precip_leach  <- max(daily_precip_leaching$precipmm, na.rm = TRUE)
leach_precip_scale <- (max_leached_daily * 0.4) / max_precip_leach
leach_plot_top     <- max_leached_daily * 1.12
max_cum_leached    <- max(cum_leaching_summary$cum_leached_mean +
                          cum_leaching_summary$cum_leached_se, na.rm = TRUE)
## Panel Builder

# ── Panel builder: daily leaching ─────────────────────────────────────────
create_leaching_daily_panel <- function(year_val, is_right, tag_label) {
  d_sub    <- filter(daily_leaching_summary, year == year_val)
  p_sub    <- filter(daily_precip_leaching, year == year_val)
  s_sub    <- filter(leaching_stars, year == year_val)
  fert_sub <- filter(fert_events, year == year_val)
  x_lims   <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot() +
    # Precipitation (inverted from top)
    geom_rect(
      data = p_sub,
      aes(xmin = date - 0.5, xmax = date + 0.5,
          ymin = leach_plot_top - precipmm * leach_precip_scale,
          ymax = leach_plot_top),
      fill = "#377EB8", alpha = 0.7
    ) +
    geom_point(data = d_sub, aes(x = date, y = mean_leached, color = Treatment),
               size = 1.4) +
    geom_errorbar(
      data = d_sub,
      aes(x = date,
          ymin = pmax(mean_leached - se_leached, 0),
          ymax = mean_leached + se_leached,
          color = Treatment),
      width = 0.2, alpha = 0.5
    ) +
    geom_vline(data = fert_sub, aes(xintercept = fertilizer_date),
               linetype = "dashed", color = "black", alpha = 0.7) +
    geom_text(data = s_sub, aes(x = date, y = star_y, label = star_label),
              size = 4.8) +
    scale_color_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right) expression("Daily Leached N (kg N ha"^-1 * " d"^-1 * ")") else "",
      limits = c(0, leach_plot_top),
      sec.axis = sec_axis(
        ~ (leach_plot_top - .) / leach_precip_scale,
        name = if (is_right) "Precipitation (mm)" else ""
      )
    ) +
    labs(x = "", tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(
      axis.text.x      = element_blank(),
      axis.ticks.x     = element_blank(),
      axis.line.x      = element_blank(),
      legend.position  = "none",
      plot.tag          = element_text(face = "bold", size = 11),
      plot.tag.position = "topleft"
    )

  if (is_right) {
    p <- p + theme(
      axis.text.y        = element_blank(),
      axis.ticks.y       = element_blank(),
      axis.line.y        = element_blank(),
      axis.title.y       = element_blank(),
      axis.text.y.right  = element_text(color = "#377EB8"),
      axis.ticks.y.right = element_line(),
      axis.title.y.right = element_text(color = "#377EB8")
    )
  } else {
    p <- p + theme(
      axis.text.y.right  = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.title.y.right = element_blank()
    )
  }
  p
}

# ── Panel builder: cumulative leaching ────────────────────────────────────
create_leaching_cum_panel <- function(year_val, is_right, tag_label) {
  c_sub    <- filter(cum_leaching_summary, year == year_val)
  fert_sub <- filter(fert_events, year == year_val)
  x_lims   <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot(c_sub, aes(x = date, y = cum_leached_mean,
                          color = Treatment, fill = Treatment)) +
    geom_ribbon(aes(ymin = pmax(cum_leached_mean - cum_leached_se, 0),
                    ymax = cum_leached_mean + cum_leached_se),
                alpha = 0.15, linewidth = 0) +
    geom_line(linewidth = 1) +
    geom_vline(data = fert_sub, aes(xintercept = fertilizer_date),
               linetype = "dashed", color = "black", alpha = 0.7) +
    scale_color_manual(values = treatment_colors) +
    scale_fill_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right) expression("Cumulative Leached N (kg N ha"^-1 * ")") else "",
      limits = c(0, max_cum_leached)
    ) +
    labs(x = "Month", tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(legend.position = "none",
          plot.tag = element_text(face = "bold", size = 11),
          plot.tag.position = "topleft")

  if (is_right) {
    p <- p + theme(
      axis.text.y  = element_blank(),
      axis.ticks.y = element_blank(),
      axis.line.y  = element_blank(),
      axis.title.y = element_blank()
    )
  }
  p
}

# ── Panel (c): daily difference boxplot ────────────────────────────────────
diff_span <- diff(range(daily_leach_diff$diff_wcc_minus_sorghum, na.rm = TRUE))
if (!is.finite(diff_span) || diff_span == 0) diff_span <- 0.05

leach_diff_anno <- daily_leach_diff |>
  summarise(y = max(diff_wcc_minus_sorghum, na.rm = TRUE), .by = year) |>
  left_join(leach_diff_stats, by = "year") |>
  filter(significant) |>
  mutate(y = y + diff_span * 0.08)

p3c <- ggplot(daily_leach_diff, aes(x = factor(year), y = diff_wcc_minus_sorghum)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
  geom_jitter(aes(color = factor(year)), width = 0.12, alpha = 0.7,
              size = 1.4, show.legend = FALSE) +
  geom_text(data = leach_diff_anno,
            aes(x = factor(year), y = y, label = letter),
            inherit.aes = FALSE, fontface = "bold", size = 3.8) +
  scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
  labs(
    x = "Year",
    y = expression("Daily Leached N Difference (kg N ha"^-1 * " d"^-1 * ", WCC - Sorghum)"),
    tag = "(c)"
  ) +
  theme_publication(base_size = 10) +
  theme(plot.tag = element_text(face = "bold", size = 11),
        plot.tag.position = "topleft")

# ── Panel (f): annual cumulative by treatment ──────────────────────────────
letter_offset <- diff(range(annual_leached$annual_leached_n, na.rm = TRUE))
if (!is.finite(letter_offset) || letter_offset == 0) letter_offset <- 0.05

annual_letter_pos <- annual_leached |>
  summarise(y = max(annual_leached_n, na.rm = TRUE), .by = Treatment) |>
  left_join(leached_letters, by = "Treatment") |>
  mutate(y = y + letter_offset * 0.08)

p3f <- ggplot(annual_leached, aes(x = Treatment, y = annual_leached_n, fill = Treatment)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.5, color = "black") +
  geom_jitter(aes(shape = factor(year)), width = 0.12, size = 1.8,
              alpha = 0.8, show.legend = FALSE) +
  geom_text(data = annual_letter_pos,
            aes(x = Treatment, y = y, label = letter),
            inherit.aes = FALSE, fontface = "bold", size = 3.8) +
  scale_fill_manual(values = treatment_colors) +
  labs(
    x = "",
    y = expression("Annual Leached N (kg N ha"^-1 * ")"),
    subtitle = paste0("One-way ANOVA p = ",
                      ifelse(anova_leached_p < 0.001, "< 0.001",
                             sprintf("%.3f", anova_leached_p))),
    tag = "(f)"
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "none",
        plot.tag = element_text(face = "bold", size = 11),
        plot.tag.position = "topleft")

# ── Build panels ──────────────────────────────────────────────────────────
p3a <- create_leaching_daily_panel(2023, is_right = FALSE, "(a)")
p3b <- create_leaching_daily_panel(2024, is_right = TRUE,  "(b)")
p3d <- create_leaching_cum_panel(2023, is_right = FALSE, "(d)")
p3e <- create_leaching_cum_panel(2024, is_right = TRUE,  "(e)")

# ── Assemble ──────────────────────────────────────────────────────────────
figure3 <- (p3a | p3b | p3c) / (p3d | p3e | p3f) +
  plot_layout(guides = "collect", widths = c(1, 1, 0.95)) &
  theme(legend.position = "top", legend.justification = "center")

figure3
```

```{r print-stats}
#| echo: false
cat("Daily difference stats:\n")
leach_diff_stats
cat("\nANOVA p-value:", anova_leached_p, "\n")
```
