---
title: "Figure 5: Biomass Yields"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
  echo: true
---

# Scientific Context

A critical question for cover crop adoption is whether the additional biomass
production impacts the primary cash crop yield. We measured end-of-season
biomass for both sorghum and cereal rye to determine if the cover crop caused a
"yield drag" on sorghum production.

Biomass is reported as oven-dried (0% moisture) equivalent dry matter.

**Note:** The sorghum biomass source files are stored in the researcher's Box
folder and not included in this repository. The code below demonstrates the
analysis workflow. Users with access to the complete dataset can update the file
paths accordingly.

# Setup

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(patchwork)

# ── Publication theme ──────────────────────────────────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(3, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "top",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}

treatment_levels <- c("Sorghum", "Sorghum + WCC")
```

# Statistical Approach

**Current method:** Welch's two-sample t-test comparing sorghum biomass
between treatments within each year. Compact letter display (CLD) marks groups
that differ significantly at *p* < 0.05.

**Assessment:** With only 4 plots per treatment per year, the t-test has low
power to detect moderate yield differences. The approach is appropriate given
the design (two groups, continuous outcome) but has limitations:

1. **Small sample size (n = 4):** The t-test relies on normality assumptions
   that are hard to verify with only 4 observations. The Welch correction for
   unequal variances helps, but the test has power to detect only very large
   effect sizes.
2. **Year treated independently:** Running separate t-tests per year loses
   the ability to borrow information across years and increases the risk of
   inconsistent conclusions.

**Suggested alternatives:**

- **Two-way ANOVA or mixed model:** `lm(biomass ~ Treatment * year)` or
  `lmer(biomass ~ Treatment + (1|year))` to jointly analyze both years. This
  pools error variance, giving better power estimates and a single treatment
  p-value.
- **Non-parametric alternative:** With n = 4, a Wilcoxon rank-sum test avoids
  normality assumptions but sacrifices even more power. Given the small sample,
  the t-test is pragmatically the better choice.
- **Equivalence testing (TOST):** If the goal is to demonstrate *no yield
  penalty*, a two-one-sided-t-test (TOST) procedure with a practically
  meaningful equivalence margin (e.g., +/- 2 Mg ha^-1^) would provide a
  stronger argument than simply "failing to reject H0."

# Data Processing

```{r load-sorghum-data}
# ── File paths (Box cloud — update as needed) ─────────────────────────────
sorghum_2023_path <- "/Users/jaredflater/Library/CloudStorage/Box-Box/CABBI/Data/MeasuredData/SABR/SmallDrainagePlots/Biomass/2023GrowingSeason/DerivedDatasets/chopper_load_weights_yields.xlsx"
sorghum_2024_path <- "/Users/jaredflater/Library/CloudStorage/Box-Box/CABBI/Data/MeasuredData/SABR/SmallDrainagePlots/Biomass/2024GrowingSeason/RecordingDataSheets/sorg_chopper_harvest_load_weights.xlsx"

# ── Constants ─────────────────────────────────────────────────────────────
plot_size_ha      <- 0.145
conversion_factor <- 0.907185  # short ton → metric ton

# Moisture-adjustment factor (oven-dry equivalent)
moisture_df  <- tibble(
  Plot     = c(2, 7, 9, 15),
  Moisture = c(0.749262537, 0.766325316, 0.731502423, 0.767241379)
)
avg_moisture <- mean(moisture_df$Moisture, na.rm = TRUE)

# ── 2023 sorghum ──────────────────────────────────────────────────────────
sorghum_2023 <- read_excel(sorghum_2023_path) |>
  summarise(total_biomass_ton = sum(`Net Weight`, na.rm = TRUE), .by = Plot) |>
  mutate(
    year      = 2023,
    Plot      = as.character(Plot),
    treatment = if_else(Plot %in% c("2", "3", "5", "15"), "Sorghum", "Sorghum + WCC"),
    biomass_mg_ha = ((total_biomass_ton * conversion_factor) / plot_size_ha) * (1 - avg_moisture)
  ) |>
  select(year, Plot, treatment, biomass_mg_ha)

# ── 2024 sorghum ──────────────────────────────────────────────────────────
sorghum_2024 <- read_excel(sorghum_2024_path) |>
  filter(Study == "Drainage") |>
  summarise(total_biomass_ton = sum(`Net Weight (t)`, na.rm = TRUE), .by = Plot) |>
  mutate(
    year      = 2024,
    Plot      = as.character(Plot),
    treatment = if_else(Plot %in% c("2", "3", "5", "15"), "Sorghum", "Sorghum + WCC"),
    biomass_mg_ha = ((total_biomass_ton * conversion_factor) / plot_size_ha) * (1 - avg_moisture)
  ) |>
  select(year, Plot, treatment, biomass_mg_ha)

sorghum_plot_data <- bind_rows(sorghum_2023, sorghum_2024) |>
  mutate(
    year      = factor(year, levels = c(2023, 2024)),
    treatment = factor(treatment, levels = treatment_levels)
  )
```

```{r load-rye-data}
# ── Rye cover crop biomass ────────────────────────────────────────────────
rye_data <- bind_rows(
  read_excel("data/BiomassData_Rye_20230522.xlsx") |> mutate(year = 2023),
  read_excel("data/BiomassData_Rye_20240515.xlsx") |> mutate(year = 2024)
) |>
  mutate(
    year          = factor(year, levels = c(2023, 2024)),
    Plot          = as.character(`Plot ID`),
    treatment     = factor("Sorghum + WCC", levels = treatment_levels),
    biomass_mg_ha = (`Dry Mass + Dry Bag Mass` - `Dry Bag Mass`) * 0.01
  ) |>
  select(year, Plot, treatment, biomass_mg_ha)
```

```{r compute-stats}
# ── Summary statistics ────────────────────────────────────────────────────
sorghum_stats <- sorghum_plot_data |>
  summarise(
    mean_sorghum = mean(biomass_mg_ha, na.rm = TRUE),
    sd_sorghum   = sd(biomass_mg_ha, na.rm = TRUE),
    n            = n(),
    se_sorghum   = sd_sorghum / sqrt(n),
    ci_sorghum   = qt(0.975, df = pmax(n - 1, 1)) * se_sorghum,
    .by = c(year, treatment)
  )

rye_stats <- rye_data |>
  summarise(
    mean_rye = mean(biomass_mg_ha, na.rm = TRUE),
    sd_rye   = sd(biomass_mg_ha, na.rm = TRUE),
    n        = n(),
    se_rye   = sd_rye / sqrt(n),
    ci_rye   = qt(0.975, df = pmax(n - 1, 1)) * se_rye,
    .by = c(year, treatment)
  )

# ── Per-year t-tests ──────────────────────────────────────────────────────
sorghum_tests_by_year <- sorghum_plot_data |>
  summarise(
    p_value = if (n_distinct(treatment) < 2) NA_real_
              else tryCatch(t.test(biomass_mg_ha ~ treatment)$p.value,
                            error = \(e) NA_real_),
    .by = year
  )

# ── Compact letter display ────────────────────────────────────────────────
sorghum_letters <- sorghum_stats |>
  left_join(sorghum_tests_by_year, by = "year") |>
  group_by(year) |>
  mutate(
    letter = case_when(
      is.na(p_value) | p_value >= 0.05 ~ "a",
      mean_sorghum == max(mean_sorghum, na.rm = TRUE) ~ "a",
      TRUE ~ "b"
    )
  ) |>
  ungroup()

letter_offsets <- sorghum_letters |>
  summarise(offset = max(mean_sorghum + ci_sorghum, na.rm = TRUE) * 0.08, .by = year)

sorghum_letters <- sorghum_letters |>
  left_join(letter_offsets, by = "year") |>
  mutate(y_label = mean_sorghum + ci_sorghum + offset)
```

# Figure 5

```{r figure5}
#| fig-width: 8
#| fig-height: 5
#| fig-dpi: 300

# ── Fill keys for layered bars ────────────────────────────────────────────
sorghum_stats <- sorghum_stats |> mutate(fill_key = as.character(treatment))
rye_overlay   <- rye_stats |> mutate(fill_key = "Rye overlay")

fill_palette <- c(
  "Sorghum"       = "#D8D97AFF",
  "Sorghum + WCC" = "#95C36EFF",
  "Rye overlay"   = "#F5DEB3"
)

# ── Rye layer (striped if ggpattern available) ────────────────────────────
rye_layer <- if (requireNamespace("ggpattern", quietly = TRUE)) {
  ggpattern::geom_col_pattern(
    data = rye_overlay,
    aes(x = treatment, y = mean_rye, fill = fill_key),
    pattern = "stripe", pattern_fill = "grey35", pattern_colour = "grey35",
    pattern_angle = 45, pattern_density = 0.45,
    alpha = 0.95, color = "black", width = 0.32, inherit.aes = FALSE
  )
} else {
  geom_col(
    data = rye_overlay,
    aes(x = treatment, y = mean_rye, fill = fill_key),
    alpha = 0.95, color = "black", width = 0.32, inherit.aes = FALSE
  )
}

# ── Panel labels ──────────────────────────────────────────────────────────
facet_panel_labels <- tibble(
  year      = factor(c(2023, 2024), levels = c(2023, 2024)),
  treatment = factor("Sorghum", levels = treatment_levels),
  y         = Inf,
  label     = c("(a)", "(b)")
)

# ── Build figure ──────────────────────────────────────────────────────────
figure5 <- ggplot() +
  # Sorghum bars
  geom_col(
    data = sorghum_stats,
    aes(x = treatment, y = mean_sorghum, fill = fill_key),
    color = "black", width = 0.7
  ) +
  geom_errorbar(
    data = sorghum_stats,
    aes(x = treatment,
        ymin = pmax(0, mean_sorghum - ci_sorghum),
        ymax = mean_sorghum + ci_sorghum),
    width = 0.18, inherit.aes = FALSE
  ) +
  # Rye overlay
  rye_layer +
  geom_errorbar(
    data = rye_overlay,
    aes(x = treatment,
        ymin = pmax(0, mean_rye - ci_rye),
        ymax = mean_rye + ci_rye),
    width = 0.1, inherit.aes = FALSE
  ) +
  # Significance letters
  geom_text(
    data = sorghum_letters,
    aes(x = treatment, y = y_label, label = letter),
    inherit.aes = FALSE, fontface = "bold", size = 3.8
  ) +
  # Panel labels via geom_text (consistent bold style)
  geom_text(
    data = facet_panel_labels,
    aes(x = treatment, y = y, label = label),
    inherit.aes = FALSE,
    hjust = -0.08, vjust = 1.15,
    fontface = "bold", size = 3.8
  ) +
  facet_wrap(~year, nrow = 1) +
  scale_fill_manual(
    values = fill_palette,
    breaks = c("Sorghum", "Sorghum + WCC", "Rye overlay"),
    name   = "Biomass component"
  ) +
  labs(
    x = "",
    y = expression("Oven-dried Biomass (Mg ha"^-1 * ")")
  ) +
  theme_publication(base_size = 10) +
  theme(
    legend.position    = "top",
    legend.justification = "center"
  )

figure5
```

```{r print-tests}
#| echo: false
cat("Per-year t-test p-values:\n")
sorghum_tests_by_year
```
