---
title: "Figure 4: Nitrous Oxide Emissions"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
  echo: true
---

# Scientific Context

Nitrous oxide (N~2~O) is a potent greenhouse gas produced during soil nitrogen
transformations. We measured N~2~O fluxes using static chambers throughout the
growing season to quantify emissions from sorghum systems with and without
winter cover crops. Fluxes were generally low and episodic, with peak emissions
often following fertilization or precipitation events.

The figure has six panels arranged in a 2 x 3 grid:

- **(a--b)** Daily N~2~O flux with precipitation overlay (2023, 2024)
- **(c)** Daily treatment difference boxplot (WCC minus Sorghum) by year
- **(d--e)** Cumulative N~2~O flux (2023, 2024)
- **(f)** Annual cumulative N~2~O by treatment (years combined)

# Setup

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(patchwork)
library(lubridate)
library(emmeans)
library(multcomp)

# ── Publication theme ──────────────────────────────────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(3, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "top",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}

treatment_colors <- c(
  "Sorghum"       = "#D8D97AFF",
  "Sorghum + Rye" = "#95C36EFF",
  "Sorghum + WCC" = "#95C36EFF"
)
```

# Statistical Approach

**Current methods:**

1. **Per-date Wilcoxon rank-sum tests** (*p* < 0.05) for asterisks on dates
   with significant treatment differences in daily N~2~O flux. Plot-level
   means are used as replicates (averaging row/interrow within a plot first).
2. **One-sample t-test** of daily differences (WCC minus Sorghum) against zero
   for the boxplot panel (c).
3. **One-way ANOVA** with Tukey-adjusted comparisons on annual cumulative
   N~2~O (pooling years) for panel (f).

**Assessment:**

- **Per-date Wilcoxon tests:** N~2~O fluxes are notoriously zero-inflated and
  right-skewed (many near-zero measurements, occasional spikes). The Wilcoxon
  test handles non-normality but with n = 4 per treatment the test has
  extremely low power; only perfect separation can achieve *p* < 0.05. The
  asterisks should be interpreted cautiously.

- **Cumulative calculation via linear interpolation:** Interpolating between
  measurement days is standard for chamber-based N~2~O studies, but it can
  miss short-lived emission pulses between sampling events. This is a known
  limitation of manual chamber methods.

- **Annual ANOVA (panel f):** Same concern as Figure 3 — pooling years
  without accounting for year structure.

**Suggested alternatives:**

- **Generalized linear mixed model:** For daily fluxes, a GLMM with Gamma or
  Tweedie family and a log link
  (`glmmTMB(flux ~ Treatment + (1|plot) + (1|date), family = Tweedie)`) would
  better handle the skewed, zero-inflated distribution while accounting for
  repeated measures.
- **Trapezoidal integration:** Replace linear interpolation with the
  trapezoidal rule for cumulative N~2~O calculation. In practice the two
  methods give very similar results for irregularly spaced data, but
  trapezoidal is the more standard approach in the soil science literature.
- **Panel f:** Same recommendation as Figure 3 — include year as a term.

# Data Processing

```{r load-data}
flux_data <- read_csv("data/cleaned_flux_data.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date), year = year(date))

rain <- read_xlsx("data/rain/historical_rain.xlsx") |>
  select(day, doy, precipmm) |>
  mutate(date = as.Date(day), year = year(date))

fert_events <- tibble(
  year            = c(2023, 2024),
  fertilizer_date = as.Date(c("2023-05-05", "2024-07-17"))
)

# Filter to sorghum treatments
flux_subset <- flux_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    Treatment = factor(Treatment, levels = c("Sorghum", "Sorghum + WCC"))
  ) |>
  filter(year %in% c(2023, 2024))
```

```{r daily-flux}
# Daily means in g N ha⁻¹ d⁻¹
daily_flux <- flux_subset |>
  summarise(
    mean_flux_g = mean(gnha_day_no_negative, na.rm = TRUE),
    se_flux_g   = sd(gnha_day_no_negative, na.rm = TRUE) /
                  sqrt(sum(!is.na(gnha_day_no_negative))),
    .by = c(Treatment, year, date)
  )
```

```{r significance-tests}
# Per-date Wilcoxon tests (plot-level means as replicates)
daily_significance <- flux_subset |>
  summarise(plot_mean_flux = mean(gnha_day_no_negative, na.rm = TRUE),
            .by = c(year, date, Treatment, plot)) |>
  summarise(
    p_value = {
      s <- plot_mean_flux[Treatment == "Sorghum"]       |> na.omit()
      w <- plot_mean_flux[Treatment == "Sorghum + WCC"] |> na.omit()
      if (length(s) >= 3 && length(w) >= 3)
        tryCatch(wilcox.test(s, w, exact = FALSE)$p.value, error = \(e) NA_real_)
      else NA_real_
    },
    .by = c(year, date)
  ) |>
  mutate(significant = !is.na(p_value) & p_value < 0.05)

star_data <- daily_flux |>
  summarise(y_top = max(mean_flux_g + se_flux_g, na.rm = TRUE), .by = c(year, date)) |>
  left_join(daily_significance, by = c("year", "date")) |>
  filter(significant) |>
  mutate(star_y = y_top * 1.1, star_label = "*")
```

```{r cumulative-flux}
# Plot-level cumulative N2O via linear interpolation
daily_flux_plot_level <- flux_subset |>
  summarise(daily_flux_kg = mean(gnha_day_no_negative, na.rm = TRUE) / 1000,
            .by = c(Treatment, year, plot, date))

cum_flux_plot_level <- daily_flux_plot_level |>
  arrange(Treatment, year, plot, date) |>
  group_by(Treatment, year, plot) |>
  group_modify(\(df, keys) {
    df <- df |>
      complete(date = seq(min(date), max(date), by = "day")) |>
      arrange(date)
    non_na <- !is.na(df$daily_flux_kg)
    if (sum(non_na) == 0) return(tibble())

    flux_interp <- if (sum(non_na) >= 2) {
      approx(df$date[non_na], df$daily_flux_kg[non_na],
             xout = df$date, rule = 1, ties = "ordered")$y
    } else {
      rep(df$daily_flux_kg[non_na][1], nrow(df))
    }

    df |>
      mutate(flux_interp = flux_interp, cum_flux_kg = cumsum(flux_interp)) |>
      filter(!is.na(flux_interp))
  }) |>
  ungroup()

cum_flux_summary <- cum_flux_plot_level |>
  summarise(
    cum_flux_mean = mean(cum_flux_kg, na.rm = TRUE),
    cum_flux_sd   = sd(cum_flux_kg, na.rm = TRUE),
    n_plots       = n(),
    cum_flux_se   = cum_flux_sd / sqrt(n_plots),
    t_crit        = if_else(n_plots > 1, qt(0.975, df = n_plots - 1), NA_real_),
    cum_flux_ci   = t_crit * cum_flux_se,
    .by = c(Treatment, year, date)
  )
```

```{r diff-and-annual}
# ── Daily difference (WCC - Sorghum) ──────────────────────────────────────
daily_n2o_diff <- daily_flux |>
  mutate(trt_key = if_else(Treatment == "Sorghum + WCC", "WCC", "Sorghum")) |>
  select(year, date, trt_key, mean_flux_g) |>
  pivot_wider(names_from = trt_key, values_from = mean_flux_g) |>
  mutate(diff_wcc_minus_sorghum = WCC - Sorghum)

one_sample_vs_zero <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, mu = 0)$p.value, error = \(e) NA_real_)
}

n2o_diff_stats <- daily_n2o_diff |>
  summarise(
    n          = sum(!is.na(diff_wcc_minus_sorghum)),
    mean_diff  = mean(diff_wcc_minus_sorghum, na.rm = TRUE),
    p_value    = one_sample_vs_zero(diff_wcc_minus_sorghum),
    significant = !is.na(p_value) & p_value < 0.05,
    letter     = if_else(significant, "a", NA_character_),
    .by = year
  )

# ── Annual cumulative N2O + one-way ANOVA ─────────────────────────────────
annual_n2o <- cum_flux_plot_level |>
  summarise(annual_cum_n2o = max(cum_flux_kg, na.rm = TRUE),
            .by = c(year, plot, Treatment)) |>
  filter(is.finite(annual_cum_n2o), annual_cum_n2o >= 0)

anova_n2o   <- aov(annual_cum_n2o ~ Treatment, data = annual_n2o)
anova_n2o_p <- summary(anova_n2o)[[1]]["Treatment", "Pr(>F)"]

n2o_letters <- tryCatch({
  emmeans(anova_n2o, ~ Treatment) |>
    cld(adjust = "tukey", Letters = letters) |>
    as_tibble() |>
    transmute(Treatment = as.character(Treatment),
              letter    = gsub("\\s+", "", .group))
}, error = \(e) {
  annual_n2o |> distinct(Treatment) |> mutate(letter = "a")
})
```

# Figure 4

```{r figure4}
#| fig-width: 12
#| fig-height: 8
#| fig-dpi: 300

# ── Precipitation trimmed to flux windows ──────────────────────────────────
flux_date_ranges <- daily_flux |>
  summarise(min_date = min(date), max_date = max(date), .by = year)

daily_precip_trimmed <- rain |>
  select(date, year, precipmm) |>
  filter(year %in% c(2023, 2024)) |>
  left_join(flux_date_ranges, by = "year") |>
  filter(date >= min_date, date <= max_date)

# ── Scaling constants ─────────────────────────────────────────────────────
global_max_flux_g     <- max(daily_flux$mean_flux_g + daily_flux$se_flux_g, na.rm = TRUE)
global_max_precip     <- max(daily_precip_trimmed$precipmm, na.rm = TRUE)
global_max_cum_flux   <- max(cum_flux_summary$cum_flux_mean + cum_flux_summary$cum_flux_ci, na.rm = TRUE)
precip_scale_factor   <- global_max_flux_g * 0.4 / global_max_precip
plot_top_flux         <- global_max_flux_g * 1.12

# ── Panel builder: daily N2O ──────────────────────────────────────────────
create_n2o_daily_panel <- function(year_val, is_right, tag_label) {
  flux_year   <- filter(daily_flux, year == year_val)
  precip_year <- filter(daily_precip_trimmed, year == year_val)
  star_year   <- filter(star_data, year == year_val)
  fert_year   <- filter(fert_events, year == year_val)
  x_lims      <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot() +
    geom_rect(
      data = precip_year,
      aes(xmin = date - 0.5, xmax = date + 0.5,
          ymin = plot_top_flux - precipmm * precip_scale_factor,
          ymax = plot_top_flux),
      fill = "#377EB8", alpha = 0.7
    ) +
    geom_point(data = flux_year,
               aes(x = date, y = mean_flux_g, color = Treatment), size = 1.4) +
    geom_errorbar(
      data = flux_year,
      aes(x = date,
          ymin = pmax(mean_flux_g - se_flux_g, 0),
          ymax = mean_flux_g + se_flux_g,
          color = Treatment),
      width = 0.2, alpha = 0.5
    ) +
    geom_vline(data = fert_year, aes(xintercept = fertilizer_date),
               linetype = "dashed", color = "black", alpha = 0.7) +
    geom_text(data = star_year,
              aes(x = date, y = star_y, label = star_label), size = 4.8) +
    scale_color_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right) expression("Daily N"[2] * "O Flux (g N ha"^-1 * " d"^-1 * ")") else "",
      limits = c(0, plot_top_flux),
      sec.axis = sec_axis(
        ~ (plot_top_flux - .) / precip_scale_factor,
        name = if (is_right) "Precipitation (mm)" else ""
      )
    ) +
    labs(x = "", tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(
      axis.text.x      = element_blank(),
      axis.ticks.x     = element_blank(),
      axis.line.x      = element_blank(),
      legend.position  = "none",
      plot.tag          = element_text(face = "bold", size = 11),
      plot.tag.position = "topleft"
    )

  if (is_right) {
    p <- p + theme(
      axis.text.y = element_blank(), axis.ticks.y = element_blank(),
      axis.line.y = element_blank(), axis.title.y = element_blank(),
      axis.text.y.right  = element_text(color = "#377EB8"),
      axis.ticks.y.right = element_line(),
      axis.title.y.right = element_text(color = "#377EB8")
    )
  } else {
    p <- p + theme(
      axis.text.y.right = element_blank(), axis.ticks.y.right = element_blank(),
      axis.title.y.right = element_blank()
    )
  }
  p
}

# ── Panel builder: cumulative N2O ─────────────────────────────────────────
create_n2o_cum_panel <- function(year_val, is_right, tag_label) {
  cum_year  <- filter(cum_flux_summary, year == year_val)
  fert_year <- filter(fert_events, year == year_val)
  x_lims    <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot(cum_year, aes(x = date, y = cum_flux_mean,
                             color = Treatment, fill = Treatment)) +
    geom_ribbon(aes(ymin = pmax(cum_flux_mean - cum_flux_ci, 0),
                    ymax = cum_flux_mean + cum_flux_ci),
                alpha = 0.15, linewidth = 0) +
    geom_line(linewidth = 1) +
    geom_vline(data = fert_year, aes(xintercept = fertilizer_date),
               linetype = "dashed", color = "black", alpha = 0.7) +
    scale_color_manual(values = treatment_colors) +
    scale_fill_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right) expression("Cumulative N"[2] * "O Flux (kg N ha"^-1 * ")") else "",
      limits = c(0, global_max_cum_flux)
    ) +
    labs(x = "Month", tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(legend.position = "none",
          plot.tag = element_text(face = "bold", size = 11),
          plot.tag.position = "topleft")

  if (is_right) {
    p <- p + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                   axis.line.y = element_blank(), axis.title.y = element_blank())
  }
  p
}

# ── Panel (c): daily difference boxplot ────────────────────────────────────
n2o_diff_span <- diff(range(daily_n2o_diff$diff_wcc_minus_sorghum, na.rm = TRUE))
if (!is.finite(n2o_diff_span) || n2o_diff_span == 0) n2o_diff_span <- 1

n2o_diff_anno <- daily_n2o_diff |>
  summarise(y = max(diff_wcc_minus_sorghum, na.rm = TRUE), .by = year) |>
  left_join(n2o_diff_stats, by = "year") |>
  filter(significant) |>
  mutate(y = y + n2o_diff_span * 0.08)

p4c <- ggplot(daily_n2o_diff, aes(x = factor(year), y = diff_wcc_minus_sorghum)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
  geom_jitter(aes(color = factor(year)), width = 0.12, alpha = 0.7,
              size = 1.4, show.legend = FALSE) +
  geom_text(data = n2o_diff_anno,
            aes(x = factor(year), y = y, label = letter),
            inherit.aes = FALSE, fontface = "bold", size = 3.8) +
  scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
  labs(
    x = "Year",
    y = expression("Daily N"[2] * "O Difference (g N ha"^-1 * " d"^-1 * ", WCC - Sorghum)"),
    tag = "(c)"
  ) +
  theme_publication(base_size = 10) +
  theme(plot.tag = element_text(face = "bold", size = 11),
        plot.tag.position = "topleft")

# ── Panel (f): annual cumulative by treatment ──────────────────────────────
n2o_letter_offset <- diff(range(annual_n2o$annual_cum_n2o, na.rm = TRUE))
if (!is.finite(n2o_letter_offset) || n2o_letter_offset == 0) n2o_letter_offset <- 0.05

n2o_letter_pos <- annual_n2o |>
  summarise(y = max(annual_cum_n2o, na.rm = TRUE), .by = Treatment) |>
  left_join(n2o_letters, by = "Treatment") |>
  mutate(y = y + n2o_letter_offset * 0.08)

p4f <- ggplot(annual_n2o, aes(x = Treatment, y = annual_cum_n2o, fill = Treatment)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.5, color = "black") +
  geom_jitter(aes(shape = factor(year)), width = 0.12, size = 1.8,
              alpha = 0.8, show.legend = FALSE) +
  geom_text(data = n2o_letter_pos,
            aes(x = Treatment, y = y, label = letter),
            inherit.aes = FALSE, fontface = "bold", size = 3.8) +
  scale_fill_manual(values = treatment_colors) +
  labs(
    x = "",
    y = expression("Annual Cumulative N"[2] * "O Flux (kg N ha"^-1 * ")"),
    subtitle = paste0("One-way ANOVA p = ",
                      ifelse(anova_n2o_p < 0.001, "< 0.001",
                             sprintf("%.3f", anova_n2o_p))),
    tag = "(f)"
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "none",
        plot.tag = element_text(face = "bold", size = 11),
        plot.tag.position = "topleft")

# ── Build panels ──────────────────────────────────────────────────────────
p4a <- create_n2o_daily_panel(2023, is_right = FALSE, "(a)")
p4b <- create_n2o_daily_panel(2024, is_right = TRUE,  "(b)")
p4d <- create_n2o_cum_panel(2023, is_right = FALSE, "(d)")
p4e <- create_n2o_cum_panel(2024, is_right = TRUE,  "(e)")

# ── Assemble ──────────────────────────────────────────────────────────────
figure4 <- (p4a | p4b | p4c) / (p4d | p4e | p4f) +
  plot_layout(guides = "collect", widths = c(1, 1, 0.95)) &
  theme(legend.position = "top", legend.justification = "center")

figure4
```

```{r print-stats}
#| echo: false
cat("Daily difference stats:\n")
n2o_diff_stats
cat("\nANOVA p-value:", anova_n2o_p, "\n")
```
