---
title: "Figure 2: Soil Microclimate"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 10
    fig-height: 10
execute:
  warning: false
  message: false
  echo: true
---

# Scientific Context

We monitored soil temperature and volumetric water content at 5 cm depth to
evaluate whether the winter cover crop altered the biophysical environment
experienced by soil microbes and plants. Differences in soil microclimate could
influence decomposition rates, nutrient mineralization, and crop performance.

This figure has six panels:

- **(a--b)** Soil temperature time series for 2023 and 2024
- **(c--d)** Soil moisture time series for 2023 and 2024
- **(e--f)** Boxplots of daily treatment differences (WCC minus Sorghum) for
  temperature and moisture

Colored status bars at the top of each time-series panel flag dates with
significant treatment differences.

# Setup

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(patchwork)
library(lubridate)

# ── Publication theme ──────────────────────────────────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(3, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "top",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}

treatment_colors <- c(
  "Sorghum" = "#D8D97AFF",
  "Sorghum + WCC" = "#95C36EFF"
)
```

# Statistical Approach

**Current method:** Per-date two-sample Welch's t-tests (Sorghum vs
Sorghum + WCC) with *p* < 0.05 to flag significant dates on the time-series
panels. The difference panels (e, f) use a one-sample t-test of daily
differences against zero within each year.

**Assessment:** The approach is reasonable for an exploratory visualization,
but has limitations:

1. **Multiple comparisons:** Running ~60--80 independent t-tests across dates
   inflates the family-wise error rate. No correction (e.g., Bonferroni,
   Benjamini--Hochberg) is applied, so the significant-date bars likely
   over-detect.
2. **Temporal autocorrelation:** Consecutive daily means from the same plots
   are not independent. A mixed model with plot as a random effect and a
   temporal correlation structure (e.g., AR(1)) would better account for this.

**Suggested alternatives:**

- **Linear mixed-effects model:** `lmer(value ~ Treatment * date + (1|plot))`
  with an AR(1) residual correlation. Test the Treatment main effect and
  Treatment-by-date interaction. This avoids the multiple-testing issue
  entirely and accounts for repeated measures.
- **If retaining per-date tests:** Apply Benjamini--Hochberg FDR correction
  across all dates within a year to control the false-discovery rate. This is a
  minimal fix that preserves the current visual style.

The current approach is retained below for consistency with the manuscript.

# Data Processing

```{r load-data}
flux_data <- read_csv("data/cleaned_flux_data.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date), year = year(date))

raw_soil <- flux_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    Treatment = factor(Treatment, levels = c("Sorghum", "Sorghum + WCC"))
  )
```

```{r compute-daily-means}
# Daily treatment means for time-series and difference panels
moist_temp <- raw_soil |>
  summarise(
    mean_soil_moisture = mean(swc_2_mean, na.rm = TRUE),
    mean_soil_temp     = mean(ts_2_mean, na.rm = TRUE),
    .by = c(Treatment, year, date)
  )

daily_diff <- moist_temp |>
  mutate(trt_key = if_else(Treatment == "Sorghum + WCC", "WCC", "Sorghum")) |>
  select(year, date, trt_key, mean_soil_temp, mean_soil_moisture) |>
  pivot_wider(names_from = trt_key, values_from = c(mean_soil_temp, mean_soil_moisture)) |>
  mutate(
    temp_diff  = mean_soil_temp_WCC  - mean_soil_temp_Sorghum,
    moist_diff = mean_soil_moisture_WCC - mean_soil_moisture_Sorghum
  )
```

```{r build-status-bars}
# Per-date significance bars (t-test, p < 0.05)
build_status_bars <- function(df, value_col) {
  df |>
    summarise(
      n_sorghum    = sum(Treatment == "Sorghum" & !is.na(.data[[value_col]])),
      n_wcc        = sum(Treatment == "Sorghum + WCC" & !is.na(.data[[value_col]])),
      mean_sorghum = mean(.data[[value_col]][Treatment == "Sorghum"], na.rm = TRUE),
      mean_wcc     = mean(.data[[value_col]][Treatment == "Sorghum + WCC"], na.rm = TRUE),
      p_value = {
        s <- .data[[value_col]][Treatment == "Sorghum"]   |> na.omit()
        w <- .data[[value_col]][Treatment == "Sorghum + WCC"] |> na.omit()
        if (length(s) >= 2 && length(w) >= 2)
          tryCatch(t.test(s, w)$p.value, error = \(e) NA_real_)
        else NA_real_
      },
      .by = c(year, date)
    ) |>
    mutate(
      has_testable = n_sorghum >= 2 & n_wcc >= 2,
      significant  = has_testable & !is.na(p_value) & p_value < 0.05,
      status = case_when(
        significant  ~ "Significant",
        has_testable ~ "No difference",
        TRUE         ~ "No data"
      ),
      greater = case_when(
        significant & mean_wcc > mean_sorghum  ~ "Sorghum + WCC",
        significant & mean_sorghum >= mean_wcc ~ "Sorghum",
        TRUE ~ NA_character_
      ),
      x    = date - 2,
      xend = date + 2
    )
}

temp_status  <- build_status_bars(raw_soil, "ts_2_mean")
moist_status <- build_status_bars(raw_soil, "swc_2_mean")

status_fill <- c(
  "Significant"   = "black",
  "No difference" = "grey65",
  "No data"       = "grey90"
)
```

```{r compute-diff-stats}
# One-sample t-test of daily differences against zero
one_sample_vs_zero <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, mu = 0)$p.value, error = \(e) NA_real_)
}

temp_diff_stats <- daily_diff |>
  summarise(
    n          = sum(!is.na(temp_diff)),
    mean_diff  = mean(temp_diff, na.rm = TRUE),
    p_value    = one_sample_vs_zero(temp_diff),
    significant = !is.na(p_value) & p_value < 0.05,
    letter     = if_else(significant, "a", NA_character_),
    .by = year
  )

moist_diff_stats <- daily_diff |>
  summarise(
    n          = sum(!is.na(moist_diff)),
    mean_diff  = mean(moist_diff, na.rm = TRUE),
    p_value    = one_sample_vs_zero(moist_diff),
    significant = !is.na(p_value) & p_value < 0.05,
    letter     = if_else(significant, "a", NA_character_),
    .by = year
  )
```

# Figure 2

```{r figure2}
#| fig-width: 10
#| fig-height: 10
#| fig-dpi: 300

# ── Global axis limits ────────────────────────────────────────────────────
global_moisture_limits <- range(moist_temp$mean_soil_moisture, na.rm = TRUE)
global_moisture_limits[2] <- global_moisture_limits[2] * 1.14

global_temp_limits <- range(moist_temp$mean_soil_temp, na.rm = TRUE)
global_temp_limits[2] <- max(40, global_temp_limits[2] * 1.08)

moisture_breaks <- pretty(global_moisture_limits, n = 5)
temp_breaks     <- pretty(global_temp_limits, n = 5)

# Status-bar vertical positions
temp_bar_y    <- global_temp_limits[2] - diff(global_temp_limits) * 0.035
temp_bar_half <- diff(global_temp_limits) * 0.01
moist_bar_y    <- global_moisture_limits[2] - diff(global_moisture_limits) * 0.035
moist_bar_half <- diff(global_moisture_limits) * 0.01

# ── Helper: time-series panel ─────────────────────────────────────────────
make_ts_panel <- function(data, status_df, y_var, y_limits, y_breaks,
                          bar_y, bar_half, y_lab, tag_label,
                          hide_x = FALSE, hide_y_left = FALSE) {
  year_val <- unique(data$year)
  x_lims   <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot(data, aes(x = date, y = .data[[y_var]],
                         color = Treatment, linetype = Treatment, shape = Treatment)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.5) +
    # Status bar background
    geom_rect(
      data = filter(status_df, year == year_val),
      aes(xmin = x, xmax = xend,
          ymin = bar_y - bar_half, ymax = bar_y + bar_half,
          fill = status),
      inherit.aes = FALSE, show.legend = FALSE
    ) +
    # Colored overlay on significant dates
    geom_segment(
      data = filter(status_df, year == year_val, significant),
      aes(x = x, xend = xend, y = bar_y, yend = bar_y, color = greater),
      linewidth = 3, inherit.aes = FALSE, show.legend = FALSE
    ) +
    scale_fill_manual(values = status_fill) +
    scale_color_manual(values = treatment_colors) +
    scale_linetype_manual(values = c("Sorghum" = "dashed", "Sorghum + WCC" = "solid")) +
    scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + WCC" = 15)) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(limits = y_limits, breaks = y_breaks,
                       expand = expansion(mult = c(0.04, 0.04))) +
    labs(x = if (hide_x) "" else "Month", y = y_lab, tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(plot.tag = element_text(face = "bold", size = 11),
          plot.tag.position = "topleft")

  if (hide_x) {
    p <- p + theme(axis.text.x = element_blank(),
                   axis.ticks.x = element_blank(),
                   axis.line.x = element_blank())
  }
  if (hide_y_left) {
    p <- p + theme(axis.text.y = element_blank(),
                   axis.ticks.y = element_blank(),
                   axis.line.y = element_blank()) +
      labs(y = "")
  }
  p + theme(legend.position = "none")
}

# ── Panels (a–d): time series ─────────────────────────────────────────────
moist_temp_2023 <- filter(moist_temp, year == 2023)
moist_temp_2024 <- filter(moist_temp, year == 2024)

p_a <- make_ts_panel(moist_temp_2023, temp_status,  "mean_soil_temp",
                     global_temp_limits, temp_breaks,
                     temp_bar_y, temp_bar_half,
                     "Soil Temperature (\u00B0C)", "(a)",
                     hide_x = TRUE)

p_b <- make_ts_panel(moist_temp_2024, temp_status,  "mean_soil_temp",
                     global_temp_limits, temp_breaks,
                     temp_bar_y, temp_bar_half,
                     "", "(b)",
                     hide_x = TRUE, hide_y_left = TRUE)

p_c <- make_ts_panel(moist_temp_2023, moist_status, "mean_soil_moisture",
                     global_moisture_limits, moisture_breaks,
                     moist_bar_y, moist_bar_half,
                     expression("Volumetric Soil Moisture (cm"^3 * " cm"^-3 * ")"),
                     "(c)")

p_d <- make_ts_panel(moist_temp_2024, moist_status, "mean_soil_moisture",
                     global_moisture_limits, moisture_breaks,
                     moist_bar_y, moist_bar_half,
                     "", "(d)",
                     hide_y_left = TRUE)

# ── Helper: difference boxplot ─────────────────────────────────────────────
make_diff_panel <- function(diff_col, diff_stats, y_lab, tag_label) {
  span <- diff(range(daily_diff[[diff_col]], na.rm = TRUE))
  if (!is.finite(span) || span == 0) span <- 1

  anno <- daily_diff |>
    summarise(y = max(.data[[diff_col]], na.rm = TRUE), .by = year) |>
    left_join(diff_stats, by = "year") |>
    filter(significant) |>
    mutate(y = y + span * 0.08)

  ggplot(daily_diff, aes(x = factor(year), y = .data[[diff_col]])) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
    geom_jitter(aes(color = factor(year)), width = 0.12, alpha = 0.7,
                size = 1.4, show.legend = FALSE) +
    geom_text(data = anno, aes(x = factor(year), y = y, label = letter),
              inherit.aes = FALSE, fontface = "bold", size = 3.8) +
    scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
    labs(x = "Year", y = y_lab, tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(plot.tag = element_text(face = "bold", size = 11),
          plot.tag.position = "topleft")
}

# ── Panels (e–f): difference boxplots ─────────────────────────────────────
p_e <- make_diff_panel("temp_diff",  temp_diff_stats,
                       expression("Daily Temperature Difference (\u00B0C, WCC - Sorghum)"),
                       "(e)")

p_f <- make_diff_panel("moist_diff", moist_diff_stats,
                       expression("Daily Moisture Difference (cm"^3 * " cm"^-3 * ", WCC - Sorghum)"),
                       "(f)")

# ── Assemble ───────────────────────────────────────────────────────────────
figure2 <- (p_a | p_b) / (p_c | p_d) / (p_e | p_f) +
  plot_layout(guides = "collect", heights = c(1, 1, 0.95)) &
  theme(legend.position = "top") &
  guides(
    color    = guide_legend(nrow = 1, byrow = TRUE),
    linetype = guide_legend(nrow = 1, byrow = TRUE),
    shape    = guide_legend(nrow = 1, byrow = TRUE)
  )

figure2 <- figure2 +
  plot_annotation(
    caption = "Status bars above time-series panels: black = significant treatment difference; gray = no difference; light gray = no data.\nColored overlays indicate which treatment had the greater daily mean on significant dates."
  )

figure2
```

```{r diff-stats-table}
#| echo: false
cat("Temperature difference stats:\n")
temp_diff_stats
cat("\nMoisture difference stats:\n")
moist_diff_stats
```
