---
title: "Figure 2: Soil Microclimate"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 10
    fig-height: 10
execute:
  warning: false
  message: false
  echo: true
---

# Scientific Context

We monitored soil temperature and volumetric water content at 5 cm depth to
evaluate whether the winter cover crop altered the biophysical environment
experienced by soil microbes and plants. Differences in soil microclimate could
influence decomposition rates, nutrient mineralization, and crop performance.

This figure has six panels:

- **(a--b)** Soil temperature time series for 2023 and 2024
- **(c--d)** Soil moisture time series for 2023 and 2024
- **(e--f)** Boxplots of daily treatment differences (WCC minus Sorghum) for
  temperature and moisture

Colored status bars at the top of each time-series panel flag dates with
significant treatment differences.

# Setup

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(patchwork)
library(lubridate)

# ── Publication theme ──────────────────────────────────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(3, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "top",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}

treatment_colors <- c(
  "Sorghum" = "#D8D97AFF",
  "Sorghum + WCC" = "#95C36EFF"
)
```

# Statistical Approach

**Current method:** Per-date two-sample Welch's t-tests (Sorghum vs
Sorghum + WCC) with *p* < 0.05 to flag significant dates on the time-series
panels. The difference panels (e, f) use a one-sample t-test of daily
differences against zero within each year.

**Assessment:** The approach is reasonable for an exploratory visualization,
but has limitations:

1. **Multiple comparisons:** Running ~60--80 independent t-tests across dates
   inflates the family-wise error rate. No correction (e.g., Bonferroni,
   Benjamini--Hochberg) is applied, so the significant-date bars likely
   over-detect.
2. **Temporal autocorrelation:** Consecutive daily means from the same plots
   are not independent. A mixed model with plot as a random effect and a
   temporal correlation structure (e.g., AR(1)) would better account for this.

**Suggested alternatives:**

- **Linear mixed-effects model:** `lmer(value ~ Treatment * date + (1|plot))`
  with an AR(1) residual correlation. Test the Treatment main effect and
  Treatment-by-date interaction. This avoids the multiple-testing issue
  entirely and accounts for repeated measures.
- **If retaining per-date tests:** Apply Benjamini--Hochberg FDR correction
  across all dates within a year to control the false-discovery rate. This is a
  minimal fix that preserves the current visual style.

The current approach is retained below for consistency with the manuscript.

# Data Processing

```{r load-data}
flux_data <- read_csv("data/cleaned_flux_data.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date), year = year(date))

raw_soil <- flux_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    Treatment = factor(Treatment, levels = c("Sorghum", "Sorghum + WCC"))
  )
```

```{r compute-daily-means}
# Daily treatment means for time-series and difference panels
moist_temp <- raw_soil |>
  summarise(
    mean_soil_moisture = mean(swc_2_mean, na.rm = TRUE),
    mean_soil_temp     = mean(ts_2_mean, na.rm = TRUE),
    .by = c(Treatment, year, date)
  )

daily_diff <- moist_temp |>
  mutate(trt_key = if_else(Treatment == "Sorghum + WCC", "WCC", "Sorghum")) |>
  select(year, date, trt_key, mean_soil_temp, mean_soil_moisture) |>
  pivot_wider(names_from = trt_key, values_from = c(mean_soil_temp, mean_soil_moisture)) |>
  mutate(
    temp_diff  = mean_soil_temp_WCC  - mean_soil_temp_Sorghum,
    moist_diff = mean_soil_moisture_WCC - mean_soil_moisture_Sorghum
  )
```

```{r build-status-bars}
# Per-date significance bars (t-test, p < 0.05)
build_status_bars <- function(df, value_col) {
  df |>
    summarise(
      n_sorghum    = sum(Treatment == "Sorghum" & !is.na(.data[[value_col]])),
      n_wcc        = sum(Treatment == "Sorghum + WCC" & !is.na(.data[[value_col]])),
      mean_sorghum = mean(.data[[value_col]][Treatment == "Sorghum"], na.rm = TRUE),
      mean_wcc     = mean(.data[[value_col]][Treatment == "Sorghum + WCC"], na.rm = TRUE),
      p_value = {
        s <- .data[[value_col]][Treatment == "Sorghum"]   |> na.omit()
        w <- .data[[value_col]][Treatment == "Sorghum + WCC"] |> na.omit()
        if (length(s) >= 2 && length(w) >= 2)
          tryCatch(t.test(s, w)$p.value, error = \(e) NA_real_)
        else NA_real_
      },
      .by = c(year, date)
    ) |>
    mutate(
      has_testable = n_sorghum >= 2 & n_wcc >= 2,
      significant  = has_testable & !is.na(p_value) & p_value < 0.05,
      status = case_when(
        significant  ~ "Significant",
        has_testable ~ "No difference",
        TRUE         ~ "No data"
      ),
      greater = case_when(
        significant & mean_wcc > mean_sorghum  ~ "Sorghum + WCC",
        significant & mean_sorghum >= mean_wcc ~ "Sorghum",
        TRUE ~ NA_character_
      ),
      x    = date - 2,
      xend = date + 2
    )
}

temp_status  <- build_status_bars(raw_soil, "ts_2_mean")
moist_status <- build_status_bars(raw_soil, "swc_2_mean")

status_fill <- c(
  "Significant"   = "black",
  "No difference" = "grey65",
  "No data"       = "grey90"
)
```

```{r compute-diff-stats}
# One-sample t-test of daily differences against zero
one_sample_vs_zero <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, mu = 0)$p.value, error = \(e) NA_real_)
}

temp_diff_stats <- daily_diff |>
  summarise(
    n          = sum(!is.na(temp_diff)),
    mean_diff  = mean(temp_diff, na.rm = TRUE),
    p_value    = one_sample_vs_zero(temp_diff),
    significant = !is.na(p_value) & p_value < 0.05,
    letter     = if_else(significant, "a", NA_character_),
    .by = year
  )

moist_diff_stats <- daily_diff |>
  summarise(
    n          = sum(!is.na(moist_diff)),
    mean_diff  = mean(moist_diff, na.rm = TRUE),
    p_value    = one_sample_vs_zero(moist_diff),
    significant = !is.na(p_value) & p_value < 0.05,
    letter     = if_else(significant, "a", NA_character_),
    .by = year
  )
```

# Figure 2

```{r figure2}
#| fig-width: 10
#| fig-height: 10
#| fig-dpi: 300

# ── Global axis limits ────────────────────────────────────────────────────
global_moisture_limits <- range(moist_temp$mean_soil_moisture, na.rm = TRUE)
global_moisture_limits[2] <- global_moisture_limits[2] * 1.14

global_temp_limits <- range(moist_temp$mean_soil_temp, na.rm = TRUE)
global_temp_limits[2] <- max(40, global_temp_limits[2] * 1.08)

moisture_breaks <- pretty(global_moisture_limits, n = 5)
temp_breaks     <- pretty(global_temp_limits, n = 5)

# Status-bar vertical positions
temp_bar_y    <- global_temp_limits[2] - diff(global_temp_limits) * 0.035
temp_bar_half <- diff(global_temp_limits) * 0.01
moist_bar_y    <- global_moisture_limits[2] - diff(global_moisture_limits) * 0.035
moist_bar_half <- diff(global_moisture_limits) * 0.01

# ── Helper: time-series panel ─────────────────────────────────────────────
make_ts_panel <- function(data, status_df, y_var, y_limits, y_breaks,
                          bar_y, bar_half, y_lab, tag_label,
                          hide_x = FALSE, hide_y_left = FALSE) {
  year_val <- unique(data$year)
  x_lims   <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot(data, aes(x = date, y = .data[[y_var]],
                         color = Treatment, linetype = Treatment, shape = Treatment)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.5) +
    # Status bar background
    geom_rect(
      data = filter(status_df, year == year_val),
      aes(xmin = x, xmax = xend,
          ymin = bar_y - bar_half, ymax = bar_y + bar_half,
          fill = status),
      inherit.aes = FALSE, show.legend = FALSE
    ) +
    # Colored overlay on significant dates
    geom_segment(
      data = filter(status_df, year == year_val, significant),
      aes(x = x, xend = xend, y = bar_y, yend = bar_y, color = greater),
      linewidth = 3, inherit.aes = FALSE, show.legend = FALSE
    ) +
    scale_fill_manual(values = status_fill) +
    scale_color_manual(values = treatment_colors) +
    scale_linetype_manual(values = c("Sorghum" = "dashed", "Sorghum + WCC" = "solid")) +
    scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + WCC" = 15)) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(limits = y_limits, breaks = y_breaks,
                       expand = expansion(mult = c(0.04, 0.04))) +
    labs(x = if (hide_x) "" else "Month", y = y_lab, tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(plot.tag = element_text(face = "bold", size = 11),
          plot.tag.position = "topleft")

  if (hide_x) {
    p <- p + theme(axis.text.x = element_blank(),
                   axis.ticks.x = element_blank(),
                   axis.line.x = element_blank())
  }
  if (hide_y_left) {
    p <- p + theme(axis.text.y = element_blank(),
                   axis.ticks.y = element_blank(),
                   axis.line.y = element_blank()) +
      labs(y = "")
  }
  p + theme(legend.position = "none")
}

# ── Panels (a–d): time series ─────────────────────────────────────────────
moist_temp_2023 <- filter(moist_temp, year == 2023)
moist_temp_2024 <- filter(moist_temp, year == 2024)

p_a <- make_ts_panel(moist_temp_2023, temp_status,  "mean_soil_temp",
                     global_temp_limits, temp_breaks,
                     temp_bar_y, temp_bar_half,
                     "Soil Temperature (\u00B0C)", "(a)",
                     hide_x = TRUE)

p_b <- make_ts_panel(moist_temp_2024, temp_status,  "mean_soil_temp",
                     global_temp_limits, temp_breaks,
                     temp_bar_y, temp_bar_half,
                     "", "(b)",
                     hide_x = TRUE, hide_y_left = TRUE)

p_c <- make_ts_panel(moist_temp_2023, moist_status, "mean_soil_moisture",
                     global_moisture_limits, moisture_breaks,
                     moist_bar_y, moist_bar_half,
                     expression("Volumetric Soil Moisture (cm"^3 * " cm"^-3 * ")"),
                     "(c)")

p_d <- make_ts_panel(moist_temp_2024, moist_status, "mean_soil_moisture",
                     global_moisture_limits, moisture_breaks,
                     moist_bar_y, moist_bar_half,
                     "", "(d)",
                     hide_y_left = TRUE)

# ── Helper: difference boxplot ─────────────────────────────────────────────
make_diff_panel <- function(diff_col, diff_stats, y_lab, tag_label) {
  span <- diff(range(daily_diff[[diff_col]], na.rm = TRUE))
  if (!is.finite(span) || span == 0) span <- 1

  anno <- daily_diff |>
    summarise(y = max(.data[[diff_col]], na.rm = TRUE), .by = year) |>
    left_join(diff_stats, by = "year") |>
    filter(significant) |>
    mutate(y = y + span * 0.08)

  ggplot(daily_diff, aes(x = factor(year), y = .data[[diff_col]])) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
    geom_jitter(aes(color = factor(year)), width = 0.12, alpha = 0.7,
                size = 1.4, show.legend = FALSE) +
    geom_text(data = anno, aes(x = factor(year), y = y, label = letter),
              inherit.aes = FALSE, fontface = "bold", size = 3.8) +
    scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
    labs(x = "Year", y = y_lab, tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(plot.tag = element_text(face = "bold", size = 11),
          plot.tag.position = "topleft")
}

# ── Panels (e–f): difference boxplots ─────────────────────────────────────
p_e <- make_diff_panel("temp_diff",  temp_diff_stats,
                       expression("Daily Temperature Difference (\u00B0C, WCC - Sorghum)"),
                       "(e)")

p_f <- make_diff_panel("moist_diff", moist_diff_stats,
                       expression("Daily Moisture Difference (cm"^3 * " cm"^-3 * ", WCC - Sorghum)"),
                       "(f)")

# ── Assemble ───────────────────────────────────────────────────────────────
figure2 <- (p_a | p_b) / (p_c | p_d) / (p_e | p_f) +
  plot_layout(guides = "collect", heights = c(1, 1, 0.95)) &
  theme(legend.position = "top") &
  guides(
    color    = guide_legend(nrow = 1, byrow = TRUE),
    linetype = guide_legend(nrow = 1, byrow = TRUE),
    shape    = guide_legend(nrow = 1, byrow = TRUE)
  )

figure2 <- figure2 +
  plot_annotation(
    caption = "Status bars above time-series panels: black = significant treatment difference; gray = no difference; light gray = no data.\nColored overlays indicate which treatment had the greater daily mean on significant dates."
  )

figure2

ggsave(filename = "figure_2_microclimate.tiff", plot = figure2, width = 174, height = 174, units = "mm", dpi = 600, device = "tiff", compression = "lzw")
```

```{r diff-stats-table}
#| echo: false
cat("Temperature difference stats:\n")
temp_diff_stats
cat("\nMoisture difference stats:\n")
moist_diff_stats
```

# Alternative Statistical Approaches

## Approach B: BH FDR-Corrected Per-Date T-Tests

Apply Benjamini–Hochberg FDR correction to the per-date t-tests produced by
`build_status_bars`. This controls the false-discovery rate across all dates
within each year.

```{r approach-b-fdr}
#| message: false
#| warning: false

# ── Apply BH correction to temperature status bars ──────────────────────
temp_status_fdr <- temp_status |>
  group_by(year) |>
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    significant_fdr = has_testable & !is.na(p_adj) & p_adj < 0.05,
    status_fdr = case_when(
      significant_fdr  ~ "Significant",
      has_testable     ~ "No difference",
      TRUE             ~ "No data"
    ),
    greater_fdr = case_when(
      significant_fdr & mean_wcc > mean_sorghum  ~ "Sorghum + WCC",
      significant_fdr & mean_sorghum >= mean_wcc ~ "Sorghum",
      TRUE ~ NA_character_
    )
  ) |>
  ungroup()

# ── Apply BH correction to moisture status bars ─────────────────────────
moist_status_fdr <- moist_status |>
  group_by(year) |>
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    significant_fdr = has_testable & !is.na(p_adj) & p_adj < 0.05,
    status_fdr = case_when(
      significant_fdr  ~ "Significant",
      has_testable     ~ "No difference",
      TRUE             ~ "No data"
    ),
    greater_fdr = case_when(
      significant_fdr & mean_wcc > mean_sorghum  ~ "Sorghum + WCC",
      significant_fdr & mean_sorghum >= mean_wcc ~ "Sorghum",
      TRUE ~ NA_character_
    )
  ) |>
  ungroup()

# ── Comparison: uncorrected vs FDR-corrected significant dates ───────────
fdr_comparison <- bind_rows(
  temp_status |>
    group_by(year) |>
    summarise(n_sig_uncorrected = sum(significant, na.rm = TRUE), .groups = "drop") |>
    left_join(
      temp_status_fdr |>
        group_by(year) |>
        summarise(n_sig_FDR = sum(significant_fdr, na.rm = TRUE), .groups = "drop"),
      by = "year"
    ) |>
    mutate(variable = "Temperature"),
  moist_status |>
    group_by(year) |>
    summarise(n_sig_uncorrected = sum(significant, na.rm = TRUE), .groups = "drop") |>
    left_join(
      moist_status_fdr |>
        group_by(year) |>
        summarise(n_sig_FDR = sum(significant_fdr, na.rm = TRUE), .groups = "drop"),
      by = "year"
    ) |>
    mutate(variable = "Moisture")
) |>
  select(year, variable, n_sig_uncorrected, n_sig_FDR)

cat("Comparison of significant dates: uncorrected vs BH FDR-corrected\n")
knitr::kable(fdr_comparison, align = "c")
```

```{r approach-b-fdr-figure}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7
#| fig-dpi: 300

# ── Helper for FDR-corrected time-series panels ─────────────────────────
make_ts_panel_fdr <- function(data, status_fdr_df, y_var, y_limits, y_breaks,
                              bar_y, bar_half, y_lab, tag_label,
                              hide_x = FALSE, hide_y_left = FALSE) {
  year_val <- unique(data$year)
  x_lims   <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot(data, aes(x = date, y = .data[[y_var]],
                         color = Treatment, linetype = Treatment, shape = Treatment)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.5) +
    # FDR-corrected status bar background
    geom_rect(
      data = filter(status_fdr_df, year == year_val),
      aes(xmin = x, xmax = xend,
          ymin = bar_y - bar_half, ymax = bar_y + bar_half,
          fill = status_fdr),
      inherit.aes = FALSE, show.legend = FALSE
    ) +
    # Colored overlay on FDR-significant dates
    geom_segment(
      data = filter(status_fdr_df, year == year_val, significant_fdr),
      aes(x = x, xend = xend, y = bar_y, yend = bar_y, color = greater_fdr),
      linewidth = 3, inherit.aes = FALSE, show.legend = FALSE
    ) +
    scale_fill_manual(values = status_fill) +
    scale_color_manual(values = treatment_colors) +
    scale_linetype_manual(values = c("Sorghum" = "dashed", "Sorghum + WCC" = "solid")) +
    scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + WCC" = 15)) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(limits = y_limits, breaks = y_breaks,
                       expand = expansion(mult = c(0.04, 0.04))) +
    labs(x = if (hide_x) "" else "Month", y = y_lab, tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(plot.tag = element_text(face = "bold", size = 11),
          plot.tag.position = "topleft")

  if (hide_x) {
    p <- p + theme(axis.text.x = element_blank(),
                   axis.ticks.x = element_blank(),
                   axis.line.x = element_blank())
  }
  if (hide_y_left) {
    p <- p + theme(axis.text.y = element_blank(),
                   axis.ticks.y = element_blank(),
                   axis.line.y = element_blank()) +
      labs(y = "")
  }
  p + theme(legend.position = "none")
}

# ── FDR-corrected panels (a-d) ──────────────────────────────────────────
p_a_fdr <- make_ts_panel_fdr(moist_temp_2023, temp_status_fdr, "mean_soil_temp",
                              global_temp_limits, temp_breaks,
                              temp_bar_y, temp_bar_half,
                              "Soil Temperature (\u00B0C)", "(a)",
                              hide_x = TRUE)

p_b_fdr <- make_ts_panel_fdr(moist_temp_2024, temp_status_fdr, "mean_soil_temp",
                              global_temp_limits, temp_breaks,
                              temp_bar_y, temp_bar_half,
                              "", "(b)",
                              hide_x = TRUE, hide_y_left = TRUE)

p_c_fdr <- make_ts_panel_fdr(moist_temp_2023, moist_status_fdr, "mean_soil_moisture",
                              global_moisture_limits, moisture_breaks,
                              moist_bar_y, moist_bar_half,
                              expression("Volumetric Soil Moisture (cm"^3 * " cm"^-3 * ")"),
                              "(c)")

p_d_fdr <- make_ts_panel_fdr(moist_temp_2024, moist_status_fdr, "mean_soil_moisture",
                              global_moisture_limits, moisture_breaks,
                              moist_bar_y, moist_bar_half,
                              "", "(d)",
                              hide_y_left = TRUE)

figure2_fdr <- (p_a_fdr | p_b_fdr) / (p_c_fdr | p_d_fdr) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top") &
  guides(
    color    = guide_legend(nrow = 1, byrow = TRUE),
    linetype = guide_legend(nrow = 1, byrow = TRUE),
    shape    = guide_legend(nrow = 1, byrow = TRUE)
  )

figure2_fdr <- figure2_fdr +
  plot_annotation(
    title   = "BH FDR-Corrected Status Bars",
    caption = "Status bars use BH-adjusted p-values (FDR < 0.05)."
  )

figure2_fdr
```

## Approach C: Period-Aggregated One-Sample T-Tests

Aggregate daily treatment differences into seasonal windows (Spring, Summer,
Fall, Winter), then test each period's mean difference against zero.

```{r approach-c-seasonal}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6
#| fig-dpi: 300

# ── Assign seasons ───────────────────────────────────────────────────────
daily_diff_season <- daily_diff |>
  mutate(
    month = month(date),
    season = case_when(
      month %in% 3:5   ~ "Spring",
      month %in% 6:8   ~ "Summer",
      month %in% 9:11  ~ "Fall",
      month %in% c(12, 1, 2) ~ "Winter"
    ),
    season = factor(season, levels = c("Spring", "Summer", "Fall", "Winter"))
  )

# ── One-sample t-tests by year x season for temperature ──────────────────
seasonal_temp_tests <- daily_diff_season |>
  group_by(year, season) |>
  summarise(
    n         = sum(!is.finite(temp_diff) == FALSE),
    mean_diff = mean(temp_diff, na.rm = TRUE),
    se_diff   = sd(temp_diff, na.rm = TRUE) / sqrt(n()),
    p_value   = if (sum(is.finite(temp_diff)) >= 2)
                  tryCatch(t.test(temp_diff[is.finite(temp_diff)], mu = 0)$p.value,
                           error = \(e) NA_real_)
                else NA_real_,
    .groups = "drop"
  ) |>
  mutate(
    significant = !is.na(p_value) & p_value < 0.05,
    sig_label   = if_else(significant, "*", ""),
    variable    = "Temperature (\u00B0C)"
  )

# ── One-sample t-tests by year x season for moisture ─────────────────────
seasonal_moist_tests <- daily_diff_season |>
  group_by(year, season) |>
  summarise(
    n         = sum(!is.finite(moist_diff) == FALSE),
    mean_diff = mean(moist_diff, na.rm = TRUE),
    se_diff   = sd(moist_diff, na.rm = TRUE) / sqrt(n()),
    p_value   = if (sum(is.finite(moist_diff)) >= 2)
                  tryCatch(t.test(moist_diff[is.finite(moist_diff)], mu = 0)$p.value,
                           error = \(e) NA_real_)
                else NA_real_,
    .groups = "drop"
  ) |>
  mutate(
    significant = !is.na(p_value) & p_value < 0.05,
    sig_label   = if_else(significant, "*", ""),
    variable    = "Moisture (cm\u00B3 cm\u207B\u00B3)"
  )

# ── Combined table ───────────────────────────────────────────────────────
seasonal_all <- bind_rows(seasonal_temp_tests, seasonal_moist_tests)

cat("Seasonal one-sample t-test results (daily diff vs 0):\n")
knitr::kable(
  seasonal_all |> select(year, season, variable, n, mean_diff, se_diff, p_value, sig_label),
  digits = 4, align = "c"
)

# ── Point plot with error bars ───────────────────────────────────────────
p_seasonal <- ggplot(seasonal_all,
                     aes(x = season, y = mean_diff, color = factor(year))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_pointrange(aes(ymin = mean_diff - se_diff, ymax = mean_diff + se_diff),
                  position = position_dodge(width = 0.4), size = 0.6) +
  geom_text(aes(label = sig_label, y = mean_diff + se_diff + 0.02),
            position = position_dodge(width = 0.4), size = 5, show.legend = FALSE,
            fontface = "bold", vjust = 0) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8"), name = "Year") +
  labs(
    x = "Season",
    y = "Mean Daily Difference (WCC \u2212 Sorghum) \u00B1 SE",
    title = "Period-Aggregated One-Sample T-Tests",
    caption = "* = p < 0.05 (one-sample t-test of daily differences against zero)"
  ) +
  theme_publication(base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

p_seasonal
```

## Approach D: Linear Mixed Model

Fit linear mixed-effects models to test Treatment effects while accounting for
repeated measures across plots.

```{r approach-d-lmm}
#| message: false
#| warning: false

library(lme4)
library(emmeans)

# Ensure year is a factor for mixed model fitting
raw_soil <- raw_soil |>
  mutate(year_fct = factor(year))

# ── Temperature model ────────────────────────────────────────────────────
lmm_temp <- lmer(ts_2_mean ~ Treatment * year_fct + (1 | plot),
                 data = raw_soil)

cat("===== Temperature Mixed Model Summary =====\n")
print(summary(lmm_temp))

cat("\n===== Temperature ANOVA (Type III) =====\n")
anova_temp <- anova(lmm_temp)
print(anova_temp)

cat("\n===== Temperature EMMs: Treatment Pairwise Comparisons =====\n")
emm_temp <- emmeans(lmm_temp, pairwise ~ Treatment)
print(emm_temp)

cat("\n===== Temperature EMMs: Treatment x Year =====\n")
emm_temp_yr <- emmeans(lmm_temp, pairwise ~ Treatment | year_fct)
print(emm_temp_yr)

# ── Moisture model ───────────────────────────────────────────────────────
lmm_moist <- lmer(swc_2_mean ~ Treatment * year_fct + (1 | plot),
                  data = raw_soil)

cat("\n===== Moisture Mixed Model Summary =====\n")
print(summary(lmm_moist))

cat("\n===== Moisture ANOVA (Type III) =====\n")
anova_moist <- anova(lmm_moist)
print(anova_moist)

cat("\n===== Moisture EMMs: Treatment Pairwise Comparisons =====\n")
emm_moist <- emmeans(lmm_moist, pairwise ~ Treatment)
print(emm_moist)

cat("\n===== Moisture EMMs: Treatment x Year =====\n")
emm_moist_yr <- emmeans(lmm_moist, pairwise ~ Treatment | year_fct)
print(emm_moist_yr)
```

## Statistical Comparison

```{r stat-comparison}
#| message: false
#| warning: false

# ── Gather results from each approach ────────────────────────────────────

# Approach A: uncorrected per-date significant counts
a_temp <- sum(temp_status$significant, na.rm = TRUE)
a_moist <- sum(moist_status$significant, na.rm = TRUE)

# Approach B: FDR-corrected per-date significant counts
b_temp <- sum(temp_status_fdr$significant_fdr, na.rm = TRUE)
b_moist <- sum(moist_status_fdr$significant_fdr, na.rm = TRUE)

# Approach C: number of significant seasons
c_temp <- sum(seasonal_temp_tests$significant, na.rm = TRUE)
c_moist <- sum(seasonal_moist_tests$significant, na.rm = TRUE)

# Approach D: Treatment p-values from ANOVA
# lmerTest provides "Pr(>F)"; without it the column may be "Pr(>Chisq)" or absent
extract_p <- function(anova_tbl, term = "Treatment") {
  tryCatch({
    row <- anova_tbl[term, ]
    p_col <- intersect(c("Pr(>F)", "Pr(>Chisq)"), names(row))
    if (length(p_col) > 0) row[[p_col[1]]] else NA_real_
  }, error = function(e) NA_real_)
}
d_temp_p <- extract_p(anova_temp)
if (is.null(d_temp_p) || length(d_temp_p) == 0) d_temp_p <- NA_real_
d_moist_p <- extract_p(anova_moist)
if (is.null(d_moist_p) || length(d_moist_p) == 0) d_moist_p <- NA_real_

comparison_table <- tibble::tibble(
  Approach = c(
    "A: Per-date t-tests (uncorrected)",
    "B: Per-date t-tests (BH FDR)",
    "C: Seasonal aggregated t-tests",
    "D: Linear mixed model"
  ),
  Temperature = c(
    paste0(a_temp, " significant dates"),
    paste0(b_temp, " significant dates"),
    paste0(c_temp, " significant seasons (of ", nrow(seasonal_temp_tests), ")"),
    paste0("Treatment p = ", formatC(d_temp_p, format = "g", digits = 4))
  ),
  Moisture = c(
    paste0(a_moist, " significant dates"),
    paste0(b_moist, " significant dates"),
    paste0(c_moist, " significant seasons (of ", nrow(seasonal_moist_tests), ")"),
    paste0("Treatment p = ", formatC(d_moist_p, format = "g", digits = 4))
  )
)

cat("===== Statistical Comparison Across Approaches =====\n\n")
knitr::kable(comparison_table, align = "lll")
```
