---
title: "Figure 3: Nitrate Leaching"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
  echo: true
---

# Scientific Context

Nitrate leaching via subsurface tile drainage represents a primary nitrogen
loss pathway in agricultural systems. This figure shows daily and cumulative
nitrogen losses through tile drainage, demonstrating the key finding that the
winter cover crop substantially reduced nitrogen leaching compared to the
no-cover control.

The figure has six panels arranged in a 2 x 3 grid:

- **(a--b)** Daily leached N with precipitation overlay (2023, 2024)
- **(c)** Daily treatment difference boxplot (WCC minus Sorghum) by year
- **(d--e)** Cumulative leached N (2023, 2024)
- **(f)** Annual cumulative leached N by treatment (years combined)

# Setup

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(patchwork)
library(lubridate)
library(janitor)
library(emmeans)
library(multcomp)

# ── Publication theme ──────────────────────────────────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(3, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "top",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}

treatment_colors <- c(
  "Sorghum"       = "#D8D97AFF",
  "Sorghum + Rye" = "#95C36EFF",
  "Sorghum + WCC" = "#95C36EFF",
  "Corn"          = "#74C8C3FF",
  "Soy"           = "#0A2E57FF"
)
```

# Statistical Approach

**Current methods:**

1. **Per-date Wilcoxon rank-sum tests** (*p* < 0.05) to place asterisks on
   dates with significant treatment differences in daily leaching.
2. **One-sample t-test** of daily differences (WCC minus Sorghum) against zero
   for the boxplot panel (c).
3. **One-way ANOVA** with Tukey-adjusted pairwise comparisons (via `emmeans`)
   on annual cumulative leaching, pooling both years, for panel (f).

**Assessment:**

- **Per-date Wilcoxon tests (panels a--b):** Same multiple-comparison concern
  as Figure 2. With only ~3--4 plots per treatment on any given date, the
  Wilcoxon test has very low power (max test statistic with n = 4 vs n = 4 can
  only achieve *p* ~ 0.03 under perfect separation). Consider whether the
  asterisks overstate certainty on individual dates.

- **Annual ANOVA (panel f):** Pooling years and ignoring the year factor treats
  2023 and 2024 observations as exchangeable replicates, which inflates the
  effective sample size. A mixed model with year as a random or fixed effect
  would be more defensible.

**Suggested alternatives:**

- **For panels a--b:** Wilcoxon with Benjamini--Hochberg FDR correction across
  dates, or replace per-date tests with a cumulative divergence test (e.g.,
  permutation test on the area between cumulative curves).
- **For panel f:** A two-way ANOVA or mixed model
  `lmer(annual_N ~ Treatment + (1|year))` that properly accounts for the
  year structure. Given only 2 years, treating year as fixed
  (`annual_N ~ Treatment + year`) may be more appropriate than random.
- **Log transformation:** Leaching data are right-skewed. A log or
  square-root transformation (or a generalized linear model with Gamma family)
  would better meet normality assumptions for ANOVA.

# Data Processing

```{r load-data}
# Constants
plot_area_ft2 <- 120 * 160
plot_area_ha  <- plot_area_ft2 / 107639.1041671

fert_events <- tibble(
  year            = c(2023, 2024),
  fertilizer_date = as.Date(c("2023-05-05", "2024-07-17"))
)

leaching_data <- read_csv("data/clean_leaching.csv", show_col_types = FALSE) |>
  clean_names() |>
  mutate(date = as.Date(date), year = year(date)) |>
  filter(year %in% c(2023, 2024))

# Precipitation (for overlay bars)
rain <- read_xlsx("data/rain/historical_rain.xlsx") |>
  select(day, doy, precipmm) |>
  mutate(date = as.Date(day), year = year(date))
```

```{r compute-summaries}
# ── Daily leaching summary ─────────────────────────────────────────────────
daily_leaching_summary <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    daily_leached_n_kgha = (total_n_loss_mg / 1e6) / plot_area_ha
  ) |>
  summarise(
    mean_leached = mean(daily_leached_n_kgha, na.rm = TRUE),
    se_leached   = sd(daily_leached_n_kgha, na.rm = TRUE) /
                   sqrt(sum(!is.na(daily_leached_n_kgha))),
    .by = c(Treatment, year, date)
  )

# ── Cumulative leaching summary ───────────────────────────────────────────
cum_leaching_summary <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    cumulative_leached_n_kgha = (cumulative_n_loss_mg / 1e6) / plot_area_ha
  ) |>
  summarise(
    cum_leached_mean = mean(cumulative_leached_n_kgha, na.rm = TRUE),
    cum_leached_sd   = sd(cumulative_leached_n_kgha, na.rm = TRUE),
    n                = n(),
    cum_leached_se   = cum_leached_sd / sqrt(n),
    .by = c(Treatment, year, date)
  )
```

```{r significance-tests}
# ── Per-date Wilcoxon tests (asterisks) ────────────────────────────────────
daily_leaching_sig <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    daily_leached_n_kgha = (total_n_loss_mg / 1e6) / plot_area_ha
  ) |>
  summarise(
    p_value = {
      s <- daily_leached_n_kgha[Treatment == "Sorghum"]       |> na.omit()
      w <- daily_leached_n_kgha[Treatment == "Sorghum + WCC"] |> na.omit()
      if (length(s) >= 3 && length(w) >= 3)
        tryCatch(wilcox.test(s, w)$p.value, error = \(e) NA_real_)
      else NA_real_
    },
    .by = c(year, date)
  ) |>
  mutate(significant = !is.na(p_value) & p_value < 0.05)

leaching_stars <- daily_leaching_summary |>
  summarise(y_top = max(mean_leached + se_leached, na.rm = TRUE), .by = c(year, date)) |>
  left_join(daily_leaching_sig, by = c("year", "date")) |>
  filter(significant) |>
  mutate(star_y = y_top * 1.1, star_label = "*")
```

```{r daily-diff}
# ── Daily difference (WCC - Sorghum) and one-sample t-test ────────────────
daily_leach_diff <- daily_leaching_summary |>
  mutate(trt_key = if_else(Treatment == "Sorghum + WCC", "WCC", "Sorghum")) |>
  select(year, date, trt_key, mean_leached) |>
  pivot_wider(names_from = trt_key, values_from = mean_leached) |>
  mutate(diff_wcc_minus_sorghum = WCC - Sorghum)

one_sample_vs_zero <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, mu = 0)$p.value, error = \(e) NA_real_)
}

leach_diff_stats <- daily_leach_diff |>
  summarise(
    n          = sum(!is.na(diff_wcc_minus_sorghum)),
    mean_diff  = mean(diff_wcc_minus_sorghum, na.rm = TRUE),
    p_value    = one_sample_vs_zero(diff_wcc_minus_sorghum),
    significant = !is.na(p_value) & p_value < 0.05,
    letter     = if_else(significant, "a", NA_character_),
    .by = year
  )
```

```{r annual-anova}
# ── Annual cumulative leaching + one-way ANOVA ────────────────────────────
annual_leached <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    plot = as.character(plot)
  ) |>
  summarise(
    annual_leached_n = (max(cumulative_n_loss_mg, na.rm = TRUE) / 1e6) / plot_area_ha,
    .by = c(year, plot, Treatment)
  ) |>
  filter(is.finite(annual_leached_n), annual_leached_n >= 0)

anova_leached   <- aov(annual_leached_n ~ Treatment, data = annual_leached)
anova_leached_p <- summary(anova_leached)[[1]]["Treatment", "Pr(>F)"]

leached_letters <- tryCatch({
  emmeans(anova_leached, ~ Treatment) |>
    cld(adjust = "tukey", Letters = letters) |>
    as_tibble() |>
    transmute(Treatment = as.character(Treatment),
              letter    = gsub("\\s+", "", .group))
}, error = \(e) {
  annual_leached |> distinct(Treatment) |> mutate(letter = "a")
})
```

# Figure 3

```{r figure3}
#| fig-width: 12
#| fig-height: 8
#| fig-dpi: 300

# ── Precipitation trimmed to leaching windows ─────────────────────────────
leach_date_ranges <- daily_leaching_summary |>
  summarise(min_date = min(date), max_date = max(date), .by = year)

daily_precip_leaching <- rain |>
  select(date, year, precipmm) |>
  filter(year %in% c(2023, 2024)) |>
  left_join(leach_date_ranges, by = "year") |>
  filter(date >= min_date, date <= max_date)

# ── Scaling constants ─────────────────────────────────────────────────────
max_leached_daily <- max(daily_leaching_summary$mean_leached +
                         daily_leaching_summary$se_leached, na.rm = TRUE)
max_precip_leach  <- max(daily_precip_leaching$precipmm, na.rm = TRUE)
leach_precip_scale <- (max_leached_daily * 0.4) / max_precip_leach
leach_plot_top     <- max_leached_daily * 1.12
max_cum_leached    <- max(cum_leaching_summary$cum_leached_mean +
                          cum_leaching_summary$cum_leached_se, na.rm = TRUE)
## Panel Builder

# ── Panel builder: daily leaching ─────────────────────────────────────────
create_leaching_daily_panel <- function(year_val, is_right, tag_label) {
  d_sub    <- filter(daily_leaching_summary, year == year_val)
  p_sub    <- filter(daily_precip_leaching, year == year_val)
  s_sub    <- filter(leaching_stars, year == year_val)
  fert_sub <- filter(fert_events, year == year_val)
  x_lims   <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot() +
    # Precipitation (inverted from top)
    geom_rect(
      data = p_sub,
      aes(xmin = date - 0.5, xmax = date + 0.5,
          ymin = leach_plot_top - precipmm * leach_precip_scale,
          ymax = leach_plot_top),
      fill = "#377EB8", alpha = 0.7
    ) +
    geom_point(data = d_sub, aes(x = date, y = mean_leached, color = Treatment),
               size = 1.4) +
    geom_errorbar(
      data = d_sub,
      aes(x = date,
          ymin = pmax(mean_leached - se_leached, 0),
          ymax = mean_leached + se_leached,
          color = Treatment),
      width = 0.2, alpha = 0.5
    ) +
    geom_vline(data = fert_sub, aes(xintercept = fertilizer_date),
               linetype = "dashed", color = "black", alpha = 0.7) +
    geom_text(data = s_sub, aes(x = date, y = star_y, label = star_label),
              size = 4.8) +
    scale_color_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right) expression("Daily Leached N (kg N ha"^-1 * " d"^-1 * ")") else "",
      limits = c(0, leach_plot_top),
      sec.axis = sec_axis(
        ~ (leach_plot_top - .) / leach_precip_scale,
        name = if (is_right) "Precipitation (mm)" else ""
      )
    ) +
    labs(x = "", tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(
      axis.text.x      = element_blank(),
      axis.ticks.x     = element_blank(),
      axis.line.x      = element_blank(),
      legend.position  = "none",
      plot.tag          = element_text(face = "bold", size = 11),
      plot.tag.position = "topleft"
    )

  if (is_right) {
    p <- p + theme(
      axis.text.y        = element_blank(),
      axis.ticks.y       = element_blank(),
      axis.line.y        = element_blank(),
      axis.title.y       = element_blank(),
      axis.text.y.right  = element_text(color = "#377EB8"),
      axis.ticks.y.right = element_line(),
      axis.title.y.right = element_text(color = "#377EB8")
    )
  } else {
    p <- p + theme(
      axis.text.y.right  = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.title.y.right = element_blank()
    )
  }
  p
}

# ── Panel builder: cumulative leaching ────────────────────────────────────
create_leaching_cum_panel <- function(year_val, is_right, tag_label) {
  c_sub    <- filter(cum_leaching_summary, year == year_val)
  fert_sub <- filter(fert_events, year == year_val)
  x_lims   <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot(c_sub, aes(x = date, y = cum_leached_mean,
                          color = Treatment, fill = Treatment)) +
    geom_ribbon(aes(ymin = pmax(cum_leached_mean - cum_leached_se, 0),
                    ymax = cum_leached_mean + cum_leached_se),
                alpha = 0.15, linewidth = 0) +
    geom_line(linewidth = 1) +
    geom_vline(data = fert_sub, aes(xintercept = fertilizer_date),
               linetype = "dashed", color = "black", alpha = 0.7) +
    scale_color_manual(values = treatment_colors) +
    scale_fill_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right) expression("Cumulative Leached N (kg N ha"^-1 * ")") else "",
      limits = c(0, max_cum_leached)
    ) +
    labs(x = "Month", tag = tag_label) +
    theme_publication(base_size = 10) +
    theme(legend.position = "none",
          plot.tag = element_text(face = "bold", size = 11),
          plot.tag.position = "topleft")

  if (is_right) {
    p <- p + theme(
      axis.text.y  = element_blank(),
      axis.ticks.y = element_blank(),
      axis.line.y  = element_blank(),
      axis.title.y = element_blank()
    )
  }
  p
}

# ── Panel (c): daily difference boxplot ────────────────────────────────────
diff_span <- diff(range(daily_leach_diff$diff_wcc_minus_sorghum, na.rm = TRUE))
if (!is.finite(diff_span) || diff_span == 0) diff_span <- 0.05

leach_diff_anno <- daily_leach_diff |>
  summarise(y = max(diff_wcc_minus_sorghum, na.rm = TRUE), .by = year) |>
  left_join(leach_diff_stats, by = "year") |>
  filter(significant) |>
  mutate(y = y + diff_span * 0.08)

p3c <- ggplot(daily_leach_diff, aes(x = factor(year), y = diff_wcc_minus_sorghum)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
  geom_jitter(aes(color = factor(year)), width = 0.12, alpha = 0.7,
              size = 1.4, show.legend = FALSE) +
  geom_text(data = leach_diff_anno,
            aes(x = factor(year), y = y, label = letter),
            inherit.aes = FALSE, fontface = "bold", size = 3.8) +
  scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
  labs(
    x = "Year",
    y = expression("Daily Leached N Difference (kg N ha"^-1 * " d"^-1 * ", WCC - Sorghum)"),
    tag = "(c)"
  ) +
  theme_publication(base_size = 10) +
  theme(plot.tag = element_text(face = "bold", size = 11),
        plot.tag.position = "topleft")

# ── Panel (f): annual cumulative by treatment ──────────────────────────────
letter_offset <- diff(range(annual_leached$annual_leached_n, na.rm = TRUE))
if (!is.finite(letter_offset) || letter_offset == 0) letter_offset <- 0.05

annual_letter_pos <- annual_leached |>
  summarise(y = max(annual_leached_n, na.rm = TRUE), .by = Treatment) |>
  left_join(leached_letters, by = "Treatment") |>
  mutate(y = y + letter_offset * 0.08)

p3f <- ggplot(annual_leached, aes(x = Treatment, y = annual_leached_n, fill = Treatment)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.5, color = "black") +
  geom_jitter(aes(shape = factor(year)), width = 0.12, size = 1.8,
              alpha = 0.8, show.legend = FALSE) +
  geom_text(data = annual_letter_pos,
            aes(x = Treatment, y = y, label = letter),
            inherit.aes = FALSE, fontface = "bold", size = 3.8) +
  scale_fill_manual(values = treatment_colors) +
  labs(
    x = "",
    y = expression("Annual Leached N (kg N ha"^-1 * ")"),
    subtitle = paste0("One-way ANOVA p = ",
                      ifelse(anova_leached_p < 0.001, "< 0.001",
                             sprintf("%.3f", anova_leached_p))),
    tag = "(f)"
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "none",
        plot.tag = element_text(face = "bold", size = 11),
        plot.tag.position = "topleft")

# ── Build panels ──────────────────────────────────────────────────────────
p3a <- create_leaching_daily_panel(2023, is_right = FALSE, "(a)")
p3b <- create_leaching_daily_panel(2024, is_right = TRUE,  "(b)")
p3d <- create_leaching_cum_panel(2023, is_right = FALSE, "(d)")
p3e <- create_leaching_cum_panel(2024, is_right = TRUE,  "(e)")

# ── Assemble ──────────────────────────────────────────────────────────────
figure3 <- (p3a | p3b | p3c) / (p3d | p3e | p3f) +
  plot_layout(guides = "collect", widths = c(1, 1, 0.95)) &
  theme(legend.position = "top", legend.justification = "center")

figure3

ggsave(filename = "figure_3_leaching.png", plot = figure3, width = 12, height = 8, dpi = 300)
```

```{r print-stats}
#| echo: false
cat("Daily difference stats:\n")
leach_diff_stats
cat("\nANOVA p-value:", anova_leached_p, "\n")
```

# Alternative Statistical Approaches

The sections below explore alternative analytical frameworks for the leaching
data. Each approach creates its own variables and leaves the existing figure
code untouched. Note that the Google Doc recommends renaming the y-axis label
to **"Leached N"** (nitrate + ammonium) in all new panels.

## Approach B: Mixed Model (Sorghum Pair)

This approach fits a linear mixed-effects model to annual cumulative leached N
for the two-treatment subset (Sorghum vs Sorghum + WCC), treating `plot` as a
random intercept and crossing Treatment with year.

```{r approach-b-mixed-model}
#| message: false
#| warning: false

library(lme4)

# Ensure year is a factor for model fitting (avoids column-name issues in emmeans)
annual_leached <- annual_leached |>
  mutate(year_fct = factor(year))

# Fit mixed model on the existing annual_leached data
mod_leach <- lmer(
  annual_leached_n ~ Treatment * year_fct + (1 | plot),
  data = annual_leached
)

cat("── Approach B: Mixed Model Summary ──────────────────────────\n")
summary(mod_leach)

# Pairwise contrasts by year
emm_leach_b <- emmeans(mod_leach, pairwise ~ Treatment | year_fct)
cat("\n── Pairwise Contrasts by Year ──────────────────────────────\n")
emm_leach_b

# Extract means and contrasts for plotting
emm_leach_b_df <- as.data.frame(emm_leach_b$emmeans)

contrast_leach_b <- as.data.frame(emm_leach_b$contrasts) |>
  mutate(
    p_label = ifelse(p.value < 0.001, "p < 0.001", sprintf("p = %.3f", p.value)),
    year = as.numeric(as.character(year_fct))
  )

# Bar chart with means ± SE and p-value annotations
bar_max_b <- max(emm_leach_b_df$emmean + emm_leach_b_df$SE, na.rm = TRUE)

p_approach_b <- ggplot(emm_leach_b_df,
                       aes(x = year_fct, y = emmean, fill = Treatment)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6,
           color = "black", alpha = 0.8) +
  geom_errorbar(
    aes(ymin = pmax(emmean - SE, 0), ymax = emmean + SE),
    position = position_dodge(width = 0.7), width = 0.2
  ) +
  geom_text(
    data = contrast_leach_b,
    aes(x = factor(year), y = bar_max_b * 1.08, label = p_label),
    inherit.aes = FALSE, size = 3.5, fontface = "italic"
  ) +
  scale_fill_manual(values = treatment_colors) +
  labs(
    x = "Year",
    y = expression("Leached N (kg N ha"^-1 * ")"),
    title = "Approach B: Mixed Model — Treatment Means ± SE by Year"
  ) +
  theme_publication(base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

p_approach_b
```

## Approach C: All-Treatment ANOVA with CLD

This section includes all four treatments (Sorghum, Sorghum + Rye, Corn, Soy)
and applies compact letter display (CLD) via Tukey HSD for pairwise
comparisons.

```{r approach-c-all-treatment}
#| message: false
#| warning: false

# Recompute annual cumulative leached N for ALL treatments
annual_leached_all <- leaching_data |>
  mutate(plot = as.character(plot)) |>
  summarise(
    annual_leached_n = (max(cumulative_n_loss_mg, na.rm = TRUE) / 1e6) / plot_area_ha,
    .by = c(year, plot, treatment)
  ) |>
  filter(is.finite(annual_leached_n), annual_leached_n >= 0) |>
  mutate(year_fct = factor(year))

# Attempt mixed model; fall back to aov if convergence fails
mod_leach_all <- tryCatch(
  {
    m <- lmer(
      annual_leached_n ~ treatment * year_fct + (1 | plot),
      data = annual_leached_all
    )
    cat("Mixed model (lmer) converged successfully.\n")
    m
  },
  error = function(e) {
    cat("lmer failed to converge — falling back to aov().\n")
    cat("Error:", conditionMessage(e), "\n")
    NULL
  }
)

if (is.null(mod_leach_all)) {
  mod_leach_all <- aov(
    annual_leached_n ~ treatment * year_fct,
    data = annual_leached_all
  )
  cat("aov() model fitted.\n")
}

cat("\n── Approach C: Model Summary ──────────────────────────────\n")
summary(mod_leach_all)

# CLD via emmeans + Tukey HSD
emm_leach_c <- emmeans(mod_leach_all, ~ treatment | year_fct)
cld_leach_c <- cld(emm_leach_c, adjust = "tukey", Letters = letters) |>
  as.data.frame() |>
  mutate(
    letter = gsub("\\s+", "", .group),
    year_num = as.numeric(as.character(year_fct))
  )

cat("\n── CLD Groups by Year ──────────────────────────────────────\n")
print(cld_leach_c |> select(treatment, year_fct, emmean, SE, letter))

# All-treatment color palette
treatment_colors_all <- c(
  "Sorghum"       = "#D8D97AFF",
  "Sorghum + Rye" = "#95C36EFF",
  "Corn"          = "#74C8C3FF",
  "Soy"           = "#0A2E57FF"
)

# Grouped bar chart with CLD letters
bar_max_c <- max(cld_leach_c$emmean + cld_leach_c$SE, na.rm = TRUE)

p_approach_c <- ggplot(cld_leach_c,
                       aes(x = factor(year_num), y = emmean, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65,
           color = "black", alpha = 0.8) +
  geom_errorbar(
    aes(ymin = pmax(emmean - SE, 0), ymax = emmean + SE),
    position = position_dodge(width = 0.75), width = 0.2
  ) +
  geom_text(
    aes(y = emmean + SE + bar_max_c * 0.03, label = letter),
    position = position_dodge(width = 0.75),
    size = 3.5, fontface = "bold", vjust = 0
  ) +
  scale_fill_manual(values = treatment_colors_all) +
  labs(
    x = "Year",
    y = expression("Leached N (kg N ha"^-1 * ")"),
    fill = "Treatment",
    title = "Approach C: All-Treatment ANOVA — CLD Groups by Year"
  ) +
  theme_publication(base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

p_approach_c
```

## Approach D: Log-Transformed Mixed Model

Leaching data are typically right-skewed. This approach fits the same Sorghum-
pair mixed model from Approach B on the log-transformed response and compares
residual diagnostics.

```{r approach-d-log-model}
#| message: false
#| warning: false

# Log-transformed mixed model (small constant avoids log(0))
mod_leach_log <- lmer(
  log(annual_leached_n + 0.01) ~ Treatment * year_fct + (1 | plot),
  data = annual_leached
)

cat("── Approach D: Log-Transformed Model Summary ────────────────\n")
summary(mod_leach_log)

# Back-transformed emmeans
emm_leach_d <- emmeans(mod_leach_log, pairwise ~ Treatment | year_fct, type = "response")
cat("\n── Back-Transformed Means ──────────────────────────────────\n")
emm_leach_d
```

```{r approach-d-diagnostics}
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 10

# Residual diagnostics comparison
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Untransformed model (Approach B) residuals
plot(fitted(mod_leach), resid(mod_leach),
     xlab = "Fitted", ylab = "Residuals",
     main = "Approach B: Residuals vs Fitted",
     pch = 16, col = alpha("steelblue", 0.7))
abline(h = 0, lty = 2, col = "red")

qqnorm(resid(mod_leach), main = "Approach B: Q-Q Plot", pch = 16,
       col = alpha("steelblue", 0.7))
qqline(resid(mod_leach), col = "red", lwd = 1.5)

# Log-transformed model (Approach D) residuals
plot(fitted(mod_leach_log), resid(mod_leach_log),
     xlab = "Fitted", ylab = "Residuals",
     main = "Approach D: Residuals vs Fitted (log)",
     pch = 16, col = alpha("darkorange", 0.7))
abline(h = 0, lty = 2, col = "red")

qqnorm(resid(mod_leach_log), main = "Approach D: Q-Q Plot (log)", pch = 16,
       col = alpha("darkorange", 0.7))
qqline(resid(mod_leach_log), col = "red", lwd = 1.5)

par(mfrow = c(1, 1))
```

## Approach E: BH FDR-Corrected Per-Date Wilcoxon

The existing per-date Wilcoxon tests (used for asterisks in panels a--b) do
not account for the large number of comparisons across dates. This approach
applies Benjamini--Hochberg FDR correction within each year.

```{r approach-e-fdr}
#| message: false
#| warning: false

# Apply BH correction within each year
fdr_leaching_sig <- daily_leaching_sig |>
  group_by(year) |>
  mutate(p_adjusted = p.adjust(p_value, method = "BH")) |>
  ungroup() |>
  mutate(significant_fdr = !is.na(p_adjusted) & p_adjusted < 0.05)

# Comparison: uncorrected vs FDR-corrected significant dates
fdr_comparison <- fdr_leaching_sig |>
  summarise(
    n_dates_tested   = sum(!is.na(p_value)),
    n_sig_uncorrected = sum(significant, na.rm = TRUE),
    n_sig_fdr        = sum(significant_fdr, na.rm = TRUE),
    pct_lost         = ifelse(
      n_sig_uncorrected > 0,
      sprintf("%.1f%%", 100 * (1 - n_sig_fdr / n_sig_uncorrected)),
      "N/A"
    ),
    .by = year
  )

cat("── Approach E: Uncorrected vs BH FDR-Corrected Significance ──\n\n")
knitr::kable(
  fdr_comparison,
  col.names = c("Year", "Dates Tested", "Sig. (Uncorrected)",
                "Sig. (BH FDR)", "% Lost"),
  align = "ccccc"
)
```

## Statistical Comparison

Summary table comparing all approaches on the leaching data.

```{r statistical-comparison}
#| message: false
#| warning: false

# ── Gather metrics from each approach ────────────────────────────────────

# Approach A: existing one-way ANOVA
approach_a_p <- ifelse(anova_leached_p < 0.001, "< 0.001",
                       sprintf("%.3f", anova_leached_p))
approach_a_nsig <- sum(daily_leaching_sig$significant, na.rm = TRUE)

# Approach B: mixed model Treatment main effect
# lmerTest provides "Pr(>F)"; without it the column may be "Pr(>Chisq)" or absent
approach_b_anova <- tryCatch({
  anova_b <- anova(mod_leach)
  ab <- anova_b["Treatment", ]
  p_col <- intersect(c("Pr(>F)", "Pr(>Chisq)"), names(ab))
  p_trt <- if (length(p_col) > 0) ab[[p_col[1]]] else NA_real_
  if (is.null(p_trt) || length(p_trt) == 0 || is.na(p_trt)) "N/A"
  else if (p_trt < 0.001) "< 0.001"
  else sprintf("%.3f", p_trt)
}, error = function(e) "N/A")

approach_b_pairwise <- contrast_leach_b |>
  mutate(entry = paste0(year, ": ", p_label)) |>
  pull(entry) |>
  paste(collapse = "; ")

# Approach C: CLD groups
approach_c_cld <- cld_leach_c |>
  mutate(entry = paste0(treatment, " (", year_fct, "): ", letter)) |>
  pull(entry) |>
  paste(collapse = "; ")

# Approach D: log-transformed model p-values
approach_d_anova <- tryCatch({
  anova_d <- anova(mod_leach_log)
  ad <- anova_d["Treatment", ]
  p_col_d <- intersect(c("Pr(>F)", "Pr(>Chisq)"), names(ad))
  p_trt_d <- if (length(p_col_d) > 0) ad[[p_col_d[1]]] else NA_real_
  if (is.null(p_trt_d) || length(p_trt_d) == 0 || is.na(p_trt_d)) "N/A"
  else if (p_trt_d < 0.001) "< 0.001"
  else sprintf("%.3f", p_trt_d)
}, error = function(e) "N/A")

# Approach E: FDR-corrected counts
approach_e_nsig <- sum(fdr_leaching_sig$significant_fdr, na.rm = TRUE)

# ── Build comparison table ───────────────────────────────────────────────
comparison_table <- tibble(
  Approach = c(
    "A: One-way ANOVA (current)",
    "B: Mixed Model (Sorghum pair)",
    "C: All-Treatment ANOVA + CLD",
    "D: Log-Transformed Mixed Model",
    "E: BH FDR-Corrected Wilcoxon"
  ),
  `Key Statistic` = c(
    paste0("ANOVA p = ", approach_a_p),
    paste0("Treatment p = ", approach_b_anova),
    "CLD groups (see below)",
    paste0("Treatment p = ", approach_d_anova),
    paste0(approach_e_nsig, " significant dates (FDR)")
  ),
  Details = c(
    paste0(approach_a_nsig, " significant Wilcoxon dates (uncorrected)"),
    approach_b_pairwise,
    approach_c_cld,
    "Log(y + 0.01); back-transformed means shown above",
    paste0("vs ", approach_a_nsig, " uncorrected; BH correction within year")
  )
)

cat("── Statistical Comparison Across Approaches ────────────────\n\n")
knitr::kable(comparison_table, align = "lll")
```
