---
title: "Figure S3: Row vs. Interrow N₂O Flux"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    fig-width: 10
    fig-height: 8
execute:
  warning: false
  message: false
  echo: true
---

# Caption

**Figure S3.** Daily N₂O flux by measurement position (Row vs. Interrow) for
Sorghum and Sorghum + WCC treatments in **(a)** 2023 and **(b)** 2024.
Boxplots show the median and interquartile range; individual observations are
overlaid as jittered points. The x-axis is log₁₀-scaled. Panels **(c)** and
**(d)** show the distribution of within-date, within-plot differences
(Interrow − Row) for each year. The x-axis uses a pseudo-log (signed
symmetric log) scale: values near zero are displayed on a linear scale,
while the tails are compressed logarithmically to reveal the shape of the
distribution without hiding the extreme observations. The red dashed line
marks zero (no difference). Most differences cluster tightly near zero;
large positive differences indicate measurement dates when Interrow flux
exceeded Row flux substantially, and vice versa.

# Setup

```{r setup}
library(tidyverse)
library(lubridate)
library(patchwork)
library(scales)

# ── Publication theme (matches figure_1_weather.qmd) ──────────────────────────
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      panel.background   = element_rect(fill = "#ffffffff", color = NA),
      plot.background    = element_rect(fill = "#ffffffff", color = NA),
      strip.background   = element_rect(fill = alpha("#A5CAD2", 0.2)),
      panel.grid.major   = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor   = element_blank(),
      panel.spacing      = unit(2, "lines"),
      panel.border       = element_rect(color = "black", fill = NA, linewidth = 1),
      text               = element_text(color = "#000000ff"),
      axis.text          = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title         = element_text(size = base_size),
      axis.title.x       = element_text(margin = margin(t = 8)),
      axis.title.y       = element_text(margin = margin(r = 8)),
      axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line          = element_line(color = "black", linewidth = 0.5),
      axis.ticks         = element_line(color = "black", linewidth = 0.5),
      legend.position    = "top",
      legend.justification = "center",
      legend.direction   = "horizontal",
      legend.box         = "horizontal",
      legend.margin      = margin(b = 1, unit = "pt"),
      legend.box.margin  = margin(b = 1, unit = "pt"),
      legend.text        = element_text(size = base_size * 0.9),
      legend.title       = element_text(size = base_size),
      legend.key.width   = unit(0.8, "cm"),
      legend.key.height  = unit(0.3, "cm")
    )
}

# ── Colors ─────────────────────────────────────────────────────────────────────
year_colors <- c("2023" = "#E41A1C", "2024" = "#377EB8")

treatment_colors <- c(
  "Sorghum"       = "#D8D97AFF",
  "Sorghum + WCC" = "#95C36EFF"
)

position_colors <- c("Row" = "grey80", "Interrow" = "grey40")
```

# Data

```{r data}
flux_data <- read_csv("data/cleaned_flux_data.csv") |>
  mutate(date = as.Date(date))

# Sorghum vs Sorghum + WCC only; relabel and set factor order
row_interrow_comparison <- flux_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment      = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    Treatment      = factor(Treatment, levels = c("Sorghum", "Sorghum + WCC")),
    rowvs_interrow = factor(rowvs_interrow, levels = c("Row", "Interrow")),
    year           = year(date)
  ) |>
  filter(year %in% c(2023, 2024)) |>
  select(year, date, Treatment, plot, rowvs_interrow, gnha_day_no_negative)

# Paired differences: Interrow − Row per plot × date
diff_analysis <- row_interrow_comparison |>
  pivot_wider(
    names_from  = rowvs_interrow,
    values_from = gnha_day_no_negative,
    names_prefix = "Flux_"
  ) |>
  mutate(Diff = Flux_Interrow - Flux_Row) |>
  filter(!is.na(Diff)) |>
  mutate(year = year(date))
```

# Summary statistics

```{r summary}
cat("=== Mean flux by year / treatment / position ===\n")
row_interrow_comparison |>
  group_by(year, Treatment, rowvs_interrow) |>
  summarise(
    n         = n(),
    mean_flux = mean(gnha_day_no_negative, na.rm = TRUE),
    median_flux = median(gnha_day_no_negative, na.rm = TRUE),
    sd_flux   = sd(gnha_day_no_negative, na.rm = TRUE),
    .groups   = "drop"
  ) |> print()

cat("\n=== One-sample t-test: mean difference ≠ 0 (Interrow − Row) ===\n")
diff_analysis |>
  group_by(year) |>
  summarise(
    n        = n(),
    mean_diff = mean(Diff),
    t_stat   = t.test(Diff)$statistic,
    p_value  = t.test(Diff)$p.value,
    .groups  = "drop"
  ) |> print()
```

# Figure

```{r figure}
# ── Panels (a) & (b): absolute flux, log x-axis ────────────────────────────────
make_flux_panel <- function(yr, tag_label) {
  ggplot(
    filter(row_interrow_comparison, year == yr),
    aes(x = gnha_day_no_negative, color = rowvs_interrow)
  ) +
    geom_boxplot(aes(y = Treatment), outlier.shape = NA, alpha = 0.7) +
    geom_jitter(
      aes(y = Treatment),
      position = position_jitterdodge(jitter.width = 0.2),
      alpha = 0.4, size = 0.7
    ) +
    scale_x_log10(labels = label_log()) +
    annotation_logticks(sides = "b") +
    scale_color_manual(values = position_colors, name = "Position") +
    labs(
      y     = "",
      x     = expression("N"[2]*"O Flux (g N ha"^{-1}*" day"^{-1}*")"),
      tag   = tag_label,
      title = as.character(yr)
    ) +
    theme_publication(base_size = 10) +
    theme(
      legend.position   = if (tag_label == "(a)") "top" else "none",
      plot.tag          = element_text(face = "bold", size = 11),
      plot.tag.position = "topleft",
      plot.title        = element_text(
        hjust = 0.5, face = "bold", size = 12,
        color = year_colors[as.character(yr)]
      )
    )
}

# ── Panels (c) & (d): difference distribution, pseudo-log x-axis ───────────────
# pseudo_log_trans: linear near 0, logarithmic in the tails — handles negatives.
# sigma controls the width of the linear region (units = g N ha-1 day-1).
make_diff_panel <- function(yr, tag_label) {
  df <- filter(diff_analysis, year == yr)

  # Compute % of observations within ±10 for annotation
  pct_near_zero <- round(mean(abs(df$Diff) <= 10) * 100, 0)
  n_extreme <- sum(abs(df$Diff) > 10)

  ggplot(df, aes(x = Diff, fill = Treatment)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins = 60, alpha = 0.75, position = "identity",
      color = "white", linewidth = 0.2
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red",
               linewidth = 0.7) +
    # Pseudo-log scale: sigma = 2 keeps small differences visible
    scale_x_continuous(
      trans  = pseudo_log_trans(sigma = 2, base = 10),
      breaks = c(-100, -10, -1, 0, 1, 10, 100),
      labels = label_comma()
    ) +
    scale_fill_manual(values = treatment_colors, name = "Treatment") +
    annotate(
      "text", x = Inf, y = Inf,
      label = paste0(pct_near_zero, "% within ±10\n",
                     n_extreme, " obs. outside ±10"),
      hjust = 1.05, vjust = 1.3, size = 2.8, color = "grey30"
    ) +
    labs(
      x     = expression("Interrow \u2212 Row (g N ha"^{-1}*" day"^{-1}*")"),
      y     = "Density",
      tag   = tag_label,
      title = as.character(yr)
    ) +
    theme_publication(base_size = 10) +
    theme(
      legend.position   = if (tag_label == "(c)") "top" else "none",
      plot.tag          = element_text(face = "bold", size = 11),
      plot.tag.position = "topleft",
      plot.title        = element_text(
        hjust = 0.5, face = "bold", size = 12,
        color = year_colors[as.character(yr)]
      )
    )
}

p_a <- make_flux_panel(2023, "(a)")
p_b <- make_flux_panel(2024, "(b)")
p_c <- make_diff_panel(2023, "(c)")
p_d <- make_diff_panel(2024, "(d)")

fig_s3 <- (p_a | p_b) / (p_c | p_d)
fig_s3
```

```{r save}
ggsave(
  "SubmissionFiguresTables/FigureS3_row_vs_interrow.tiff",
  plot        = fig_s3,
  width       = 174, height = 200, units = "mm", dpi = 600, bg = "white",
  compression = "lzw"
)
```
