---
title: "Figure 2: N₂O Flux with Plot-Level Cumulative Uncertainty"
format: 
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: false
---

# Setup and Data Loading

```{r setup}
#| label: setup
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(lubridate)
library(patchwork)
library(readxl)

# Re-establish dplyr preferences after loading packages
if (requireNamespace("conflicted", quietly = TRUE)) {
  library(conflicted)
  conflicts_prefer(dplyr::select)
  conflicts_prefer(dplyr::filter)
  conflicts_prefer(dplyr::summarise)
}

# Custom theme for consistency (from Final_Figures_Tables.qmd)
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.background = element_rect(fill = "#ffffffff", color = NA),
      plot.background = element_rect(fill = "#ffffffff", color = NA),
      panel.grid.major = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor = element_blank(),
      text = element_text(color = "#000000ff"),
      axis.text = element_text(color = "#000000ff"),
      strip.background = element_rect(fill = alpha("#A5CAD2", 0.2)),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.spacing = unit(2, "lines")
    )
}

# Treatment colors (from Final_Figures_Tables.qmd)
treatment_colors <- c(
  "Sorghum" = "#D8D97AFF",
  "Sorghum + Rye" = "#95C36EFF", 
  "Corn" = "#74C8C3FF",
  "Soy" = "#0A2E57FF"
)

scale_color_treatments <- function() scale_color_manual(values = treatment_colors)
scale_fill_treatments <- function() scale_fill_manual(values = treatment_colors)
```

# Data Import

```{r data-import}
#| label: data-import
#| message: false

# Import flux data 
flux_data <- read_csv("data/cleaned_flux_data.csv") %>%
  mutate(date = as.Date(date)) 

# Import historical rain data
rain <- read_xlsx("data/rain/historical_rain.xlsx")

rain <- rain %>%
  select(day, doy, precipmm) %>%
  mutate(
    date = as.Date(day),
    year = year(date)
  ) %>%
  group_by(year) %>%
  arrange(date) %>%
  mutate(cumulative_precip = cumsum(precipmm))

rain <- rain %>%
  mutate(doy_date = as.Date(doy - 1, origin = "2000-01-01"))
```

# Plot-Level Cumulative Flux Calculation

This implements a more statistically appropriate uncertainty method:
- Work at the plot level (Treatment × year × plot)
- Interpolate daily fluxes between measurement dates
- Compute cumulative flux over time per plot
- Summarize across plots to get mean, SE, and 95% CI

```{r plot-level-cumulative}
#| label: plot-level-cumulative

# Filter to target treatments and years
flux_filtered <- flux_data %>%
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) %>%
  filter(year %in% c(2023, 2024))

# Calculate daily flux per plot (averaging row and interrow within each plot-date)
daily_flux_plot <- flux_filtered %>%
  group_by(plot, treatment, year, date) %>%
  summarise(
    flux = mean(gnha_day_no_negative, na.rm = TRUE) / 1000,  # Convert g to kg
    .groups = "drop"
  ) %>%
  arrange(treatment, year, plot, date)

# Get date ranges per treatment-year for interpolation bounds
date_ranges <- daily_flux_plot %>%
  group_by(treatment, year) %>%
  summarise(
    min_date = min(date, na.rm = TRUE),
    max_date = max(date, na.rm = TRUE),
    .groups = "drop"
  )

# Implement plot-level cumulative flux with interpolation
cum_flux_plot_level <- daily_flux_plot %>%
  # Create all combinations of treatment-year-plot
  group_by(treatment, year, plot) %>%
  # Get the date range for this treatment-year
  left_join(date_ranges, by = c("treatment", "year")) %>%
  group_modify(~ {
    df <- .x
    if (nrow(df) < 2) return(tibble())
    
    # Create full date sequence within measurement range
    all_dates <- seq(first(df$min_date), first(df$max_date), by = "day")
    
    # Create tibble with all dates
    full_dates <- tibble(date = all_dates)
    
    # Join with observed data
    df_full <- full_dates %>%
      left_join(df %>% select(date, flux), by = "date")
    
    # Interpolate flux values
    valid_data <- df %>% filter(!is.na(flux))
    if (nrow(valid_data) >= 2) {
      df_full <- df_full %>%
        mutate(
          flux_interp = approx(
            x = valid_data$date,
            y = valid_data$flux,
            xout = date,
            rule = 1,
            ties = "ordered"
          )$y
        )
    } else {
      df_full <- df_full %>%
        mutate(flux_interp = NA_real_)
    }
    
    # Calculate cumulative flux
    df_full %>%
      filter(!is.na(flux_interp)) %>%
      mutate(cum_flux = cumsum(flux_interp))
  }) %>%
  ungroup()

# Summarize across plots to get mean and CI
cum_flux_summary <- cum_flux_plot_level %>%
  group_by(treatment, year, date) %>%
  summarise(
    cum_flux_mean = mean(cum_flux, na.rm = TRUE),
    cum_flux_sd = sd(cum_flux, na.rm = TRUE),
    n_plots = n(),
    cum_flux_se = cum_flux_sd / sqrt(n_plots),
    t_crit = qt(0.975, df = pmax(n_plots - 1, 1)),
    cum_flux_ci = t_crit * cum_flux_se,
    .groups = "drop"
  ) %>%
  # Handle cases where CI might be NA
  mutate(
    cum_flux_ci = replace_na(cum_flux_ci, 0)
  )

# Create a version with observed points marked
cum_flux_observed <- cum_flux_plot_level %>%
  filter(!is.na(flux)) %>%
  group_by(treatment, year, date) %>%
  summarise(
    cum_flux_mean = mean(cum_flux, na.rm = TRUE),
    .groups = "drop"
  )

cat("✓ Plot-level cumulative flux calculation complete\n")
cat("Summary data:", nrow(cum_flux_summary), "rows\n")
cat("Treatments:", paste(unique(cum_flux_summary$treatment), collapse = ", "), "\n")
cat("Years:", paste(unique(cum_flux_summary$year), collapse = ", "), "\n")
```

# Build Daily Flux Summary

```{r daily-flux-summary}
#| label: daily-flux-summary

# Build daily treatment means and SE for N2O flux (for daily plot)
daily_flux <- flux_filtered %>%
  mutate(Treatment = treatment) %>%
  group_by(Treatment, year, date) %>%
  summarise(
    mean_flux = mean(gnha_day_no_negative, na.rm = TRUE) / 1000,  # Convert g to kg
    se_flux = sd(gnha_day_no_negative, na.rm = TRUE) / sqrt(sum(!is.na(gnha_day_no_negative))) / 1000,
    .groups = "drop"
  )

# Get flux measurement date ranges for each year
flux_date_ranges <- daily_flux %>%
  group_by(year) %>%
  summarise(
    min_date = min(date, na.rm = TRUE),
    max_date = max(date, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate daily significance tests between treatments
daily_significance <- flux_filtered %>%
  group_by(year, date) %>%
  summarise(
    p_value = {
      sorghum_flux <- gnha_day_no_negative[treatment == "Sorghum"]
      sorghum_rye_flux <- gnha_day_no_negative[treatment == "Sorghum + Rye"]
      sorghum_flux <- sorghum_flux[!is.na(sorghum_flux)]
      sorghum_rye_flux <- sorghum_rye_flux[!is.na(sorghum_rye_flux)]
      if (length(sorghum_flux) >= 2 && length(sorghum_rye_flux) >= 2) {
        tryCatch({
          t_result <- t.test(sorghum_flux, sorghum_rye_flux)
          t_result$p.value
        }, error = function(e) NA_real_)
      } else {
        NA_real_
      }
    },
    .groups = "drop"
  ) %>%
  mutate(
    significant = p_value < 0.05,
    star_label = ifelse(significant & !is.na(p_value), "*", "")
  )

# Get max flux values for star positioning
max_flux_by_date <- daily_flux %>%
  group_by(year, date) %>%
  summarise(max_flux = max(mean_flux + se_flux, na.rm = TRUE), .groups = "drop")

# Combine significance data with position data
star_data <- daily_significance %>%
  left_join(max_flux_by_date, by = c("year", "date")) %>%
  filter(star_label == "*") %>%
  mutate(star_y = max_flux * 1.1)

# Filter precipitation to match flux measurement periods only
daily_precip_trimmed <- rain %>%
  transmute(date, year, precipmm) %>%
  filter(year %in% c(2023, 2024)) %>%
  left_join(flux_date_ranges, by = "year") %>%
  filter(date >= min_date, date <= max_date) %>%
  select(date, year, precipmm)

# Fertilizer event dates
fert_events <- tibble(
  year = c(2023, 2024),
  fertilizer_date = as.Date(c("2023-05-05", "2024-07-17"))
)
```

# Create Figure 2 with Plot-Level Cumulative Uncertainty

```{r figure2-plotlevel}
#| label: fig-figure2-plotlevel
#| fig-cap: "Daily and cumulative N₂O flux with plot-level between-plot 95% CI"
#| fig-width: 13.3
#| fig-height: 7.5

# Calculate global maxima for consistent y-axes
global_max_flux <- max(daily_flux$mean_flux + daily_flux$se_flux, na.rm = TRUE)
global_max_precip <- max(daily_precip_trimmed$precipmm, na.rm = TRUE)
global_max_cum_flux <- max(cum_flux_summary$cum_flux_mean + cum_flux_summary$cum_flux_ci, na.rm = TRUE)

# Scale precipitation to fit on flux scale, positioned from top down
global_precip_scale_factor <- global_max_flux * 0.4 / global_max_precip
global_plot_top <- global_max_flux * 1.1

create_dual_axis_plot <- function(year_val, is_right_panel = FALSE) {
  # Filter data for the specific year
  flux_year <- daily_flux %>% filter(year == year_val)
  precip_year <- daily_precip_trimmed %>% filter(year == year_val)
  star_year <- star_data %>% filter(year == year_val)
  fert_year <- fert_events %>% filter(year == year_val)
  cum_flux_year <- cum_flux_summary %>% 
    filter(year == year_val) %>%
    rename(Treatment = treatment)
  cum_flux_points <- cum_flux_observed %>% 
    filter(year == year_val) %>%
    rename(Treatment = treatment)

  # Full-year limits (force Jan 1 → Dec 31)
  year_start <- as.Date(sprintf("%d-01-01", year_val))
  year_end <- as.Date(sprintf("%d-12-31", year_val))

  # Daily flux plot with precipitation
  daily_plot <- ggplot() +
    # Precipitation bars (from top down)
    geom_rect(data = precip_year,
              aes(xmin = date - 0.5, xmax = date + 0.5,
                  ymin = global_plot_top - precipmm * global_precip_scale_factor, ymax = global_plot_top),
              fill = "#377EB8", alpha = 0.7) +
    # Daily flux points
    geom_point(data = flux_year,
               aes(x = date, y = mean_flux, color = Treatment),
               size = 2) +
    geom_errorbar(data = flux_year,
                  aes(x = date, ymin = mean_flux - se_flux, ymax = mean_flux + se_flux, color = Treatment),
                  width = 0.2, alpha = 0.5) +
    # Fertilizer events
    geom_vline(data = fert_year,
               aes(xintercept = as.numeric(fertilizer_date)),
               linetype = "dashed", color = "red", alpha = 0.7) +
    # Significance stars
    geom_text(data = star_year,
              aes(x = date, y = star_y, label = star_label),
              color = "black", size = 8) +
    # Scales and styling
    scale_color_treatments() +
    guides(color = "none") +
    scale_x_date(date_breaks = "1 month", date_labels = "%b",
                 limits = c(year_start, year_end)) +
    scale_y_continuous(
      name = if (!is_right_panel) expression("Daily N"[2]*"O Flux (kg N ha"^-1*" day"^-1*")") else "",
      limits = c(0, global_plot_top),
      sec.axis = sec_axis(~ (global_plot_top - .) / global_precip_scale_factor,
                          name = if (is_right_panel) "Daily Precipitation (mm)" else "")
    ) +
    theme_publication() +
    theme(
      legend.position = "none",
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = if (is_right_panel) element_blank() else element_text(),
      axis.ticks.y = if (is_right_panel) element_blank() else element_line(),
      axis.title.y = if (is_right_panel) element_blank() else element_text(),
      axis.text.y.right = if (is_right_panel) element_text(color = "#377EB8") else element_blank(),
      axis.ticks.y.right = if (is_right_panel) element_line() else element_blank(),
      axis.title.y.right = if (is_right_panel) element_text(color = "#377EB8") else element_blank()
    ) +
    labs(title = paste("", year_val), x = "") +
    geom_hline(yintercept = 0, linetype = "solid", alpha = 0.3)

  # Cumulative flux plot with PLOT-LEVEL 95% CI
  cumulative_plot <- ggplot(cum_flux_year, aes(x = date, y = cum_flux_mean, color = Treatment)) +
    geom_ribbon(aes(ymin = pmax(cum_flux_mean - cum_flux_ci, 0), 
                    ymax = cum_flux_mean + cum_flux_ci, 
                    fill = Treatment),
                alpha = 0.2, linewidth = 0) +
    geom_line(linewidth = 1.1) +
    geom_point(data = cum_flux_points, 
               aes(x = date, y = cum_flux_mean, color = Treatment), 
               size = 1.8) +
    scale_color_treatments() +
    scale_fill_treatments() +
    scale_x_date(date_breaks = "1 month", date_labels = "%b",
                 limits = c(year_start, year_end)) +
    scale_y_continuous(
      name = if (!is_right_panel) expression("Cumulative N"[2]*"O Flux (kg N ha"^-1*")") else "",
      limits = c(0, global_max_cum_flux)
    ) +
    labs(x = "Month") +
    theme_publication() +
    theme(
      legend.position = "none",
      axis.text.y = if (is_right_panel) element_blank() else element_text(),
      axis.ticks.y = if (is_right_panel) element_blank() else element_line(),
      axis.title.y = if (is_right_panel) element_blank() else element_text()
    )

  # Combine daily and cumulative plots vertically
  (daily_plot / cumulative_plot) +
    patchwork::plot_layout(heights = c(1.2, 1))
}

# Create plots for both years
plot_2023 <- create_dual_axis_plot(2023, is_right_panel = FALSE)
plot_2024 <- create_dual_axis_plot(2024, is_right_panel = TRUE)

# Combine plots side by side
figure2 <- (plot_2023 | plot_2024) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom", 
        legend.justification = "center",
        legend.box.just = "center")

figure2
```

# Save Figure

```{r save-figure}
#| label: save-figure

# Save the figure with specified dimensions
ggsave(
  filename = "figures/figure2_flux_precip_cumulative_2023_2024_plotlevelCI.png",
  plot = figure2,
  width = 174,
  height = 174,
  units = "mm",
  dpi = 600,
  bg = "white"
)

cat("✓ Figure saved to figures/figure2_flux_precip_cumulative_2023_2024_plotlevelCI.png\n")
cat("  Dimensions: 174 mm × 174 mm @ 600 dpi\n")
```

# Summary

This figure implements a more statistically appropriate uncertainty method for cumulative N₂O flux:

1. **Plot-level calculation**: Cumulative flux is computed for each individual plot (Treatment × year × plot)
2. **Interpolation**: Daily fluxes are interpolated between measurement dates within each plot
3. **Between-plot variability**: The ribbon shows 95% CI based on the variability between plots within each treatment
4. **Statistical basis**: CI = t_crit × SE, where SE = SD / √n and t_crit comes from the t-distribution with n-1 degrees of freedom

This approach is more appropriate than propagating measurement error because it captures the actual biological variability between experimental units (plots) rather than measurement uncertainty within units.
