---
title: "Sorghum and Rye Project: Final Figures and Tables"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-overflow: wrap
    fig-width: 10
    fig-height: 6
    embed-resources: true
    theme: default
execute:
  echo: true
  warning: false
  message: false
  error: false
  cache: false
---

# Setup and Data Loading

```{r setup}
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(lubridate)
library(gt)
library(janitor)
library(broom)
library(readxl)
library(patchwork)

# Re-establish dplyr preferences after loading packages
if (requireNamespace("conflicted", quietly = TRUE)) {
  library(conflicted)
  conflicts_prefer(dplyr::select)
  conflicts_prefer(dplyr::filter)
  conflicts_prefer(dplyr::summarise)
}

# Custom theme for consistency
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.background = element_rect(fill = "#ffffffff", color = NA),
      plot.background = element_rect(fill = "#ffffffff", color = NA),
      panel.grid.major = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor = element_blank(),
      text = element_text(color = "#000000ff"),
      axis.text = element_text(color = "#000000ff"),
      axis.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold", size = base_size * 1.2),
      strip.text = element_text(face = "bold"),
      strip.background = element_rect(fill = alpha("#A5CAD2", 0.2)),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.spacing = unit(2, "lines")
    )
}

# Treatment colors
treatment_colors <- c(
  "Sorghum" = "#D8D97AFF",
  "Sorghum + Rye" = "#95C36EFF", 
  "Corn" = "#74C8C3FF",
  "Soy" = "#0A2E57FF"
)

scale_color_treatments <- function() scale_color_manual(values = treatment_colors)
scale_fill_treatments <- function() scale_fill_manual(values = treatment_colors)
```

# Data Import and Processing

```{r data-import}
#| message: false

# Import soil nitrogen data
# Nitrate and ammonia in ppm, not sure if moisture corrected
soil_data <- read_csv("data/soils/cleaned_soil_data.csv") %>%
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) %>%
  mutate(
    date = as.Date(date),
    total_inorganic_n = ammonia_ppm + nitrate_ppm
  ) 

# Import flux data 
# N2O flux in g N ha-1 day-1 
flux_data <- read_csv("data/cleaned_flux_data.csv") %>%
  mutate(date = as.Date(date)) 

# Import nitrogen leaching data
# Cumulative N loss in mg per day, same for nitrate and ammonium
leaching_data <- read_csv("data/clean_leaching.csv") %>%
  clean_names() %>%
  mutate(
    date = as.Date(date),
    year = year(date)
  ) %>%
  filter(year %in% c(2023, 2024)) %>%  # Filter out bad years
  filter(!is.na(year))

# Import historical rain and gdd data
gdd <- read_xlsx("data/rain/historical_rain.xlsx")

gdd <- gdd %>%
  select(day, doy, gdd_50_86) %>%
  mutate(
    date = as.Date(day),
    year = year(date),
  ) %>%
  group_by(year) %>%
  arrange(date) %>%
  mutate(cumulative_gdd = cumsum(gdd_50_86))

gdd <- gdd %>%
  mutate(doy_date = as.Date(doy - 1, origin = "2000-01-01"))

rain <- read_xlsx("data/rain/historical_rain.xlsx")

rain <- rain %>%
  select(day, doy, precipmm) %>%
  mutate(
    date = as.Date(day),
    year = year(date),
  ) %>%
  group_by(year) %>%
  arrange(date) %>%
  mutate(cumulative_precip = cumsum(precipmm))

rain <- rain %>%
  mutate(doy_date = as.Date(doy - 1, origin = "2000-01-01"))
```

# Helper Functions

```{r helper-functions}
# Function for pairwise t-tests within groups
#| label: get-pvals
get_pairwise_pvals <- function(data, variable, group_vars) {
  data %>%
    group_by(across(all_of(group_vars))) %>%
    group_modify(~ {
      treatments <- unique(.x$treatment)
      if (length(treatments) < 2) {
        return(tibble(p_value = NA_real_))
      }
      
      if (length(treatments) == 2) {
        t_result <- t.test(.x[[variable]][.x$treatment == treatments[1]], 
                          .x[[variable]][.x$treatment == treatments[2]])
        return(tibble(p_value = t_result$p.value))
      } else {
        pairwise_result <- pairwise.t.test(.x[[variable]], .x$treatment, 
                                          p.adjust.method = "none")
        return(tibble(p_value = min(pairwise_result$p.value, na.rm = TRUE)))
      }
    }) %>%
    ungroup()
}

# Function to add significance letters
add_significance_letters <- function(data, p_col) {
  data %>%
    mutate(
      significance_letter = case_when(
        is.na({{ p_col }}) ~ "",
        {{ p_col }} < 0.001 ~ "a",
        {{ p_col }} < 0.01 ~ "b", 
        {{ p_col }} < 0.05 ~ "c",
        TRUE ~ ""
      ),
      result_with_sig = if_else(
        significance_letter == "", 
        result,
        paste0(result, " ", significance_letter)
      )
    )
}

# Function to create ANOVA source tables
create_source_table <- function(model_results) {
  tibble(
    source = c("Date", "Treatment", "Date × Treatment"),
    p_value = summary(model_results)[[1]][c(2, 3, 6), "Pr(>F)"]
  ) %>%
    mutate(
      formatted_p = case_when(
        p_value < 0.001 ~ "< 0.001",
        p_value < 0.01 ~ sprintf("%.3f", p_value),
        TRUE ~ sprintf("%.2f", p_value)
      )
    )
}
```

# Table 2: Salt-extractable Inorganic Nitrogen

```{r table-2}
#| label: tbl-soil-nitrogen
#| tbl-cap: "Salt-extractable inorganic nitrogen"

# Calculate summary statistics
soil_summary <- soil_data %>%
  group_by(year, date, treatment) %>%
  summarise(
    across(c(ammonia_ppm, nitrate_ppm, total_inorganic_n), 
           list(mean = ~ mean(.x, na.rm = TRUE),
                se = ~ sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x))))),
    .groups = "drop"
  ) 

# Get p-values for pairwise comparisons
pvals_ammonium <- get_pairwise_pvals(soil_data, "ammonia_ppm", c("year", "date"))
pvals_nitrate <- get_pairwise_pvals(soil_data, "nitrate_ppm", c("year", "date"))  
pvals_total_n <- get_pairwise_pvals(soil_data, "total_inorganic_n", c("year", "date"))

# Combine with p-values and format
soil_table_data <- soil_summary %>%
  left_join(pvals_ammonium %>% rename(ammonia_p = p_value), by = c("year", "date")) %>%
  left_join(pvals_nitrate %>% rename(nitrate_p = p_value), by = c("year", "date")) %>%
  left_join(pvals_total_n %>% rename(total_n_p = p_value), by = c("year", "date")) %>%
  pivot_longer(
    cols = c(ammonia_ppm_mean, ammonia_ppm_se, nitrate_ppm_mean, nitrate_ppm_se, 
             total_inorganic_n_mean, total_inorganic_n_se),
    names_to = c("variable", "stat"),
    names_pattern = "(.*)_(mean|se)",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(
    result = sprintf("%.2f ± %.2f", mean, se),
    p_val = case_when(
      variable == "ammonia_ppm" ~ ammonia_p,
      variable == "nitrate_ppm" ~ nitrate_p,
      variable == "total_inorganic_n" ~ total_n_p
    ),
    variable = case_when(
      variable == "ammonia_ppm" ~ "ammonium",
      variable == "nitrate_ppm" ~ "nitrate", 
      variable == "total_inorganic_n" ~ "total_n"
    )
  ) %>%
  add_significance_letters(p_val) %>%
  select(year, date, treatment, variable, result_with_sig) %>%
  pivot_wider(names_from = variable, values_from = result_with_sig)

# Format dates and years for display
soil_table_formatted <- soil_table_data %>%
  mutate(
    year = as.character(year),
    date = format(as.Date(date), "%B %d")
  ) %>%
  arrange(year, as.Date(paste(year, date), format = "%Y %B %d"), treatment) %>%
  group_by(year) %>%
  mutate(
    year_display = if_else(row_number() == 1, year, ""),
    is_new_date = date != lag(date, default = "")
  ) %>%
  ungroup()

# Create ANOVA source of variation
aov_ammonium <- aov(ammonia_ppm ~ year * date * treatment, data = soil_data)
aov_nitrate <- aov(nitrate_ppm ~ year * date * treatment, data = soil_data)
aov_total_n <- aov(total_inorganic_n ~ year * date * treatment, data = soil_data)

source_variation <- tibble(
  source = c("Date", "Treatment", "Date × Treatment"),
  ammonium = summary(aov_ammonium)[[1]][c(2, 3, 6), "Pr(>F)"],
  nitrate = summary(aov_nitrate)[[1]][c(2, 3, 6), "Pr(>F)"],
  total_n = summary(aov_total_n)[[1]][c(2, 3, 6), "Pr(>F)"]
) %>%
  mutate(across(c(ammonium, nitrate, total_n), ~ case_when(
    .x < 0.001 ~ "< 0.001",
    .x < 0.01 ~ sprintf("%.3f", .x),
    TRUE ~ sprintf("%.2f", .x)
  ))) %>%
  mutate(
    year_display = "",
    date = "",
    treatment = source,
    .before = 1
  ) %>%
  select(-source)

# Combine data with source of variation
separator_row <- tibble(
  year_display = "Source of Variation", date = "", treatment = "",
  ammonium = "", nitrate = "", total_n = "", is_new_date = FALSE
)

table2_complete <- bind_rows(
  soil_table_formatted,
  separator_row,
  source_variation %>% mutate(is_new_date = FALSE)
)

# Create gt table
table2_gt <- table2_complete %>%
  select(-year) %>%
  select(year_display, date, treatment, ammonium, nitrate, total_n, is_new_date) %>%
  gt() %>%
  tab_header(title = md("**Table 2. Salt-extractable inorganic nitrogen**")) %>%
  cols_label(
    year_display = "Year",
    date = "Date", 
    treatment = "Treatment",
    ammonium = md("Ammonium<br>(mg kg<sup>-1</sup>)"),
    nitrate = md("Nitrate<br>(mg kg<sup>-1</sup>)"),
    total_n = md("Total Inorganic N<br>(mg kg<sup>-1</sup>)")
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", weight = px(1), color = "gray"),
    locations = cells_body(rows = is_new_date == TRUE)
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", weight = px(2)),
    locations = cells_body(rows = year_display == "Source of Variation")
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = year_display == "Source of Variation")
  ) %>%
  tab_style(
    style = cell_text(align = "center", weight = "bold"),
    locations = cells_body(columns = year_display, rows = year_display == "Source of Variation")
  ) %>%
  cols_hide(columns = is_new_date) %>%
  tab_footnote(
    footnote = "Letters indicate significance levels: a = p < 0.001, b = p < 0.01, c = p < 0.05",
    locations = cells_title()
  )

# Save table
gtsave(table2_gt, "tables/table2_soil_nitrogen.rtf")

table2_gt
```

# N2O Data Processing and Analysis

```{r n2o-processing}
# Process daily flux data with interpolation
daily_flux_summary <- flux_data %>%
  group_by(plot, rowvs_interrow, treatment, year, date) %>%
  summarise(mean_flux = mean(gnha_day_no_negative, na.rm = TRUE), .groups = "drop") %>%
  arrange(plot, rowvs_interrow, treatment, year, date)

# Interpolation function
interpolate_daily_flux <- function(data) {
  # Get unique plot combinations
  plot_combinations <- data %>%
    distinct(plot, rowvs_interrow, treatment, year)
  
  # Interpolate for each combination
  map_dfr(seq_len(nrow(plot_combinations)), function(i) {
    combo <- plot_combinations[i, ]
    
    subset_data <- data %>%
      filter(
        plot == combo$plot,
        rowvs_interrow == combo$rowvs_interrow,
        treatment == combo$treatment,
        year == combo$year
      ) %>%
      arrange(date)
    
    # Create full year sequence
    start_date <- as.Date(paste0(combo$year, "-01-01"))
    end_date <- as.Date(paste0(combo$year, "-12-31"))
    all_days <- seq(start_date, end_date, by = "day")
    
    # Interpolate if we have sufficient data
    if (nrow(subset_data) >= 2 && !all(is.na(subset_data$mean_flux))) {
      valid_data <- subset_data[!is.na(subset_data$mean_flux), ]
      
      if (nrow(valid_data) >= 2) {
        interpolated_flux <- approx(
          x = as.numeric(valid_data$date),
          y = valid_data$mean_flux,
          xout = as.numeric(all_days),
          rule = 1
        )$y
      } else {
        interpolated_flux <- rep(NA_real_, length(all_days))
      }
    } else {
      interpolated_flux <- rep(NA_real_, length(all_days))
    }
    
    tibble(
      plot = combo$plot,
      rowvs_interrow = combo$rowvs_interrow,
      treatment = combo$treatment,
      year = combo$year,
      date = all_days,
      interpolated_flux = interpolated_flux
    )
  })
}

# Apply interpolation
interpolated_flux_data <- interpolate_daily_flux(daily_flux_summary)

# Calculate cumulative N2O per plot
plot_cumulative_n2o <- interpolated_flux_data %>%
  group_by(plot, rowvs_interrow, treatment, year) %>%
  summarise(
    annual_cumulative_n2o = sum(interpolated_flux, na.rm = TRUE) / 1000, # Convert g to kg
    .groups = "drop"
  ) %>%
  filter(is.finite(annual_cumulative_n2o))

cat("✓ N2O interpolation and cumulative calculations complete\n")
cat("Plot-level cumulative N2O data:", nrow(plot_cumulative_n2o), "observations\n")
```

# Nitrogen Leaching Data Processing

```{r leaching-processing}
# Calculate annual cumulative nitrogen leaching per plot
# Convert mg per plot to kg N ha-1
plot_area_ft2 <- 120*160  # in ft^2
plot_area_ha <- plot_area_ft2 / 107639.1041671  # Convert ft^2 to ha
annual_leaching <- leaching_data %>%
  group_by(plot, treatment, year) %>%
  summarise(
    annual_cumulative_leaching = (max(cumulative_n_loss_mg, na.rm = TRUE) / 1000000) / plot_area_ha, # mg to kg to kg/ha
    .groups = "drop"
  ) %>%
  filter(is.finite(annual_cumulative_leaching), annual_cumulative_leaching >= 0)

cat("✓ Nitrogen leaching calculations complete\n")
cat("Plot-level leaching data:", nrow(annual_leaching), "observations\n")
```

# Table 3: Annual Nitrogen Losses

```{r table-3}
#| label: tbl-nitrogen-losses
#| tbl-cap: "Annual nitrogen losses by treatment"

# Summarize N2O data by treatment and year
n2o_summary <- plot_cumulative_n2o %>%
  group_by(treatment, year) %>%
  summarise(
    n_plots = n(),
    mean_n2o = mean(annual_cumulative_n2o, na.rm = TRUE),
    sd_n2o = sd(annual_cumulative_n2o, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(n2o_result = sprintf("%.3f ± %.3f", mean_n2o, sd_n2o))

# Summarize leaching data by treatment and year
leaching_summary <- annual_leaching %>%
  group_by(treatment, year) %>%
  summarise(
    n_plots = n(),
    mean_leaching = mean(annual_cumulative_leaching, na.rm = TRUE),
    sd_leaching = sd(annual_cumulative_leaching, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(leaching_result = sprintf("%.3f ± %.3f", mean_leaching, sd_leaching))

# Calculate 2-year totals
n2o_two_year <- plot_cumulative_n2o %>%
  group_by(plot, rowvs_interrow, treatment) %>%
  summarise(total_n2o = sum(annual_cumulative_n2o, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  summarise(
    mean_n2o = mean(total_n2o, na.rm = TRUE),
    sd_n2o = sd(total_n2o, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    year = "2023-2024",
    n2o_result = sprintf("%.3f ± %.3f", mean_n2o, sd_n2o)
  )

leaching_two_year <- annual_leaching %>%
  group_by(plot, treatment) %>%
  summarise(total_leaching = sum(annual_cumulative_leaching, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  summarise(
    mean_leaching = mean(total_leaching, na.rm = TRUE),
    sd_leaching = sd(total_leaching, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    year = "2023-2024", 
    leaching_result = sprintf("%.3f ± %.3f", mean_leaching, sd_leaching)
  )
compare_df_cols(n2o_two_year, leaching_two_year, n2o_summary, leaching_summary)
# Combine annual and 2-year data
combined_nitrogen_data <- bind_rows(
  n2o_summary %>% mutate(year = as.character(year)) %>% select(year, treatment, n2o_result),
  n2o_two_year %>% select(year, treatment, n2o_result)
) %>%
  full_join(
    bind_rows(
      leaching_summary %>% mutate(year = as.character(year)) %>% select(year, treatment, leaching_result),
      leaching_two_year %>% select(year, treatment, leaching_result)
    ),
    by = c("year", "treatment")
  ) %>%
  mutate(
    year_order = case_when(
      year == "2023" ~ 1,
      year == "2024" ~ 2,
      year == "2023-2024" ~ 3
    ),
    total_environmental_cost = "TBD" # Placeholder for future calculation
  ) %>%
  arrange(year_order, treatment)

# Source of variation analysis
# N2O ANOVA
combined_n2o_data <- plot_cumulative_n2o %>% mutate(year_factor = as.factor(year))
aov_n2o <- aov(annual_cumulative_n2o ~ year_factor * treatment, data = combined_n2o_data)

# Leaching ANOVA  
combined_leaching_data <- annual_leaching %>% mutate(year_factor = as.factor(year))
aov_leaching <- aov(annual_cumulative_leaching ~ year_factor * treatment, data = combined_leaching_data)

# Source of variation table
source_variation_n <- tibble(
  source = c("Year", "Treatment", "Year × Treatment"),
  n2o_p = summary(aov_n2o)[[1]][1:3, "Pr(>F)"],
  leaching_p = summary(aov_leaching)[[1]][1:3, "Pr(>F)"]
) %>%
  mutate(
    across(c(n2o_p, leaching_p), ~ case_when(
      .x < 0.001 ~ "< 0.001",
      .x < 0.01 ~ sprintf("%.3f", .x),
      TRUE ~ sprintf("%.2f", .x)
    )),
    year = "",
    year_order = 4,
    treatment = source,
    n2o_result = n2o_p,
    leaching_result = leaching_p,
    total_environmental_cost = "TBD"
  ) %>%
  select(-source, -n2o_p, -leaching_p)

# Create complete table
separator_n <- tibble(
  year = "Source of Variation", year_order = 3.5, treatment = "",
  n2o_result = "", leaching_result = "", total_environmental_cost = ""
)

table3_complete <- bind_rows(
  combined_nitrogen_data,
  separator_n,
  source_variation_n
) %>%
  arrange(year_order, treatment) %>%
  mutate(
    year_display = case_when(
      year == "Source of Variation" ~ year,
      year != lag(year, default = "") ~ as.character(year),
      TRUE ~ ""
    ),
    is_new_section = year != lag(year, default = "") & year != "Source of Variation"
  )

# Create gt table
table3_gt <- table3_complete %>%
  select(
    Year = year_display,
    Treatment = treatment,
    `N2O Loss` = n2o_result,
    `N Leaching` = leaching_result, 
    `Total Environmental Cost` = total_environmental_cost,
    is_new_section
  ) %>%
  gt() %>%
  tab_header(title = md("**Table 3. Annual nitrogen losses**")) %>%
  cols_label(
    `N2O Loss` = md("N<sub>2</sub>O Loss<br>(kg N ha<sup>-1</sup>)"),
    `N Leaching` = md("N Leaching<br>(kg N ha<sup>-1</sup>)"),
    `Total Environmental Cost` = md("Total Environmental Cost<br>(kg N equivalent)")
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", weight = px(1), color = "gray"),
    locations = cells_body(rows = is_new_section == TRUE)
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", weight = px(2)),
    locations = cells_body(rows = Year == "Source of Variation")
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = Year == "Source of Variation")  
  ) %>%
  cols_hide(columns = is_new_section) %>%
  tab_footnote(
    footnote = "Values are mean ± standard deviation. Letters indicate significance levels: a = p < 0.001, b = p < 0.01, c = p < 0.05",
    locations = cells_title()
  )

# Save table
gtsave(table3_gt, "tables/table3_nitrogen_losses.rtf")

table3_gt
```

# Figure: test: Flow and Nitrogen Leaching Relationship

This will give a good idea of the variation between flow and nitrogen leaching across treatments and years.

```{r debug-leaching-data}
#| label: debug-leaching
#| echo: true

# Let's examine the leaching data structure to understand the issue
cat("=== LEACHING DATA STRUCTURE ===\n")
cat("Total rows in leaching_data:", nrow(leaching_data), "\n")
cat("Years present:", paste(sort(unique(leaching_data$year)), collapse = ", "), "\n")
cat("Treatments present:", paste(sort(unique(leaching_data$treatment)), collapse = ", "), "\n")
cat("Plots present:", paste(sort(unique(leaching_data$plot)), collapse = ", "), "\n\n")

# Check treatment-year combinations
cat("=== TREATMENT-YEAR COMBINATIONS ===\n")
treatment_year_summary <- leaching_data %>%
  group_by(year, treatment) %>%
  summarise(
    n_observations = n(),
    unique_plots = length(unique(plot)),
    plots = paste(sort(unique(plot)), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(year, treatment)

print(treatment_year_summary)

# Check for potential data issues
cat("\n=== POTENTIAL ISSUES ===\n")
duplicate_check <- leaching_data %>%
  group_by(plot, treatment, year, date) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n > 1)

if(nrow(duplicate_check) > 0) {
  cat("WARNING: Duplicate plot-treatment-year-date combinations found!\n")
  print(head(duplicate_check))
} else {
  cat("No duplicate plot-treatment-year-date combinations found.\n")
}

# Check if same plot appears in multiple treatments within same year
plot_treatment_check <- leaching_data %>%
  distinct(plot, treatment, year) %>%
  group_by(plot, year) %>%
  summarise(treatments = paste(sort(treatment), collapse = ", "), n_treatments = n(), .groups = "drop") %>%
  filter(n_treatments > 1)

if(nrow(plot_treatment_check) > 0) {
  cat("\nWARNING: Plots appear in multiple treatments within same year:\n")
  print(plot_treatment_check)
} else {
  cat("\nGood: Each plot has only one treatment per year.\n")
}
```

```{r figure-test}
#| label: fig-flow-leaching
#| fig-cap: "Annual tile drain flow and cumulative nitrogen losses by treatment"
#| fig-width: 10
#| fig-height: 6

# Calculate annual totals from leaching data - INDIVIDUAL PLOTS
plot_flow_n_data <- leaching_data %>%
  group_by(plot, treatment, year) %>%
  summarise(
    total_flow_L = sum(flow_l, na.rm = TRUE),  # Sum daily flow to get annual total
    total_n_loss_mg = max(cumulative_n_loss_mg, na.rm = TRUE),  # Use existing cumulative N
    .groups = "drop"
  ) %>%
  # Remove any infinite or negative values
  filter(is.finite(total_flow_L), is.finite(total_n_loss_mg), 
         total_flow_L >= 0, total_n_loss_mg >= 0) %>%
  # Create plot labels and grouping variables  
  mutate(
    plot_label = paste("Plot", plot),
    year_treatment = paste(year, treatment, sep = " "),
    treatment_year = paste(treatment, year, sep = "_"),
    # Create unique plot identifier within treatment-year
    plot_id = paste(plot, treatment, year, sep = "_")
  ) %>%
  # Arrange by treatment and year for better visual grouping
  arrange(year, treatment, plot)

# Check the data structure - should match expected sample sizes
cat("Plots per treatment-year combination:\n")
print(plot_flow_n_data %>% 
  group_by(year, treatment) %>% 
  summarise(n_plots = n(), plots = paste(sort(unique(plot)), collapse = ", "), .groups = "drop"))

# Create individual plot bars showing variation within treatments
flow_n_plot <- ggplot(plot_flow_n_data, aes(x = year_treatment, group = plot)) +
  # Flow bars for individual plots - side by side
  geom_col(aes(y = total_flow_L, fill = treatment), 
           position = position_dodge(width = 0.8), alpha = 0.8, width = 0.15) +
  # Nitrogen points (scaled to flow axis) for individual plots
  geom_point(aes(y = total_n_loss_mg * max(total_flow_L, na.rm = TRUE) / max(total_n_loss_mg, na.rm = TRUE), 
                 color = treatment), 
             position = position_dodge(width = 0.8), size = 4, shape = 18) +
  # Scales and formatting
  scale_fill_treatments() +
  scale_color_treatments() +
  scale_y_continuous(
    name = "Annual Tile Drain Flow (L)",
    labels = scales::comma,
    sec.axis = sec_axis(
      trans = ~ . * max(plot_flow_n_data$total_n_loss_mg, na.rm = TRUE) / max(plot_flow_n_data$total_flow_L, na.rm = TRUE),
      name = "Cumulative N Loss (mg)",
      labels = scales::comma
    )
  ) +
  labs(
    x = "Year and Treatment",
    title = "Individual Plot Flow and Nitrogen Losses by Treatment",
    subtitle = "Each bar = individual plot. Bars = Flow (L), Diamonds = N Loss (mg). Shows plot-to-plot variation.",
    fill = "Treatment",
    color = "Treatment"
  ) +
  theme_publication() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "bottom",
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.title.y.right = element_text(angle = 90, vjust = 0.5),
    panel.grid.major.x = element_blank()
  ) +
  # Add subtle separators between treatment groups
  facet_grid(. ~ year, scales = "free_x", space = "free_x", 
             labeller = labeller(year = function(x) paste("Year", x)))

flow_n_plot

ggsave("figures/figure2_flow_nitrogen_relationship.png", width = 10, height = 6, dpi = 300)
```

# Figure 1: Precipitation and Temperature 

```{r figure-weather}
#| label: fig-weather
#| fig-cap: "Daily precipitation, temperature, and soil moisture over the study period"
#| fig-width: 10
#| fig-height: 6

# Exclude current years from the envelope
hist_env <- gdd %>%
  filter(!(year %in% c(2023, 2024))) %>%
  group_by(doy_date) %>%
  summarize(
    min_cum = min(cumulative_gdd, na.rm = TRUE),
    max_cum = max(cumulative_gdd, na.rm = TRUE)
  )

gdd_2023 <- filter(gdd, year == 2023)
gdd_2024 <- filter(gdd, year == 2024)

ggplot() +
  # Shaded region for historical envelope
  geom_ribbon(
    data = hist_env,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  # 2023 solid line
  geom_line(
    data = gdd_2023,
    aes(x = doy_date, y = cumulative_gdd),
    color = "#E41A1C", size = 1.2, linetype = "solid"
  ) +
  # 2024 dashed line
  geom_line(
    data = gdd_2024,
    aes(x = doy_date, y = cumulative_gdd),
    color = "#377EB8", size = 1.2, linetype = "dashed"
  ) +
  scale_x_date(
    date_breaks = "1 month", # Breaks at each month
    date_labels = "%b" # Label as abbreviated month names
  ) +
  labs(
    title = "Cumulative Rainfall: Historical Range vs. Recent Years",
    x = "Day of Year",
    y = "Cumulative Precipitation (mm)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # Custom legend via manual color/linetype mapping:
  scale_color_manual(
    name = "Year",
    values = c("2023" = "#E41A1C", "2024" = "#377EB8")
  ) +
  scale_linetype_manual(
    name = "Year",
    values = c("2023" = "solid", "2024" = "dashed")
  )

gdd_highlight <- gdd %>%
  filter(year %in% c(2023, 2024)) %>%
  mutate(year = as.character(year))

ggplot() +
  geom_ribbon(
    data = hist_env,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  geom_line(
    data = gdd_highlight,
    aes(x = doy_date, y = cumulative_gdd, color = year, linetype = year),
    size = 1.2
  ) +
  scale_color_manual(
    name = "Year",
    values = c("2023" = "#E41A1C", "2024" = "#377EB8"),
    labels = c("2023", "2024")
  ) +
  scale_linetype_manual(
    name = "Year",
    values = c("2023" = "solid", "2024" = "dashed"),
    labels = c("2023", "2024")
  ) +
  scale_x_date(
    date_breaks = "1 month", # Breaks at each month
    date_labels = "%b" # Label as abbreviated month names
  ) +
  labs(
    title = "Cumulative Rainfall: Historical Range vs. Recent Years",
    x = "Day of Year",
    y = "Cumulative Growing Degree Days (GDD)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

mean_cum <- gdd %>%
  filter(!(year %in% c(2023, 2024))) %>% # use only historical for mean
  group_by(doy_date) %>%
  summarize(cumulative_gdd = mean(cumulative_gdd, na.rm = TRUE)) %>%
  mutate(year = "Mean") %>%
  filter(doy_date != "2000-12-31")

gdd_highlight <- gdd %>%
  filter(year %in% c(2023, 2024)) %>%
  mutate(year = as.character(year)) %>%
  bind_rows(mean_cum)

gdd_plot_v2 <- ggplot() +
  geom_ribbon(
    data = hist_env,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  geom_line(
    data = gdd_highlight,
    aes(x = doy_date, y = cumulative_gdd, color = year, linetype = year),
    size = 1.2
  ) +
  scale_color_manual(
    name = "Year",
    values = c(
      "2023" = "#E41A1C",
      "2024" = "#377EB8",
      "Mean" = "black"
    ),
    labels = c(
      "2023" = "2023",
      "2024" = "2024",
      "Mean" = "Mean (Historical)"
    )
  ) +
  scale_linetype_manual(
    name = "Year",
    values = c(
      "2023" = "solid",
      "2024" = "dashed",
      "Mean" = "dotdash"
    ),
    labels = c(
      "2023" = "2023",
      "2024" = "2024",
      "Mean" = "Mean (Historical)"
    )
  ) +
  scale_x_date(
    date_breaks = "1 month", # Breaks at each month
    date_labels = "%b" # Label as abbreviated month names
  ) +
  labs(
    x = "",  # Remove x-axis label since it aligns with bottom plots
    y = "Cumulative Growing Degree Days (GDD)"
  ) +
  theme_publication() +
  theme(
    axis.text.x = element_blank(),  # Remove x-axis text since it aligns with bottom plots
    axis.ticks.x = element_blank(),  # Remove x-axis ticks
    legend.position = "top"
  )

hist_env <- rain %>%
  filter(!(year %in% c(2023, 2024))) %>%
  group_by(doy_date) %>%
  summarize(
    min_cum = min(cumulative_precip, na.rm = TRUE),
    max_cum = max(cumulative_precip, na.rm = TRUE)
  )

rain_2023 <- filter(rain, year == 2023)
rain_2024 <- filter(rain, year == 2024)

ggplot() +
  # Shaded region for historical envelope
  geom_ribbon(
    data = hist_env,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  # 2023 solid line
  geom_line(
    data = rain_2023,
    aes(x = doy_date, y = cumulative_precip),
    color = "#E41A1C", size = 1.2, linetype = "solid"
  ) +
  # 2024 dashed line
  geom_line(
    data = rain_2024,
    aes(x = doy_date, y = cumulative_precip),
    color = "#377EB8", size = 1.2, linetype = "dashed"
  ) +
  scale_x_date(
    date_breaks = "1 month", # Breaks at each month
    date_labels = "%b" # Label as abbreviated month names
  ) +
  labs(
    title = "Cumulative Rainfall: Historical Range vs. Recent Years",
    x = "Day of Year",
    y = "Cumulative Precipitation (mm)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # Custom legend via manual color/linetype mapping:
  scale_color_manual(
    name = "Year",
    values = c("2023" = "#E41A1C", "2024" = "#377EB8")
  ) +
  scale_linetype_manual(
    name = "Year",
    values = c("2023" = "solid", "2024" = "dashed")
  )

rain_highlight <- rain %>%
  filter(year %in% c(2023, 2024)) %>%
  mutate(year = as.character(year))

ggplot() +
  geom_ribbon(
    data = hist_env,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  geom_line(
    data = rain_highlight,
    aes(x = doy_date, y = cumulative_precip, color = year, linetype = year),
    size = 1.2
  ) +
  scale_color_manual(
    name = "Year",
    values = c("2023" = "#E41A1C", "2024" = "#377EB8"),
    labels = c("2023", "2024")
  ) +
  scale_linetype_manual(
    name = "Year",
    values = c("2023" = "solid", "2024" = "dashed"),
    labels = c("2023", "2024")
  ) +
  scale_x_date(
    date_breaks = "1 month", # Breaks at each month
    date_labels = "%b" # Label as abbreviated month names
  ) +
  labs(
    title = "Cumulative Rainfall: Historical Range vs. Recent Years",
    x = "Day of Year",
    y = "Cumulative Precipitation (mm)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

mean_cum <- rain %>%
  filter(!(year %in% c(2023, 2024))) %>% # use only historical for mean
  group_by(doy_date) %>%
  summarize(cumulative_precip = mean(cumulative_precip, na.rm = TRUE)) %>%
  mutate(year = "Mean") %>%
  filter(doy_date != "2000-12-31")

rain_highlight <- rain %>%
  filter(year %in% c(2023, 2024)) %>%
  mutate(year = as.character(year)) %>%
  bind_rows(mean_cum)

rain_plot_v2 <- ggplot() +
  geom_ribbon(
    data = hist_env,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  geom_line(
    data = rain_highlight,
    aes(x = doy_date, y = cumulative_precip, color = year, linetype = year),
    size = 1.2
  ) +
  scale_color_manual(
    name = "Year",
    values = c(
      "2023" = "#E41A1C",
      "2024" = "#377EB8",
      "Mean" = "black"
    ),
    labels = c(
      "2023" = "2023",
      "2024" = "2024",
      "Mean" = "Mean (Historical)"
    )
  ) +
  scale_linetype_manual(
    name = "Year",
    values = c(
      "2023" = "solid",
      "2024" = "dashed",
      "Mean" = "dotdash"
    ),
    labels = c(
      "2023" = "2023",
      "2024" = "2024",
      "Mean" = "Mean (Historical)"
    )
  ) +
  scale_x_date(
    date_breaks = "1 month", # Breaks at each month
    date_labels = "%b" # Label as abbreviated month names
  ) +
  labs(
    x = "",  # Remove x-axis label since it aligns with bottom plots
    y = "Cumulative Precipitation (mm)"
  ) +
  theme_publication() +
  theme(
    axis.text.x = element_blank(),  # Remove x-axis text since it aligns with bottom plots
    axis.ticks.x = element_blank(),  # Remove x-axis ticks
    legend.position = "top"
  )

rain_plot_v2
# Modify pdplot to have centered legend
pdplot <- rain_plot_v2 | gdd_plot_v2 

ggsave(
  "figures/gdd_plot_v2.png",
  pdplot,
  width = 8.5, height = 11, dpi = 350
)
pdplot
```

# Figure 1: Soil Moisture and Temperature (with per-date significance)
```{r}
#| label: fig-soil-moisture-temp
#| fig-cap: "Daily soil moisture and temperature over the study period (asterisks indicate p < 0.05 between treatments for that date)"
#| fig-width: 10
#| fig-height: 8

moist_temp <- flux_data %>%
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) %>%
  rename(Treatment = treatment) %>%
  group_by(Treatment, year, date) %>%
  summarise(
    mean_soil_moisture = mean(swc_2_mean, na.rm = TRUE),
    mean_soil_temp = mean(ts_2_mean, na.rm = TRUE),
    .groups = "drop"
  )
colnames(moist_temp)
# --- Per-date (year + date) t-tests between treatments on raw replicate data ---
raw_soil <- flux_data %>%
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) %>%
  rename(Treatment = treatment) %>%
  mutate(year = year(date))

colnames(raw_soil)

get_date_pvals <- function(df, value_col) {
  df %>%
    group_by(year, date) %>%
    summarise(
      p = {
        vals <- .data[[value_col]]
        trt  <- Treatment
        # Need exactly the two treatments and at least one non-NA in each
        has_two <- length(unique(na.omit(trt))) == 2
        counts  <- table(trt[!is.na(vals)])
        ok <- has_two && all(counts >= 1) && length(na.omit(vals)) >= 2
        if (ok) {
          out <- tryCatch(t.test(vals ~ trt)$p.value, error = function(e) NA_real_)
          out
        } else NA_real_
      },
      .groups = "drop"
    ) %>%
    filter(!is.na(p) & p < 0.05)
}

pvals_moist <- get_date_pvals(raw_soil, "swc_2_mean")
pvals_temp  <- get_date_pvals(raw_soil, "ts_2_mean")

# Star positions (5% headroom above that date's max mean)
moisture_stars <- moist_temp %>%
  group_by(year, date) %>%
  summarise(y_star = max(mean_soil_moisture, na.rm = TRUE) * 1.05, .groups = "drop") %>%
  inner_join(pvals_moist, by = c("year", "date")) %>%
  filter(is.finite(y_star))

temp_stars <- moist_temp %>%
  group_by(year, date) %>%
  summarise(y_star = max(mean_soil_temp, na.rm = TRUE) * 1.05, .groups = "drop") %>%
  inner_join(pvals_temp, by = c("year", "date")) %>%
  filter(is.finite(y_star))

# Split data by year
moist_temp_2023 <- filter(moist_temp, year == 2023)
moist_temp_2024 <- filter(moist_temp, year == 2024)

# 2023 Soil Moisture (top left)
p1 <- ggplot(moist_temp_2023, aes(x = date, y = mean_soil_moisture, color = Treatment, shape = Treatment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_treatments() +
  scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + Rye" = 15)) +  # circle and square
  scale_x_date(date_breaks = "1 month", date_labels = "%b", 
               limits = c(as.Date("2023-01-01"), as.Date("2023-12-31"))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  labs(
    x = "",  # No x-axis label for top plots
    y = "2023 Soil Moisture (m³/m³)"
  ) +
  geom_text(data = moisture_stars %>% filter(year == 2023),
            aes(x = date, y = y_star), label = "*", color = "black", size = 4, vjust = 0, inherit.aes = FALSE) +
  theme_publication() +
  theme(
    axis.text.x = element_blank(),  # Remove x-axis text for top plots
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

# 2024 Soil Moisture (bottom left)  
p2 <- ggplot(moist_temp_2024, aes(x = date, y = mean_soil_moisture, color = Treatment, shape = Treatment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_treatments() +
  scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + Rye" = 15)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b", 
               limits = c(as.Date("2024-01-01"), as.Date("2024-12-31"))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  labs(
    x = "Month",
    y = "2024 Soil Moisture (m³/m³)"
  ) +
  geom_text(data = moisture_stars %>% filter(year == 2024),
            aes(x = date, y = y_star), label = "*", color = "black", size = 4, vjust = 0, inherit.aes = FALSE) +
  theme_publication() +
  theme(
    legend.position = "bottom"
  ) +
  labs(color = "Treatment", shape = "Treatment")

# 2023 Soil Temperature (top right)
p3 <- ggplot(moist_temp_2023, aes(x = date, y = mean_soil_temp, color = Treatment, shape = Treatment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_treatments() +
  scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + Rye" = 15)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b", 
               limits = c(as.Date("2023-01-01"), as.Date("2023-12-31"))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  labs(
    x = "",
    y = "2023 Soil Temperature (°C)"
  ) +
  geom_text(data = temp_stars %>% filter(year == 2023),
            aes(x = date, y = y_star), label = "*", color = "black", size = 4, vjust = 0, inherit.aes = FALSE) +
  theme_publication() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

# 2024 Soil Temperature (bottom right)
p4 <- ggplot(moist_temp_2024, aes(x = date, y = mean_soil_temp, color = Treatment, shape = Treatment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_treatments() +
  scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + Rye" = 15)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b", 
               limits = c(as.Date("2024-01-01"), as.Date("2024-12-31"))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  labs(
    x = "Month",
    y = "2024 Soil Temperature (°C)",
    color = "Treatment",
    shape = "Treatment"
  ) +
  geom_text(data = temp_stars %>% filter(year == 2024),
            aes(x = date, y = y_star), label = "*", color = "black", size = 4, vjust = 0, inherit.aes = FALSE) +
  theme_publication() +
  theme(legend.position = "none")
p4
# Modify soil_4panel to also have centered top legend
soil_4panel <- (p1 | p3) / (p2 | p4) 

six_panel_plot <-
  (pdplot / soil_4panel) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(
    legend.position      = "top",
    legend.justification = "center",
    legend.direction     = "horizontal",
    legend.box           = "horizontal",   # keeps multiple guides in one row when possible
    legend.margin        = margin(b = 4)
  )

ggsave("figures/six_panel_weather_soil.png", six_panel_plot, width = 10, height = 12, dpi = 300)
six_panel_plot
```

# Summary

This document provides clean, efficient analysis following tidyverse principles:

-   **Consistent data processing** using `group_by()` and `summarise()`
-   **Functional programming** with custom helper functions
-   **Clear separation** between data processing and visualization
-   **Reproducible workflow** with proper chunk organization
-   **Professional styling** consistent across all outputs

Each table and figure is generated in its own chunk for modularity and clarity.