---
title: "Sorghum and Rye Project: Final Figures and Tables"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-overflow: wrap
    fig-width: 10
    fig-height: 6
    embed-resources: true
    theme: default
execute:
  echo: true
  warning: false
  message: false
  error: false
  cache: false
---

# Setup and Data Loading

```{r setup}
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(lubridate)
library(gt)
library(janitor)
library(broom)

# Re-establish dplyr preferences after loading packages
if (requireNamespace("conflicted", quietly = TRUE)) {
  library(conflicted)
  conflicts_prefer(dplyr::select)
  conflicts_prefer(dplyr::filter)
  conflicts_prefer(dplyr::summarise)
}

# Custom theme for consistency
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.background = element_rect(fill = "#F9F6ED", color = NA),
      plot.background = element_rect(fill = "#F9F6ED", color = NA),
      panel.grid.major = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor = element_blank(),
      text = element_text(color = "#464A4F"),
      axis.text = element_text(color = "#464A4F"),
      axis.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold", size = base_size * 1.2),
      strip.text = element_text(face = "bold"),
      strip.background = element_rect(fill = alpha("#A5CAD2", 0.2)),
      legend.background = element_rect(fill = "#F9F6ED"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.spacing = unit(2, "lines")
    )
}

# Treatment colors
treatment_colors <- c(
  "Sorghum" = "#714C36",
  "Sorghum + Rye" = "#2B4B72", 
  "Corn" = "#1A3B57",
  "Soy" = "#B7B5B3"
)

scale_color_treatments <- function() scale_color_manual(values = treatment_colors)
scale_fill_treatments <- function() scale_fill_manual(values = treatment_colors)
```

# Data Import and Processing

```{r data-import}
#| message: false

# Import soil nitrogen data
# Nitrate and ammonia in ppm, not sure if moisture corrected
soil_data <- read_csv("data/soils/cleaned_soil_data.csv") %>%
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) %>%
  mutate(
    date = as.Date(date),
    total_inorganic_n = ammonia_ppm + nitrate_ppm
  ) 

# Import flux data 
# N2O flux in g N ha-1 day-1 
flux_data <- read_csv("data/cleaned_flux_data.csv") %>%
  mutate(date = as.Date(date)) 

# Import nitrogen leaching data
# Cumulative N loss in mg per day, same for nitrate and ammonium
leaching_data <- read_csv("data/2023_2024_SABR_tile_cumulative_n.csv") %>%
  clean_names() %>%
  mutate(
    date = as.Date(date),
    year = year(date)
  ) %>%
  filter(year %in% c(2023, 2024)) %>%  # Filter out bad years
  filter(!is.na(year))

cat("✓ Data imported successfully\n")
cat("Soil data:", nrow(soil_data), "observations\n")
cat("Flux data:", nrow(flux_data), "observations\n") 
cat("Leaching data:", nrow(leaching_data), "observations\n")
# Print the class of year columns in each data frame
cat("Soil data year column class:", class(soil_data$year), "\n")
cat("Flux data year column class:", class(flux_data$year), "\n")
cat("Leaching data year column class:", class(leaching_data$year), "\n")
```

# Helper Functions

```{r helper-functions}
# Function for pairwise t-tests within groups
get_pairwise_pvals <- function(data, variable, group_vars) {
  data %>%
    group_by(across(all_of(group_vars))) %>%
    group_modify(~ {
      treatments <- unique(.x$treatment)
      if (length(treatments) < 2) {
        return(tibble(p_value = NA_real_))
      }
      
      if (length(treatments) == 2) {
        t_result <- t.test(.x[[variable]][.x$treatment == treatments[1]], 
                          .x[[variable]][.x$treatment == treatments[2]])
        return(tibble(p_value = t_result$p.value))
      } else {
        pairwise_result <- pairwise.t.test(.x[[variable]], .x$treatment, 
                                          p.adjust.method = "none")
        return(tibble(p_value = min(pairwise_result$p.value, na.rm = TRUE)))
      }
    }) %>%
    ungroup()
}

# Function to add significance letters
add_significance_letters <- function(data, p_col) {
  data %>%
    mutate(
      significance_letter = case_when(
        is.na({{ p_col }}) ~ "",
        {{ p_col }} < 0.001 ~ "a",
        {{ p_col }} < 0.01 ~ "b", 
        {{ p_col }} < 0.05 ~ "c",
        TRUE ~ ""
      ),
      result_with_sig = if_else(
        significance_letter == "", 
        result,
        paste0(result, " ", significance_letter)
      )
    )
}

# Function to create ANOVA source tables
create_source_table <- function(model_results) {
  tibble(
    source = c("Date", "Treatment", "Date × Treatment"),
    p_value = summary(model_results)[[1]][c(2, 3, 6), "Pr(>F)"]
  ) %>%
    mutate(
      formatted_p = case_when(
        p_value < 0.001 ~ "< 0.001",
        p_value < 0.01 ~ sprintf("%.3f", p_value),
        TRUE ~ sprintf("%.2f", p_value)
      )
    )
}
```

# Table 2: Salt-extractable Inorganic Nitrogen

```{r table-2}
#| label: tbl-soil-nitrogen
#| tbl-cap: "Salt-extractable inorganic nitrogen"

# Calculate summary statistics
soil_summary <- soil_data %>%
  group_by(year, date, treatment) %>%
  summarise(
    across(c(ammonia_ppm, nitrate_ppm, total_inorganic_n), 
           list(mean = ~ mean(.x, na.rm = TRUE),
                se = ~ sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x))))),
    .groups = "drop"
  ) 

# Get p-values for pairwise comparisons
pvals_ammonium <- get_pairwise_pvals(soil_data, "ammonia_ppm", c("year", "date"))
pvals_nitrate <- get_pairwise_pvals(soil_data, "nitrate_ppm", c("year", "date"))  
pvals_total_n <- get_pairwise_pvals(soil_data, "total_inorganic_n", c("year", "date"))

# Combine with p-values and format
soil_table_data <- soil_summary %>%
  left_join(pvals_ammonium %>% rename(ammonia_p = p_value), by = c("year", "date")) %>%
  left_join(pvals_nitrate %>% rename(nitrate_p = p_value), by = c("year", "date")) %>%
  left_join(pvals_total_n %>% rename(total_n_p = p_value), by = c("year", "date")) %>%
  pivot_longer(
    cols = c(ammonia_ppm_mean, ammonia_ppm_se, nitrate_ppm_mean, nitrate_ppm_se, 
             total_inorganic_n_mean, total_inorganic_n_se),
    names_to = c("variable", "stat"),
    names_pattern = "(.*)_(mean|se)",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(
    result = sprintf("%.2f ± %.2f", mean, se),
    p_val = case_when(
      variable == "ammonia_ppm" ~ ammonia_p,
      variable == "nitrate_ppm" ~ nitrate_p,
      variable == "total_inorganic_n" ~ total_n_p
    ),
    variable = case_when(
      variable == "ammonia_ppm" ~ "ammonium",
      variable == "nitrate_ppm" ~ "nitrate", 
      variable == "total_inorganic_n" ~ "total_n"
    )
  ) %>%
  add_significance_letters(p_val) %>%
  select(year, date, treatment, variable, result_with_sig) %>%
  pivot_wider(names_from = variable, values_from = result_with_sig)

# Format dates and years for display
soil_table_formatted <- soil_table_data %>%
  mutate(
    year = as.character(year),
    date = format(as.Date(date), "%B %d")
  ) %>%
  arrange(year, as.Date(paste(year, date), format = "%Y %B %d"), treatment) %>%
  group_by(year) %>%
  mutate(
    year_display = if_else(row_number() == 1, year, ""),
    is_new_date = date != lag(date, default = "")
  ) %>%
  ungroup()

# Create ANOVA source of variation
aov_ammonium <- aov(ammonia_ppm ~ year * date * treatment, data = soil_data)
aov_nitrate <- aov(nitrate_ppm ~ year * date * treatment, data = soil_data)
aov_total_n <- aov(total_inorganic_n ~ year * date * treatment, data = soil_data)

source_variation <- tibble(
  source = c("Date", "Treatment", "Date × Treatment"),
  ammonium = summary(aov_ammonium)[[1]][c(2, 3, 6), "Pr(>F)"],
  nitrate = summary(aov_nitrate)[[1]][c(2, 3, 6), "Pr(>F)"],
  total_n = summary(aov_total_n)[[1]][c(2, 3, 6), "Pr(>F)"]
) %>%
  mutate(across(c(ammonium, nitrate, total_n), ~ case_when(
    .x < 0.001 ~ "< 0.001",
    .x < 0.01 ~ sprintf("%.3f", .x),
    TRUE ~ sprintf("%.2f", .x)
  ))) %>%
  mutate(
    year_display = "",
    date = "",
    treatment = source,
    .before = 1
  ) %>%
  select(-source)

# Combine data with source of variation
separator_row <- tibble(
  year_display = "Source of Variation", date = "", treatment = "",
  ammonium = "", nitrate = "", total_n = "", is_new_date = FALSE
)

table2_complete <- bind_rows(
  soil_table_formatted,
  separator_row,
  source_variation %>% mutate(is_new_date = FALSE)
)

# Create gt table
table2_gt <- table2_complete %>%
  select(-year) %>%
  select(year_display, date, treatment, ammonium, nitrate, total_n, is_new_date) %>%
  gt() %>%
  tab_header(title = md("**Table 2. Salt-extractable inorganic nitrogen**")) %>%
  cols_label(
    year_display = "Year",
    date = "Date", 
    treatment = "Treatment",
    ammonium = md("Ammonium<br>(mg kg<sup>-1</sup>)"),
    nitrate = md("Nitrate<br>(mg kg<sup>-1</sup>)"),
    total_n = md("Total Inorganic N<br>(mg kg<sup>-1</sup>)")
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", weight = px(1), color = "gray"),
    locations = cells_body(rows = is_new_date == TRUE)
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", weight = px(2)),
    locations = cells_body(rows = year_display == "Source of Variation")
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = year_display == "Source of Variation")
  ) %>%
  tab_style(
    style = cell_text(align = "center", weight = "bold"),
    locations = cells_body(columns = year_display, rows = year_display == "Source of Variation")
  ) %>%
  cols_hide(columns = is_new_date) %>%
  tab_footnote(
    footnote = "Letters indicate significance levels: a = p < 0.001, b = p < 0.01, c = p < 0.05",
    locations = cells_title()
  )

# Save table
gtsave(table2_gt, "tables/table2_soil_nitrogen.rtf")

table2_gt
```

# N2O Data Processing and Analysis

```{r n2o-processing}
# Process daily flux data with interpolation
daily_flux_summary <- flux_data %>%
  group_by(plot, rowvs_interrow, treatment, year, date) %>%
  summarise(mean_flux = mean(gnha_day_no_negative, na.rm = TRUE), .groups = "drop") %>%
  arrange(plot, rowvs_interrow, treatment, year, date)

# Interpolation function
interpolate_daily_flux <- function(data) {
  # Get unique plot combinations
  plot_combinations <- data %>%
    distinct(plot, rowvs_interrow, treatment, year)
  
  # Interpolate for each combination
  map_dfr(seq_len(nrow(plot_combinations)), function(i) {
    combo <- plot_combinations[i, ]
    
    subset_data <- data %>%
      filter(
        plot == combo$plot,
        rowvs_interrow == combo$rowvs_interrow,
        treatment == combo$treatment,
        year == combo$year
      ) %>%
      arrange(date)
    
    # Create full year sequence
    start_date <- as.Date(paste0(combo$year, "-01-01"))
    end_date <- as.Date(paste0(combo$year, "-12-31"))
    all_days <- seq(start_date, end_date, by = "day")
    
    # Interpolate if we have sufficient data
    if (nrow(subset_data) >= 2 && !all(is.na(subset_data$mean_flux))) {
      valid_data <- subset_data[!is.na(subset_data$mean_flux), ]
      
      if (nrow(valid_data) >= 2) {
        interpolated_flux <- approx(
          x = as.numeric(valid_data$date),
          y = valid_data$mean_flux,
          xout = as.numeric(all_days),
          rule = 1
        )$y
      } else {
        interpolated_flux <- rep(NA_real_, length(all_days))
      }
    } else {
      interpolated_flux <- rep(NA_real_, length(all_days))
    }
    
    tibble(
      plot = combo$plot,
      rowvs_interrow = combo$rowvs_interrow,
      treatment = combo$treatment,
      year = combo$year,
      date = all_days,
      interpolated_flux = interpolated_flux
    )
  })
}

# Apply interpolation
interpolated_flux_data <- interpolate_daily_flux(daily_flux_summary)

# Calculate cumulative N2O per plot
plot_cumulative_n2o <- interpolated_flux_data %>%
  group_by(plot, rowvs_interrow, treatment, year) %>%
  summarise(
    annual_cumulative_n2o = sum(interpolated_flux, na.rm = TRUE) / 1000, # Convert g to kg
    .groups = "drop"
  ) %>%
  filter(is.finite(annual_cumulative_n2o))

cat("✓ N2O interpolation and cumulative calculations complete\n")
cat("Plot-level cumulative N2O data:", nrow(plot_cumulative_n2o), "observations\n")
```

# Nitrogen Leaching Data Processing

```{r leaching-processing}
# Calculate annual cumulative nitrogen leaching per plot
# Convert mg per plot to kg N ha-1
plot_area_ha <- 0.1  # PLACEHOLDER: Assuming 0.1 hectare plots - VERIFY AND UPDATE!
annual_leaching <- leaching_data %>%
  group_by(plot, treatment, year) %>%
  summarise(
    annual_cumulative_leaching = max(cumulative_n_loss_mg, na.rm = TRUE) / 1000 / plot_area_ha, # mg to kg to kg/ha
    .groups = "drop"
  ) %>%
  filter(is.finite(annual_cumulative_leaching), annual_cumulative_leaching >= 0)

cat("✓ Nitrogen leaching calculations complete\n")
cat("Plot-level leaching data:", nrow(annual_leaching), "observations\n")
```

# Table 3: Annual Nitrogen Losses

```{r table-3}
#| label: tbl-nitrogen-losses
#| tbl-cap: "Annual nitrogen losses by treatment"

# Summarize N2O data by treatment and year
n2o_summary <- plot_cumulative_n2o %>%
  group_by(treatment, year) %>%
  summarise(
    n_plots = n(),
    mean_n2o = mean(annual_cumulative_n2o, na.rm = TRUE),
    sd_n2o = sd(annual_cumulative_n2o, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(n2o_result = sprintf("%.3f ± %.3f", mean_n2o, sd_n2o))

# Summarize leaching data by treatment and year
leaching_summary <- annual_leaching %>%
  group_by(treatment, year) %>%
  summarise(
    n_plots = n(),
    mean_leaching = mean(annual_cumulative_leaching, na.rm = TRUE),
    sd_leaching = sd(annual_cumulative_leaching, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(leaching_result = sprintf("%.3f ± %.3f", mean_leaching, sd_leaching))

# Calculate 2-year totals
n2o_two_year <- plot_cumulative_n2o %>%
  group_by(plot, rowvs_interrow, treatment) %>%
  summarise(total_n2o = sum(annual_cumulative_n2o, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  summarise(
    mean_n2o = mean(total_n2o, na.rm = TRUE),
    sd_n2o = sd(total_n2o, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    year = "2023-2024",
    n2o_result = sprintf("%.3f ± %.3f", mean_n2o, sd_n2o)
  )

leaching_two_year <- annual_leaching %>%
  group_by(plot, treatment) %>%
  summarise(total_leaching = sum(annual_cumulative_leaching, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  summarise(
    mean_leaching = mean(total_leaching, na.rm = TRUE),
    sd_leaching = sd(total_leaching, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    year = "2023-2024", 
    leaching_result = sprintf("%.3f ± %.3f", mean_leaching, sd_leaching)
  )
compare_df_cols(n2o_two_year, leaching_two_year, n2o_summary, leaching_summary)
# Combine annual and 2-year data
combined_nitrogen_data <- bind_rows(
  n2o_summary %>% mutate(year = as.character(year)) %>% select(year, treatment, n2o_result),
  n2o_two_year %>% select(year, treatment, n2o_result)
) %>%
  full_join(
    bind_rows(
      leaching_summary %>% mutate(year = as.character(year)) %>% select(year, treatment, leaching_result),
      leaching_two_year %>% select(year, treatment, leaching_result)
    ),
    by = c("year", "treatment")
  ) %>%
  mutate(
    year_order = case_when(
      year == "2023" ~ 1,
      year == "2024" ~ 2,
      year == "2023-2024" ~ 3
    ),
    total_environmental_cost = "TBD" # Placeholder for future calculation
  ) %>%
  arrange(year_order, treatment)

# Source of variation analysis
# N2O ANOVA
combined_n2o_data <- plot_cumulative_n2o %>% mutate(year_factor = as.factor(year))
aov_n2o <- aov(annual_cumulative_n2o ~ year_factor * treatment, data = combined_n2o_data)

# Leaching ANOVA  
combined_leaching_data <- annual_leaching %>% mutate(year_factor = as.factor(year))
aov_leaching <- aov(annual_cumulative_leaching ~ year_factor * treatment, data = combined_leaching_data)

# Source of variation table
source_variation_n <- tibble(
  source = c("Year", "Treatment", "Year × Treatment"),
  n2o_p = summary(aov_n2o)[[1]][1:3, "Pr(>F)"],
  leaching_p = summary(aov_leaching)[[1]][1:3, "Pr(>F)"]
) %>%
  mutate(
    across(c(n2o_p, leaching_p), ~ case_when(
      .x < 0.001 ~ "< 0.001",
      .x < 0.01 ~ sprintf("%.3f", .x),
      TRUE ~ sprintf("%.2f", .x)
    )),
    year = "",
    year_order = 4,
    treatment = source,
    n2o_result = n2o_p,
    leaching_result = leaching_p,
    total_environmental_cost = "TBD"
  ) %>%
  select(-source, -n2o_p, -leaching_p)

# Create complete table
separator_n <- tibble(
  year = "Source of Variation", year_order = 3.5, treatment = "",
  n2o_result = "", leaching_result = "", total_environmental_cost = ""
)

table3_complete <- bind_rows(
  combined_nitrogen_data,
  separator_n,
  source_variation_n
) %>%
  arrange(year_order, treatment) %>%
  mutate(
    year_display = case_when(
      year == "Source of Variation" ~ year,
      year != lag(year, default = "") ~ as.character(year),
      TRUE ~ ""
    ),
    is_new_section = year != lag(year, default = "") & year != "Source of Variation"
  )

# Create gt table
table3_gt <- table3_complete %>%
  select(
    Year = year_display,
    Treatment = treatment,
    `N2O Loss` = n2o_result,
    `N Leaching` = leaching_result, 
    `Total Environmental Cost` = total_environmental_cost,
    is_new_section
  ) %>%
  gt() %>%
  tab_header(title = md("**Table 3. Annual nitrogen losses**")) %>%
  cols_label(
    `N2O Loss` = md("N<sub>2</sub>O Loss<br>(kg N ha<sup>-1</sup>)"),
    `N Leaching` = md("N Leaching<br>(kg N ha<sup>-1</sup>)"),
    `Total Environmental Cost` = md("Total Environmental Cost<br>(kg N equivalent)")
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", weight = px(1), color = "gray"),
    locations = cells_body(rows = is_new_section == TRUE)
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", weight = px(2)),
    locations = cells_body(rows = Year == "Source of Variation")
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = Year == "Source of Variation")  
  ) %>%
  cols_hide(columns = is_new_section) %>%
  tab_footnote(
    footnote = "Values are mean ± standard deviation. Letters indicate significance levels: a = p < 0.001, b = p < 0.01, c = p < 0.05",
    locations = cells_title()
  )

# Save table
gtsave(table3_gt, "tables/table3_nitrogen_losses.rtf")

table3_gt
```

# Figure 1: Daily N2O Flux Time Series

```{r figure-1}
#| label: fig-flux-timeseries
#| fig-cap: "Daily N2O flux by treatment over time"
#| fig-width: 12
#| fig-height: 8

# Aggregate interpolated data for plotting
daily_flux_aggregated <- interpolated_flux_data %>%
  group_by(treatment, year, date) %>%
  summarise(
    n_plots = sum(!is.na(interpolated_flux)),
    mean_flux = mean(interpolated_flux, na.rm = TRUE),
    se_flux = if_else(
      n_plots > 1, 
      sd(interpolated_flux, na.rm = TRUE) / sqrt(n_plots), 
      NA_real_
    ),
    .groups = "drop"
  )

# Create time series plot
ggplot(daily_flux_aggregated, aes(x = date, y = mean_flux, color = treatment)) +
  geom_ribbon(
    aes(ymin = mean_flux - se_flux, ymax = mean_flux + se_flux, fill = treatment),
    alpha = 0.2, color = NA
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 1, alpha = 0.7) +
  facet_grid(. ~ year, scales = "free_x", space = "free_x") +
  scale_color_treatments() +
  scale_fill_treatments() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    x = "Month",
    y = expression("Daily N"[2]*"O Flux (g N ha"^"-1"*")"),
    color = "Treatment",
    fill = "Treatment"
  ) +
  theme_publication() +
  theme(legend.position = "bottom")

ggsave("figures/figure1_flux_timeseries.png", width = 12, height = 8, dpi = 300)
```

# Summary

This document provides clean, efficient analysis following tidyverse principles:

-   **Consistent data processing** using `group_by()` and `summarise()`
-   **Functional programming** with custom helper functions
-   **Clear separation** between data processing and visualization
-   **Reproducible workflow** with proper chunk organization
-   **Professional styling** consistent across all outputs

Each table and figure is generated in its own chunk for modularity and clarity.