---
title: "Reproducible Research Companion: Sorghum & Cover Crops"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
    toc-depth: 3
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  echo: true
---

# Introduction

This document serves as a reproducible companion to the manuscript titled **"Integrating Winter Cover Crops With Biomass Sorghum Can Reduce Nitrogen Losses"**. Here, we present the complete analytical workflow that generated every figure and table in the paper, allowing readers to understand and reproduce our findings.

The study compared biomass sorghum systems with and without winter cereal rye cover crops across two years (2023-2024), measuring:

-   Weather conditions and soil microclimate
-   Soil inorganic nitrogen concentrations
-   Nitrate leaching via tile drainage
-   Nitrous oxide (N₂O) emissions
-   Biomass yields
-   Environmental damage costs

All code is shown by default to promote transparency and reproducibility.

# Global Setup

This section loads all required libraries and defines shared functions used throughout the analysis.

```{r setup}
#| message: false
#| warning: false

# Load required libraries
library(tidyverse)
library(readxl)
library(patchwork)
library(gt)
library(rstatix)
library(emmeans)
library(multcomp)
library(multcompView)
library(lubridate)
library(janitor)
library(glue)

# Define publication theme
theme_publication <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      # Backgrounds
      panel.background = element_rect(fill = "#ffffffff", color = NA),
      plot.background = element_rect(fill = "#ffffffff", color = NA),
      strip.background = element_rect(fill = alpha("#A5CAD2", 0.2)),
      
      # Grid
      panel.grid.major = element_line(color = alpha("#B7B5B3", 0.2)),
      panel.grid.minor = element_blank(),
      panel.spacing = unit(3, "lines"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1),

      # Text & Axis
      text = element_text(color = "#000000ff"),
      axis.text = element_text(color = "#000000ff", size = base_size * 0.8),
      axis.title = element_text(size = base_size),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 8)),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line = element_line(color = "black", linewidth = 0.5),
      axis.ticks = element_line(color = "black", linewidth = 0.5),
      
      # Legend
      legend.position = "top",
      legend.justification = "center",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      legend.margin = margin(b = 1, unit = "pt"),
      legend.box.margin = margin(b = 1, unit = "pt"),
      legend.text = element_text(size = base_size * 0.9),
      legend.title = element_text(size = base_size),
      legend.key.width = unit(0.8, "cm"),
      legend.key.height = unit(0.3, "cm")
    )
}

# Treatment color palette
treatment_colors <- c(
  "Sorghum" = "#D8D97AFF",
  "Sorghum + Rye" = "#95C36EFF", 
  "Sorghum + WCC" = "#95C36EFF",
  "Corn" = "#74C8C3FF",
  "Soy" = "#0A2E57FF"
)

scale_color_treatments <- function() scale_color_manual(values = treatment_colors)
scale_fill_treatments <- function() scale_fill_manual(values = treatment_colors)

# Shared panel-label styling for figure consistency (Figs 1-5)
panel_label_size <- 3.8
panel_label_face <- "bold"
panel_label_hjust <- -0.08
panel_label_vjust <- 1.15

add_panel_label <- function(label) {
  annotate(
    "text",
    x = -Inf, y = Inf,
    label = label,
    hjust = panel_label_hjust,
    vjust = panel_label_vjust,
    fontface = panel_label_face,
    size = panel_label_size
  )
}
```

# Weather Conditions (Figure 1)

**Scientific Context:** Weather conditions varied substantially between study years. The year 2023 was characterized by below-average precipitation, while 2024 experienced conditions closer to the historical mean. These differences influenced soil moisture dynamics, crop growth, and nitrogen cycling processes.

```{r figure1-weather}
#| fig-width: 10
#| fig-height: 4.5
# Load weather data
rain <- read_xlsx("data/rain/historical_rain.xlsx")
gdd <- read_xlsx("data/rain/historical_rain.xlsx")

# Process precipitation data
rain <- rain %>%
  select(day, doy, precipmm) %>%
  mutate(
    date = as.Date(day),
    year = year(date)
  ) %>%
  group_by(year) %>%
  arrange(date) %>%
  mutate(cumulative_precip = cumsum(precipmm)) %>%
  mutate(doy_date = as.Date(doy - 1, origin = "2000-01-01"))

# Process growing degree days
gdd <- gdd %>%
  select(day, doy, gdd_50_86) %>%
  mutate(
    date = as.Date(day),
    year = year(date)
  ) %>%
  group_by(year) %>%
  arrange(date) %>%
  mutate(cumulative_gdd = cumsum(gdd_50_86)) %>%
  mutate(doy_date = as.Date(doy - 1, origin = "2000-01-01"))

# Historical envelopes (excluding 2023 and 2024)
hist_env_rain <- rain |>
  filter(!(year %in% c(2023, 2024))) |>
  group_by(doy_date) |>
  summarize(
    min_cum = min(cumulative_precip, na.rm = TRUE),
    max_cum = max(cumulative_precip, na.rm = TRUE),
    .groups = "drop"
  )

hist_env_gdd <- gdd |>
  filter(!(year %in% c(2023, 2024))) |>
  group_by(doy_date) |>
  summarize(
    min_cum = min(cumulative_gdd, na.rm = TRUE),
    max_cum = max(cumulative_gdd, na.rm = TRUE),
    .groups = "drop"
  )

# 34-year historical mean lines
mean_cum_rain <- rain |>
  filter(!(year %in% c(2023, 2024))) |>
  group_by(doy_date) |>
  summarize(cumulative_precip = mean(cumulative_precip, na.rm = TRUE), .groups = "drop") |>
  mutate(year = "Mean") |>
  filter(doy_date != "2000-12-31")

mean_cum_gdd <- gdd |>
  filter(!(year %in% c(2023, 2024))) |>
  group_by(doy_date) |>
  summarize(cumulative_gdd = mean(cumulative_gdd, na.rm = TRUE), .groups = "drop") |>
  mutate(year = "Mean") |>
  filter(doy_date != "2000-12-31")

# Combine highlighted years with historical mean
rain_highlight <- rain |>
  filter(year %in% c(2023, 2024)) |>
  mutate(year = as.character(year)) |>
  bind_rows(mean_cum_rain)

gdd_highlight <- gdd |>
  filter(year %in% c(2023, 2024)) |>
  mutate(year = as.character(year)) |>
  bind_rows(mean_cum_gdd)

rain_plot <- ggplot() +
  geom_ribbon(
    data = hist_env_rain,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  geom_line(
    data = rain_highlight,
    aes(x = doy_date, y = cumulative_precip, color = year, linetype = year),
    linewidth = 0.7
  ) +
  scale_color_manual(
    name = "Year",
    values = c("2023" = "#E41A1C", "2024" = "#377EB8", "Mean" = "black"),
    labels = c("2023" = "2023", "2024" = "2024", "Mean" = "34-year Mean")
  ) +
  scale_linetype_manual(
    name = "Year",
    values = c("2023" = "solid", "2024" = "dashed", "Mean" = "dotdash"),
    labels = c("2023" = "2023", "2024" = "2024", "Mean" = "34-year Mean")
  ) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  labs(x = "", y = "Cumulative Precipitation (mm)") +
  theme_publication(base_size = 10) +
  add_panel_label("a)")

gdd_plot <- ggplot() +
  geom_ribbon(
    data = hist_env_gdd,
    aes(x = doy_date, ymin = min_cum, ymax = max_cum),
    fill = "gray80", alpha = 0.6
  ) +
  geom_line(
    data = gdd_highlight,
    aes(x = doy_date, y = cumulative_gdd, color = year, linetype = year),
    linewidth = 0.7
  ) +
  scale_color_manual(
    name = "Year",
    values = c("2023" = "#E41A1C", "2024" = "#377EB8", "Mean" = "black"),
    labels = c("2023" = "2023", "2024" = "2024", "Mean" = "34-year Mean")
  ) +
  scale_linetype_manual(
    name = "Year",
    values = c("2023" = "solid", "2024" = "dashed", "Mean" = "dotdash"),
    labels = c("2023" = "2023", "2024" = "2024", "Mean" = "34-year Mean")
  ) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  labs(x = "", y = "Cumulative Growing Degree Days (GDD)") +
  theme_publication(base_size = 10) +
  add_panel_label("b)")

figure1 <- (rain_plot | gdd_plot) +
  plot_layout(guides = "collect", widths = c(1.05, 1.05)) &
  theme_publication(base_size = 10) &
  theme(
    legend.position = "top",
    legend.text = element_text(size = 9),
    legend.key.width = unit(1.1, "cm"),
    plot.margin = margin(t = 4, r = 6, b = 3, l = 6, unit = "pt")
  ) &
  guides(
    color = guide_legend(nrow = 1, byrow = TRUE),
    linetype = guide_legend(nrow = 1, byrow = TRUE)
  )

figure1
```

# Soil Microclimate (Figure 2)

**Scientific Context:** We monitored soil temperature and volumetric water content at 5 cm depth to evaluate whether the winter cover crop altered the biophysical environment experienced by soil microbes and plants. Differences in soil microclimate could influence decomposition rates, nutrient mineralization, and crop performance.

```{r figure2-microclimate}
# Load flux data (contains soil sensors)
flux_data <- read_csv("data/cleaned_flux_data.csv") |>
  mutate(
    date = as.Date(date),
    year = year(date)
  )

raw_soil <- flux_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    Treatment = factor(Treatment, levels = c("Sorghum", "Sorghum + WCC"))
  )

# Daily treatment means for time series and difference panels
moist_temp <- raw_soil |>
  group_by(Treatment, year, date) |>
  summarise(
    mean_soil_moisture = mean(swc_2_mean, na.rm = TRUE),
    mean_soil_temp = mean(ts_2_mean, na.rm = TRUE),
    .groups = "drop"
  )

daily_diff <- moist_temp |>
  mutate(trt_key = if_else(Treatment == "Sorghum + WCC", "WCC", "Sorghum")) |>
  select(year, date, trt_key, mean_soil_temp, mean_soil_moisture) |>
  pivot_wider(names_from = trt_key, values_from = c(mean_soil_temp, mean_soil_moisture)) |>
  mutate(
    temp_diff = mean_soil_temp_WCC - mean_soil_temp_Sorghum,
    moist_diff = mean_soil_moisture_WCC - mean_soil_moisture_Sorghum
  )

# Date-status bars: significant difference, no difference, or no data
build_status_bars <- function(df, value_col) {
  df |>
    group_by(year, date) |>
    summarise(
      n_sorghum = sum(Treatment == "Sorghum" & !is.na(.data[[value_col]])),
      n_wcc = sum(Treatment == "Sorghum + WCC" & !is.na(.data[[value_col]])),
      mean_sorghum = mean(.data[[value_col]][Treatment == "Sorghum"], na.rm = TRUE),
      mean_wcc = mean(.data[[value_col]][Treatment == "Sorghum + WCC"], na.rm = TRUE),
      p_value = {
        sorghum_vals <- .data[[value_col]][Treatment == "Sorghum"]
        wcc_vals <- .data[[value_col]][Treatment == "Sorghum + WCC"]
        sorghum_vals <- sorghum_vals[!is.na(sorghum_vals)]
        wcc_vals <- wcc_vals[!is.na(wcc_vals)]
        if (length(sorghum_vals) >= 2 && length(wcc_vals) >= 2) {
          tryCatch(t.test(sorghum_vals, wcc_vals)$p.value, error = function(e) NA_real_)
        } else {
          NA_real_
        }
      },
      .groups = "drop"
    ) |>
    mutate(
      has_testable = n_sorghum >= 2 & n_wcc >= 2,
      significant = has_testable & !is.na(p_value) & p_value < 0.05,
      status = case_when(
        significant ~ "Significant",
        has_testable ~ "No difference",
        TRUE ~ "No data"
      ),
      greater = case_when(
        significant & mean_wcc > mean_sorghum ~ "Sorghum + WCC",
        significant & mean_sorghum >= mean_wcc ~ "Sorghum",
        TRUE ~ NA_character_
      ),
      x = date - 2,
      xend = date + 2
    )
}

temp_status <- build_status_bars(raw_soil, "ts_2_mean")
moist_status <- build_status_bars(raw_soil, "swc_2_mean")

status_fill <- c(
  "Significant" = "black",
  "No difference" = "grey65",
  "No data" = "grey90"
)

moist_temp_2023 <- filter(moist_temp, year == 2023)
moist_temp_2024 <- filter(moist_temp, year == 2024)

global_moisture_limits <- range(moist_temp$mean_soil_moisture, na.rm = TRUE)
global_moisture_limits[2] <- global_moisture_limits[2] * 1.14

global_temp_limits <- range(moist_temp$mean_soil_temp, na.rm = TRUE)
global_temp_limits[2] <- max(40, global_temp_limits[2] * 1.08)

moisture_breaks <- pretty(global_moisture_limits, n = 5)
temp_breaks <- pretty(global_temp_limits, n = 5)

temp_bar_y <- global_temp_limits[2] - diff(global_temp_limits) * 0.035
temp_bar_half <- diff(global_temp_limits) * 0.01
moist_bar_y <- global_moisture_limits[2] - diff(global_moisture_limits) * 0.035
moist_bar_half <- diff(global_moisture_limits) * 0.01

# Shared one-sample t-test helper for difference panels
one_sample_vs_zero <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, mu = 0)$p.value, error = function(e) NA_real_)
}

temp_diff_stats <- daily_diff |>
  group_by(year) |>
  summarise(
    n = sum(!is.na(temp_diff)),
    mean_diff = mean(temp_diff, na.rm = TRUE),
    p_value = one_sample_vs_zero(temp_diff),
    significant = !is.na(p_value) & p_value < 0.05,
    letter = if_else(significant, "a", NA_character_),
    .groups = "drop"
  )

moist_diff_stats <- daily_diff |>
  group_by(year) |>
  summarise(
    n = sum(!is.na(moist_diff)),
    mean_diff = mean(moist_diff, na.rm = TRUE),
    p_value = one_sample_vs_zero(moist_diff),
    significant = !is.na(p_value) & p_value < 0.05,
    letter = if_else(significant, "a", NA_character_),
    .groups = "drop"
  )

# Panel a) 2023 temperature
p3 <- ggplot(moist_temp_2023, aes(x = date, y = mean_soil_temp, color = Treatment, linetype = Treatment, shape = Treatment)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  geom_rect(
    data = temp_status |> filter(year == 2023),
    aes(
      xmin = x, xmax = xend,
      ymin = temp_bar_y - temp_bar_half,
      ymax = temp_bar_y + temp_bar_half,
      fill = status
    ),
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  geom_segment(
    data = temp_status |> filter(year == 2023, significant),
    aes(x = x, xend = xend, y = temp_bar_y, yend = temp_bar_y, color = greater),
    linewidth = 3,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = status_fill) +
  scale_color_manual(values = treatment_colors) +
  scale_linetype_manual(values = c("Sorghum" = "dashed", "Sorghum + WCC" = "solid")) +
  scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + WCC" = 15)) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = c(as.Date("2023-01-01"), as.Date("2023-12-31"))) +
  scale_y_continuous(limits = global_temp_limits, breaks = temp_breaks, expand = expansion(mult = c(0.04, 0.04))) +
  labs(x = "", y = "Soil Temperature (°C)", color = "Treatment", linetype = "Treatment", shape = "Treatment") +
  theme_publication(base_size = 10) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    legend.position = "none"
  ) +
    add_panel_label("a)")

# Panel b) 2024 temperature
p4 <- ggplot(moist_temp_2024, aes(x = date, y = mean_soil_temp, color = Treatment, linetype = Treatment, shape = Treatment)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  geom_rect(
    data = temp_status |> filter(year == 2024),
    aes(
      xmin = x, xmax = xend,
      ymin = temp_bar_y - temp_bar_half,
      ymax = temp_bar_y + temp_bar_half,
      fill = status
    ),
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  geom_segment(
    data = temp_status |> filter(year == 2024, significant),
    aes(x = x, xend = xend, y = temp_bar_y, yend = temp_bar_y, color = greater),
    linewidth = 3,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = status_fill) +
  scale_color_manual(values = treatment_colors) +
  scale_linetype_manual(values = c("Sorghum" = "dashed", "Sorghum + WCC" = "solid")) +
  scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + WCC" = 15)) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = c(as.Date("2024-01-01"), as.Date("2024-12-31"))) +
  scale_y_continuous(limits = global_temp_limits, breaks = temp_breaks, expand = expansion(mult = c(0.04, 0.04))) +
  labs(x = "", y = "", color = "Treatment", linetype = "Treatment", shape = "Treatment") +
  theme_publication(base_size = 10) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    legend.position = "none"
  ) +
    add_panel_label("b)")

# Panel c) 2023 moisture
p1 <- ggplot(moist_temp_2023, aes(x = date, y = mean_soil_moisture, color = Treatment, linetype = Treatment, shape = Treatment)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  geom_rect(
    data = moist_status |> filter(year == 2023),
    aes(
      xmin = x, xmax = xend,
      ymin = moist_bar_y - moist_bar_half,
      ymax = moist_bar_y + moist_bar_half,
      fill = status
    ),
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  geom_segment(
    data = moist_status |> filter(year == 2023, significant),
    aes(x = x, xend = xend, y = moist_bar_y, yend = moist_bar_y, color = greater),
    linewidth = 3,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = status_fill) +
  scale_color_manual(values = treatment_colors) +
  scale_linetype_manual(values = c("Sorghum" = "dashed", "Sorghum + WCC" = "solid")) +
  scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + WCC" = 15)) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = c(as.Date("2023-01-01"), as.Date("2023-12-31"))) +
  scale_y_continuous(limits = global_moisture_limits, breaks = moisture_breaks, expand = expansion(mult = c(0.04, 0.04))) +
  labs(
    x = "Month",
    y = expression("Volumetric Soil Moisture (cm"^3 * " " * "cm"^-3 * ")"),
    color = "Treatment",
    linetype = "Treatment",
    shape = "Treatment"
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "none") +
    add_panel_label("c)")

# Panel d) 2024 moisture
p2 <- ggplot(moist_temp_2024, aes(x = date, y = mean_soil_moisture, color = Treatment, linetype = Treatment, shape = Treatment)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  geom_rect(
    data = moist_status |> filter(year == 2024),
    aes(
      xmin = x, xmax = xend,
      ymin = moist_bar_y - moist_bar_half,
      ymax = moist_bar_y + moist_bar_half,
      fill = status
    ),
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  geom_segment(
    data = moist_status |> filter(year == 2024, significant),
    aes(x = x, xend = xend, y = moist_bar_y, yend = moist_bar_y, color = greater),
    linewidth = 3,
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = status_fill) +
  scale_color_manual(values = treatment_colors) +
  scale_linetype_manual(values = c("Sorghum" = "dashed", "Sorghum + WCC" = "solid")) +
  scale_shape_manual(values = c("Sorghum" = 16, "Sorghum + WCC" = 15)) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = c(as.Date("2024-01-01"), as.Date("2024-12-31"))) +
  scale_y_continuous(limits = global_moisture_limits, breaks = moisture_breaks, expand = expansion(mult = c(0.04, 0.04))) +
  labs(x = "Month", y = "", color = "Treatment", linetype = "Treatment", shape = "Treatment") +
  theme_publication(base_size = 10) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    legend.position = "none"
  ) +
    add_panel_label("d)")

# Temperature difference panel (e)
temp_span <- diff(range(daily_diff$temp_diff, na.rm = TRUE))
if (!is.finite(temp_span) || temp_span == 0) temp_span <- 1
temp_anno <- daily_diff |>
  group_by(year) |>
  summarise(y = max(temp_diff, na.rm = TRUE), .groups = "drop") |>
  left_join(temp_diff_stats, by = "year") |>
  filter(significant) |>
  mutate(y = y + temp_span * 0.08)

p5 <- ggplot(daily_diff, aes(x = factor(year), y = temp_diff)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
  geom_jitter(
    aes(color = factor(year)),
    width = 0.12,
    height = 0,
    alpha = 0.7,
    size = 1.4,
    show.legend = FALSE
  ) +
  geom_text(
    data = temp_anno,
    aes(x = factor(year), y = y, label = letter),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.8
  ) +
  scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
  labs(
    x = "Year",
    y = expression("Daily Temperature Difference (°C, WCC - Sorghum)")
  ) +
  theme_publication(base_size = 10) +
    add_panel_label("e)")

# Moisture difference panel (f)
moist_span <- diff(range(daily_diff$moist_diff, na.rm = TRUE))
if (!is.finite(moist_span) || moist_span == 0) moist_span <- 0.01
moist_anno <- daily_diff |>
  group_by(year) |>
  summarise(y = max(moist_diff, na.rm = TRUE), .groups = "drop") |>
  left_join(moist_diff_stats, by = "year") |>
  filter(significant) |>
  mutate(y = y + moist_span * 0.08)

p6 <- ggplot(daily_diff, aes(x = factor(year), y = moist_diff)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
  geom_jitter(
    aes(color = factor(year)),
    width = 0.12,
    height = 0,
    alpha = 0.7,
    size = 1.4,
    show.legend = FALSE
  ) +
  geom_text(
    data = moist_anno,
    aes(x = factor(year), y = y, label = letter),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.8
  ) +
  scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
  labs(
    x = "Year",
    y = expression("Daily Moisture Difference (cm"^3 * " " * "cm"^-3 * ", WCC - Sorghum)")
  ) +
  theme_publication(base_size = 10) +
    add_panel_label("f)")

figure2 <- (p3 | p4) / (p1 | p2) / (p5 | p6) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1, 0.95)) &
  theme(legend.position = "top") &
  guides(
    color = guide_legend(nrow = 1, byrow = TRUE),
    linetype = guide_legend(nrow = 1, byrow = TRUE),
    shape = guide_legend(nrow = 1, byrow = TRUE)
  )

figure2 <- figure2 +
  patchwork::plot_annotation(
    caption = "Status bars above time-series panels: black = significant treatment difference; gray = no difference; light gray = no data. Colored overlays indicate which treatment had the greater daily mean on significant dates."
  )

temp_diff_stats
moist_diff_stats
figure2
```

# Soil Inorganic Nitrogen (Table 2)

**Scientific Context:** We analyzed soil ammonium (NH₄⁺-N) and nitrate (NO₃⁻-N) concentrations to determine whether the winter cover crop immobilized soil nitrogen, potentially reducing availability for the subsequent sorghum crop. Samples were collected from 0-30 cm depth at multiple time points during the growing season.

```{r table2-soil-nitrogen}
# Load soil data
soil_data <- read_csv("data/soils/cleaned_soil_data.csv", show_col_types = FALSE)

# Prepare data for analysis
df_analysis <- soil_data %>%
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) %>%
  mutate(
    total_n = ammonia_mg_per_kg + nitrate_mg_per_kg,
    nitrate_prop = nitrate_mg_per_kg / total_n * 100,
    year = as.character(year)
  )

# Function to run Kruskal-Wallis & Get Letters per Year
get_year_stats <- function(data, yr) {
  df_yr <- data %>% filter(year == yr)
  
  # Global P-value
  kw <- kruskal_test(df_yr, total_n ~ treatment)
  p_val <- kw$p
  
  # Letters (Dunn's Test)
  if(p_val < 0.05) {
    dunn <- dunn_test(df_yr, total_n ~ treatment, p.adjust.method = "holm")
    dunn <- dunn %>% mutate(pair = paste(group1, group2, sep="-"))
    p_vec <- setNames(dunn$p.adj, dunn$pair)
    lets <- multcompLetters(p_vec)$Letters
  } else {
    lets <- setNames(rep("", length(unique(df_yr$treatment))), unique(df_yr$treatment))
  }
  
  # Calculate Summary Stats
  stats <- df_yr %>%
    group_by(treatment) %>%
    summarise(
      n_samples = n(),
      mean = mean(total_n, na.rm=TRUE),
      sd = sd(total_n, na.rm=TRUE),
      min = min(total_n, na.rm=TRUE),
      median = median(total_n, na.rm=TRUE),
      max = max(total_n, na.rm=TRUE)
    ) %>%
    mutate(
      letter = lets[treatment],
      year = yr,
      p_global = p_val
    )
  
  return(stats)
}

# Run for both years
stats_2023 <- get_year_stats(df_analysis, "2023")
stats_2024 <- get_year_stats(df_analysis, "2024")

# Format P-values
p_val_2023 <- unique(stats_2023$p_global)
year_label_23 <- ifelse(p_val_2023 < 0.001, 
                        "2023\n(P < 0.001)", 
                        paste0("2023\n(P = ", sprintf("%.3f", p_val_2023), ")"))

p_val_2024 <- unique(stats_2024$p_global)
year_label_24 <- ifelse(p_val_2024 < 0.001, 
                        "2024\n(P < 0.001)", 
                        paste0("2024\n(P = ", sprintf("%.3f", p_val_2024), ")"))

# Prepare table data
gt_ready <- bind_rows(stats_2023, stats_2024) %>%
  mutate(
    `Mean ± Standard Deviation` = glue("{sprintf('%.2f', mean)} ± {sprintf('%.2f', sd)}{letter}"),
    Year = case_when(
      year == "2023" ~ year_label_23,
      year == "2024" ~ year_label_24
    )
  ) %>%
  select(
    Year, 
    Treatment = treatment, 
    `Number of Soil Samples` = n_samples, 
    `Mean ± Standard Deviation`, 
    Minimum = min, 
    Median = median, 
    Maximum = max
  )

# Generate table
table2 <- gt_ready %>%
  gt() %>%
  cols_label(
    Year = "Year",
    Treatment = "Treatment",
    `Number of Soil Samples` = "N",
    `Mean ± Standard Deviation` = "Mean ± SD",
    Minimum = "Min",
    Median = "Med",
    Maximum = "Max"
  ) %>%
  fmt_number(columns = c(Minimum, Median, Maximum), decimals = 2) %>%
  cols_align(align = "center", columns = everything()) %>%
  cols_align(align = "left", columns = c(Treatment)) %>%
  fmt_markdown(columns = Year) %>%
  tab_header(
    title = md("**Table 2. Soil salt-extractable inorganic nitrogen (ammonium + nitrate)**")
  )

# Add footnote
nitrate_range <- range(df_analysis$nitrate_prop, na.rm = TRUE)
footnote_text <- glue("Nitrate ranged from {sprintf('%.1f', nitrate_range[1])} to {sprintf('%.1f', nitrate_range[2])}% of inorganic N. Letters indicate significance groups within year (Kruskal-Wallis with Dunn's test, α = 0.05).")

table2 <- table2 %>%
  tab_footnote(footnote = footnote_text)

table2
```

# Nitrate Leaching (Figure 3)

**Scientific Context:** Nitrate leaching via subsurface tile drainage represents a primary nitrogen loss pathway in agricultural systems. This figure shows daily and cumulative nitrogen losses through tile drainage, demonstrating the key finding that the winter cover crop substantially reduced nitrogen leaching compared to the no-cover control.

```{r figure3-leaching}
# 1. Load and process leaching data
plot_area_ft2 <- 120 * 160
plot_area_ha <- plot_area_ft2 / 107639.1041671
fert_events <- tibble(
  year = c(2023, 2024),
  fertilizer_date = as.Date(c("2023-05-05", "2024-07-17"))
)

leaching_data <- read_csv("data/clean_leaching.csv") |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(date),
    year = year(date)
  ) |>
  filter(year %in% c(2023, 2024))

# 2. Daily and cumulative summaries (Sorghum vs Sorghum + WCC)
daily_leaching_summary <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    daily_leached_n_kgha = (total_n_loss_mg / 1e6) / plot_area_ha
  ) |>
  group_by(Treatment, year, date) |>
  summarise(
    mean_leached = mean(daily_leached_n_kgha, na.rm = TRUE),
    se_leached = sd(daily_leached_n_kgha, na.rm = TRUE) / sqrt(sum(!is.na(daily_leached_n_kgha))),
    .groups = "drop"
  )

cum_leaching_summary <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    cumulative_leached_n_kgha = (cumulative_n_loss_mg / 1e6) / plot_area_ha
  ) |>
  group_by(Treatment, year, date) |>
  summarise(
    cum_leached_mean = mean(cumulative_leached_n_kgha, na.rm = TRUE),
    cum_leached_sd = sd(cumulative_leached_n_kgha, na.rm = TRUE),
    n = n(),
    cum_leached_se = cum_leached_sd / sqrt(n),
    .groups = "drop"
  )

# 3. Date-specific treatment tests (asterisks)
daily_leaching_sig <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    daily_leached_n_kgha = (total_n_loss_mg / 1e6) / plot_area_ha
  ) |>
  group_by(year, date) |>
  summarise(
    p_value = {
      sorghum_vals <- daily_leached_n_kgha[Treatment == "Sorghum"]
      wcc_vals <- daily_leached_n_kgha[Treatment == "Sorghum + WCC"]
      sorghum_vals <- sorghum_vals[!is.na(sorghum_vals)]
      wcc_vals <- wcc_vals[!is.na(wcc_vals)]
      if (length(sorghum_vals) >= 3 && length(wcc_vals) >= 3) {
        tryCatch(wilcox.test(sorghum_vals, wcc_vals)$p.value, error = function(e) NA_real_)
      } else {
        NA_real_
      }
    },
    .groups = "drop"
  ) |>
  mutate(significant = !is.na(p_value) & p_value < 0.05)

leaching_stars <- daily_leaching_summary |>
  group_by(year, date) |>
  summarise(y_top = max(mean_leached + se_leached, na.rm = TRUE), .groups = "drop") |>
  left_join(daily_leaching_sig, by = c("year", "date")) |>
  filter(significant) |>
  mutate(star_y = y_top * 1.1, star_label = "*")

# 4. Precipitation trimmed to leaching measurement windows
leach_date_ranges <- daily_leaching_summary |>
  group_by(year) |>
  summarise(min_date = min(date), max_date = max(date), .groups = "drop")

daily_precip_leaching <- rain |>
  select(date, year, precipmm) |>
  filter(year %in% c(2023, 2024)) |>
  left_join(leach_date_ranges, by = "year") |>
  filter(date >= min_date, date <= max_date)

max_leached_daily <- max(daily_leaching_summary$mean_leached + daily_leaching_summary$se_leached, na.rm = TRUE)
max_precip_leach <- max(daily_precip_leaching$precipmm, na.rm = TRUE)
leach_precip_scale <- (max_leached_daily * 0.4) / max_precip_leach
leach_plot_top <- max_leached_daily * 1.12
max_cum_leached <- max(cum_leaching_summary$cum_leached_mean + cum_leaching_summary$cum_leached_se, na.rm = TRUE)

# 5. New daily-difference panel data (daily means, WCC - Sorghum)
daily_leach_diff <- daily_leaching_summary |>
  mutate(trt_key = if_else(Treatment == "Sorghum + WCC", "WCC", "Sorghum")) |>
  select(year, date, trt_key, mean_leached) |>
  pivot_wider(names_from = trt_key, values_from = mean_leached) |>
  mutate(diff_wcc_minus_sorghum = WCC - Sorghum)

one_sample_vs_zero <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, mu = 0)$p.value, error = function(e) NA_real_)
}

leach_diff_stats <- daily_leach_diff |>
  group_by(year) |>
  summarise(
    n = sum(!is.na(diff_wcc_minus_sorghum)),
    mean_diff = mean(diff_wcc_minus_sorghum, na.rm = TRUE),
    p_value = one_sample_vs_zero(diff_wcc_minus_sorghum),
    significant = !is.na(p_value) & p_value < 0.05,
    letter = if_else(significant, "a", NA_character_),
    .groups = "drop"
  )

# 6. New annual panel data (combine years; one-way ANOVA by treatment)
annual_leached <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    plot = as.character(plot)
  ) |>
  group_by(year, plot, Treatment) |>
  summarise(
    annual_leached_n = (max(cumulative_n_loss_mg, na.rm = TRUE) / 1e6) / plot_area_ha,
    .groups = "drop"
  ) |>
  filter(is.finite(annual_leached_n), annual_leached_n >= 0)

anova_leached <- aov(annual_leached_n ~ Treatment, data = annual_leached)
anova_leached_p <- summary(anova_leached)[[1]]["Treatment", "Pr(>F)"]

leached_letters <- tryCatch({
  emm <- emmeans::emmeans(anova_leached, ~ Treatment)
  out <- multcomp::cld(emm, adjust = "tukey", Letters = letters) |>
    as_tibble() |>
    transmute(
      Treatment = as.character(Treatment),
      letter = gsub("\\s+", "", .group)
    )
  out
}, error = function(e) {
  annual_leached |>
    distinct(Treatment) |>
    mutate(letter = "a")
})

# 7. Plot builders for daily and cumulative leaching panels
create_leaching_daily_panel <- function(year_val, is_right_panel = FALSE, label_id) {
  d_sub <- filter(daily_leaching_summary, year == year_val)
  p_sub <- filter(daily_precip_leaching, year == year_val)
  s_sub <- filter(leaching_stars, year == year_val)
  fert_sub <- filter(fert_events, year == year_val)
  x_lims <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot() +
    geom_rect(
      data = p_sub,
      aes(
        xmin = date - 0.5,
        xmax = date + 0.5,
        ymin = leach_plot_top - precipmm * leach_precip_scale,
        ymax = leach_plot_top
      ),
      fill = "#377EB8",
      alpha = 0.7
    ) +
    geom_point(
      data = d_sub,
      aes(x = date, y = mean_leached, color = Treatment),
      size = 1.4
    ) +
    geom_errorbar(
      data = d_sub,
      aes(
        x = date,
        ymin = pmax(mean_leached - se_leached, 0),
        ymax = mean_leached + se_leached,
        color = Treatment
      ),
      width = 0.2,
      alpha = 0.5
    ) +
    geom_vline(
      data = fert_sub,
      aes(xintercept = fertilizer_date),
      linetype = "dashed",
      color = "black",
      alpha = 0.7
    ) +
    geom_text(
      data = s_sub,
      aes(x = date, y = star_y, label = star_label),
      size = 4.8
    ) +
    scale_color_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right_panel) expression("Daily Leached N (kg N ha"^-1 * " d"^-1 * ")") else "",
      limits = c(0, leach_plot_top),
      sec.axis = sec_axis(
        ~ (leach_plot_top - .) / leach_precip_scale,
        name = if (is_right_panel) "Precipitation (mm)" else ""
      )
    ) +
    labs(x = "") +
    theme_publication(base_size = 10) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.line.x = element_blank(),
      axis.text.y = if (is_right_panel) element_blank() else element_text(),
      axis.ticks.y = if (is_right_panel) element_blank() else element_line(),
      axis.line.y = if (is_right_panel) element_blank() else element_line(),
      axis.title.y = if (is_right_panel) element_blank() else element_text(margin = margin(r = 8)),
      axis.text.y.right = if (is_right_panel) element_text(color = "#377EB8") else element_blank(),
      axis.ticks.y.right = if (is_right_panel) element_line() else element_blank(),
      axis.title.y.right = if (is_right_panel) element_text(color = "#377EB8") else element_blank(),
      legend.position = "none"
    ) +
    add_panel_label(label_id)

  p
}

create_leaching_cum_panel <- function(year_val, is_right_panel = FALSE, label_id) {
  c_sub <- filter(cum_leaching_summary, year == year_val)
  fert_sub <- filter(fert_events, year == year_val)
  x_lims <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot(c_sub, aes(x = date, y = cum_leached_mean, color = Treatment, fill = Treatment)) +
    geom_ribbon(
      aes(
        ymin = pmax(cum_leached_mean - cum_leached_se, 0),
        ymax = cum_leached_mean + cum_leached_se
      ),
      alpha = 0.15,
      linewidth = 0
    ) +
    geom_line(linewidth = 1) +
    geom_vline(
      data = fert_sub,
      aes(xintercept = fertilizer_date),
      linetype = "dashed",
      color = "black",
      alpha = 0.7
    ) +
    scale_color_manual(values = treatment_colors) +
    scale_fill_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right_panel) expression("Cumulative Leached N (kg N ha"^-1 * ")") else "",
      limits = c(0, max_cum_leached)
    ) +
    labs(x = "Month") +
    theme_publication(base_size = 10) +
    theme(
      axis.text.y = if (is_right_panel) element_blank() else element_text(),
      axis.ticks.y = if (is_right_panel) element_blank() else element_line(),
      axis.line.y = if (is_right_panel) element_blank() else element_line(),
      axis.title.y = if (is_right_panel) element_blank() else element_text(margin = margin(r = 8)),
      legend.position = "none"
    ) +
    add_panel_label(label_id)

  p
}

# 8. New panel c: daily treatment difference by year
diff_span <- diff(range(daily_leach_diff$diff_wcc_minus_sorghum, na.rm = TRUE))
if (!is.finite(diff_span) || diff_span == 0) diff_span <- 0.05
leach_diff_anno <- daily_leach_diff |>
  group_by(year) |>
  summarise(y = max(diff_wcc_minus_sorghum, na.rm = TRUE), .groups = "drop") |>
  left_join(leach_diff_stats, by = "year") |>
  filter(significant) |>
  mutate(y = y + diff_span * 0.08)

p3c <- ggplot(daily_leach_diff, aes(x = factor(year), y = diff_wcc_minus_sorghum)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
  geom_jitter(
    aes(color = factor(year)),
    width = 0.12,
    alpha = 0.7,
    size = 1.4,
    show.legend = FALSE
  ) +
  geom_text(
    data = leach_diff_anno,
    aes(x = factor(year), y = y, label = letter),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.8
  ) +
  scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
  labs(
    x = "Year",
    y = expression("Daily Leached N Difference (kg N ha"^-1 * " d"^-1 * ", WCC - Sorghum)")
  ) +
  theme_publication(base_size = 10) +
    add_panel_label("c)")

# 9. New panel f: annual/cumulative by treatment with years combined + ANOVA
letter_offset <- diff(range(annual_leached$annual_leached_n, na.rm = TRUE))
if (!is.finite(letter_offset) || letter_offset == 0) letter_offset <- 0.05
annual_letter_pos <- annual_leached |>
  group_by(Treatment) |>
  summarise(y = max(annual_leached_n, na.rm = TRUE), .groups = "drop") |>
  left_join(leached_letters, by = "Treatment") |>
  mutate(y = y + letter_offset * 0.08)

p3f <- ggplot(annual_leached, aes(x = Treatment, y = annual_leached_n, fill = Treatment)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.5, color = "black") +
  geom_jitter(
    aes(shape = factor(year)),
    width = 0.12,
    size = 1.8,
    alpha = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = annual_letter_pos,
    aes(x = Treatment, y = y, label = letter),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.8
  ) +
  scale_fill_manual(values = treatment_colors) +
  labs(
    x = "",
    y = expression("Annual Leached N (kg N ha"^-1 * ")"),
    subtitle = paste0("One-way ANOVA p = ", ifelse(anova_leached_p < 0.001, "< 0.001", sprintf("%.3f", anova_leached_p)))
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "none") +
    add_panel_label("f)")

# 10. Assemble six-panel Figure 3
p3a <- create_leaching_daily_panel(2023, is_right_panel = FALSE, label_id = "a)")
p3b <- create_leaching_daily_panel(2024, is_right_panel = TRUE, label_id = "b)")
p3d <- create_leaching_cum_panel(2023, is_right_panel = FALSE, label_id = "d)")
p3e <- create_leaching_cum_panel(2024, is_right_panel = TRUE, label_id = "e)")

figure3 <- (p3a | p3b | p3c) /
  (p3d | p3e | p3f) +
  patchwork::plot_layout(guides = "collect", widths = c(1, 1, 0.95)) &
  theme(legend.position = "top", legend.justification = "center")

leach_diff_stats
anova_leached_p
figure3
```

# Explainer Figure: Why the \~90% Leaching Reduction Was Not Significant

**Scientific Context:** The cover crop treatment reduced mean annual leaching by roughly 90% in both years, but the test used log-transformed plot-level data (log10(leaching + 1)) and the sorghum treatment showed high plot-to-plot variability with only four plots per treatment. This figure shows the raw scale (large percent reduction) and the log scale used for inference (overlapping distributions and p \> 0.05).

```{r figure3-leaching-nonsig}
# Ensure constants are available
if (!exists("plot_area_ha")) {
  plot_area_ft2 <- 120 * 160
  plot_area_ha <- plot_area_ft2 / 107639.1041671
}

# Annual plot-level leaching for Sorghum vs Sorghum + WCC
annual_leaching_sr <- leaching_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(plot = as.numeric(plot)) |>
  group_by(plot, treatment, year) |>
  summarise(
    leaching_kg_ha = (max(cumulative_n_loss_mg, na.rm = TRUE) / 1e6) / plot_area_ha,
    .groups = "drop"
  ) |>
  filter(is.finite(leaching_kg_ha), leaching_kg_ha >= 0) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", "Sorghum"),
    Treatment = factor(Treatment, levels = c("Sorghum", "Sorghum + WCC"))
  )

# Percent reduction on arithmetic means
pct_reduction <- annual_leaching_sr |>
  group_by(year, Treatment) |>
  summarise(mean_leach = mean(leaching_kg_ha), .groups = "drop") |>
  tidyr::pivot_wider(names_from = Treatment, values_from = mean_leach) |>
  mutate(pct_reduction = 100 * (1 - (`Sorghum + WCC` / Sorghum)))

# p-values on the log10 scale used for inference
pvals <- annual_leaching_sr |>
  group_by(year) |>
  summarise(
    p_value = t.test(log10(leaching_kg_ha + 1) ~ Treatment)$p.value,
    .groups = "drop"
  ) |>
  mutate(
    p_label = if_else(p_value < 0.001, "p < 0.001", paste0("p = ", sprintf("%.3f", p_value)))
  )

# Plot data for raw and log scales
plot_data <- annual_leaching_sr |>
  mutate(
    scale = "Arithmetic (kg N ha^-1)",
    value = leaching_kg_ha
  ) |>
  bind_rows(
    annual_leaching_sr |>
      mutate(
        scale = "Log10(kg N ha^-1 + 1)",
        value = log10(leaching_kg_ha + 1)
      )
  )

# Mean ± SD helper (avoids Hmisc dependency)
mean_sd <- function(x) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  tibble::tibble(y = m, ymin = m - s, ymax = m + s)
}

# Annotation positions
y_positions <- plot_data |>
  group_by(year, scale) |>
  summarise(y_max = max(value, na.rm = TRUE), .groups = "drop") |>
  mutate(y = if_else(y_max == 0, 0.1, y_max * 1.15))

anno_pct <- pct_reduction |>
  left_join(
    y_positions |> filter(scale == "Arithmetic (kg N ha^-1)"),
    by = "year"
  ) |>
  mutate(
    scale = "Arithmetic (kg N ha^-1)",
    label = paste0("WCC mean is ", sprintf("%.0f", pct_reduction), "% lower"),
    x = 1.5
  )

anno_p <- pvals |>
  left_join(
    y_positions |> filter(scale == "Log10(kg N ha^-1 + 1)"),
    by = "year"
  ) |>
  mutate(
    scale = "Log10(kg N ha^-1 + 1)",
    label = p_label,
    x = 1.5
  )

fig_leach_nonsig <- ggplot(plot_data, aes(x = Treatment, y = value, color = Treatment)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.2) +
  geom_jitter(width = 0.1, size = 2.2, alpha = 0.9) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 3, fill = "white", color = "black") +
  stat_summary(
    fun.data = mean_sd,
    geom = "errorbar",
    width = 0.15,
    color = "black"
  ) +
  geom_text(data = anno_pct, aes(x = x, y = y, label = label), inherit.aes = FALSE, size = 3.2) +
  geom_text(data = anno_p, aes(x = x, y = y, label = label), inherit.aes = FALSE, size = 3.2) +
  facet_grid(scale ~ year, scales = "free_y") +
  scale_color_treatments() +
  scale_fill_treatments() +
  labs(x = "", y = NULL) +
  theme_publication(base_size = 10) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "#F2F2F2", color = "gray50")
  )

fig_leach_nonsig

ggsave(
  "figures/figure3_leaching_nonsig_explainer.png",
  fig_leach_nonsig,
  width = 7.5,
  height = 5,
  dpi = 300,
  bg = "white"
)
```

# Nitrous Oxide Emissions (Figure 4)

**Scientific Context:** Nitrous oxide (N₂O) is a potent greenhouse gas produced during soil nitrogen transformations. We measured N₂O fluxes using static chambers throughout the growing season to quantify emissions from sorghum systems with and without winter cover crops. Fluxes were generally low and episodic, with peak emissions often following fertilization or precipitation events.

```{r figure4-n2o}
# Prepare N2O flux data (Sorghum vs Sorghum + WCC)
flux_subset <- flux_data |>
  filter(treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  mutate(
    Treatment = if_else(treatment == "Sorghum + Rye", "Sorghum + WCC", treatment),
    Treatment = factor(Treatment, levels = c("Sorghum", "Sorghum + WCC")),
    year = year(date)
  ) |>
  filter(year %in% c(2023, 2024))

# Daily means in g N ha^-1 d^-1 for readability
daily_flux <- flux_subset |>
  group_by(Treatment, year, date) |>
  summarise(
    mean_flux_g = mean(gnha_day_no_negative, na.rm = TRUE),
    se_flux_g = sd(gnha_day_no_negative, na.rm = TRUE) / sqrt(sum(!is.na(gnha_day_no_negative))),
    .groups = "drop"
  )

# Date-specific treatment tests for asterisks
daily_significance <- flux_subset |>
  group_by(year, date, Treatment, plot) |>
  summarise(plot_mean_flux = mean(gnha_day_no_negative, na.rm = TRUE), .groups = "drop") |>
  group_by(year, date) |>
  summarise(
    p_value = {
      sorghum_vals <- plot_mean_flux[Treatment == "Sorghum"]
      wcc_vals <- plot_mean_flux[Treatment == "Sorghum + WCC"]
      sorghum_vals <- sorghum_vals[!is.na(sorghum_vals)]
      wcc_vals <- wcc_vals[!is.na(wcc_vals)]
      if (length(sorghum_vals) >= 3 && length(wcc_vals) >= 3) {
        tryCatch(wilcox.test(sorghum_vals, wcc_vals, exact = FALSE)$p.value, error = function(e) NA_real_)
      } else {
        NA_real_
      }
    },
    .groups = "drop"
  ) |>
  mutate(significant = !is.na(p_value) & p_value < 0.05)

star_data <- daily_flux |>
  group_by(year, date) |>
  summarise(y_top = max(mean_flux_g + se_flux_g, na.rm = TRUE), .groups = "drop") |>
  left_join(daily_significance, by = c("year", "date")) |>
  filter(significant) |>
  mutate(star_y = y_top * 1.1, star_label = "*")

# Cumulative N2O in kg N ha^-1
daily_flux_plot_level <- flux_subset |>
  group_by(Treatment, year, plot, date) |>
  summarise(daily_flux_kg = mean(gnha_day_no_negative, na.rm = TRUE) / 1000, .groups = "drop")

cum_flux_plot_level <- daily_flux_plot_level |>
  arrange(Treatment, year, plot, date) |>
  group_by(Treatment, year, plot) |>
  group_modify(~ {
    df <- .x |>
      complete(date = seq(min(date), max(date), by = "day")) |>
      arrange(date)
    non_na <- !is.na(df$daily_flux_kg)
    if (sum(non_na) == 0) return(tibble())

    flux_interp <- if (sum(non_na) >= 2) {
      approx(
        x = df$date[non_na],
        y = df$daily_flux_kg[non_na],
        xout = df$date,
        rule = 1,
        ties = "ordered"
      )$y
    } else {
      rep(df$daily_flux_kg[non_na][1], nrow(df))
    }

    df |>
      mutate(flux_interp = flux_interp, cum_flux_kg = cumsum(flux_interp)) |>
      filter(!is.na(flux_interp))
  }) |>
  ungroup()

cum_flux_summary <- cum_flux_plot_level |>
  group_by(Treatment, year, date) |>
  summarise(
    cum_flux_mean = mean(cum_flux_kg, na.rm = TRUE),
    cum_flux_sd = sd(cum_flux_kg, na.rm = TRUE),
    n_plots = n(),
    cum_flux_se = cum_flux_sd / sqrt(n_plots),
    t_crit = if_else(n_plots > 1, qt(0.975, df = n_plots - 1), NA_real_),
    cum_flux_ci = t_crit * cum_flux_se,
    .groups = "drop"
  )

# Precipitation trimmed to flux windows
flux_date_ranges <- daily_flux |>
  group_by(year) |>
  summarise(min_date = min(date), max_date = max(date), .groups = "drop")

daily_precip_trimmed <- rain |>
  select(date, year, precipmm) |>
  filter(year %in% c(2023, 2024)) |>
  left_join(flux_date_ranges, by = "year") |>
  filter(date >= min_date, date <= max_date)

# Scaling constants
global_max_flux_g <- max(daily_flux$mean_flux_g + daily_flux$se_flux_g, na.rm = TRUE)
global_max_precip <- max(daily_precip_trimmed$precipmm, na.rm = TRUE)
global_max_cum_flux <- max(cum_flux_summary$cum_flux_mean + cum_flux_summary$cum_flux_ci, na.rm = TRUE)
precip_scale_factor <- global_max_flux_g * 0.4 / global_max_precip
plot_top_flux <- global_max_flux_g * 1.12

# New panel c: daily difference (WCC - Sorghum), one-sample t-test vs 0
daily_n2o_diff <- daily_flux |>
  mutate(trt_key = if_else(Treatment == "Sorghum + WCC", "WCC", "Sorghum")) |>
  select(year, date, trt_key, mean_flux_g) |>
  pivot_wider(names_from = trt_key, values_from = mean_flux_g) |>
  mutate(diff_wcc_minus_sorghum = WCC - Sorghum)

one_sample_vs_zero <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, mu = 0)$p.value, error = function(e) NA_real_)
}

n2o_diff_stats <- daily_n2o_diff |>
  group_by(year) |>
  summarise(
    n = sum(!is.na(diff_wcc_minus_sorghum)),
    mean_diff = mean(diff_wcc_minus_sorghum, na.rm = TRUE),
    p_value = one_sample_vs_zero(diff_wcc_minus_sorghum),
    significant = !is.na(p_value) & p_value < 0.05,
    letter = if_else(significant, "a", NA_character_),
    .groups = "drop"
  )

# New panel f: annual cumulative N2O by treatment (combine years) + one-way ANOVA
annual_n2o <- cum_flux_plot_level |>
  group_by(year, plot, Treatment) |>
  summarise(annual_cum_n2o = max(cum_flux_kg, na.rm = TRUE), .groups = "drop") |>
  filter(is.finite(annual_cum_n2o), annual_cum_n2o >= 0)

anova_n2o <- aov(annual_cum_n2o ~ Treatment, data = annual_n2o)
anova_n2o_p <- summary(anova_n2o)[[1]]["Treatment", "Pr(>F)"]

n2o_letters <- tryCatch({
  emm <- emmeans::emmeans(anova_n2o, ~ Treatment)
  out <- multcomp::cld(emm, adjust = "tukey", Letters = letters) |>
    as_tibble() |>
    transmute(
      Treatment = as.character(Treatment),
      letter = gsub("\\s+", "", .group)
    )
  out
}, error = function(e) {
  annual_n2o |>
    distinct(Treatment) |>
    mutate(letter = "a")
})

# Daily and cumulative panel builders
create_n2o_daily_panel <- function(year_val, is_right_panel = FALSE, label_id) {
  flux_year <- filter(daily_flux, year == year_val)
  precip_year <- filter(daily_precip_trimmed, year == year_val)
  star_year <- filter(star_data, year == year_val)
  fert_year <- filter(fert_events, year == year_val)
  x_lims <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot() +
    geom_rect(
      data = precip_year,
      aes(
        xmin = date - 0.5,
        xmax = date + 0.5,
        ymin = plot_top_flux - precipmm * precip_scale_factor,
        ymax = plot_top_flux
      ),
      fill = "#377EB8",
      alpha = 0.7
    ) +
    geom_point(
      data = flux_year,
      aes(x = date, y = mean_flux_g, color = Treatment),
      size = 1.4
    ) +
    geom_errorbar(
      data = flux_year,
      aes(x = date, ymin = pmax(mean_flux_g - se_flux_g, 0), ymax = mean_flux_g + se_flux_g, color = Treatment),
      width = 0.2,
      alpha = 0.5
    ) +
    geom_vline(
      data = fert_year,
      aes(xintercept = fertilizer_date),
      linetype = "dashed",
      color = "black",
      alpha = 0.7
    ) +
    geom_text(
      data = star_year,
      aes(x = date, y = star_y, label = star_label),
      size = 4.8
    ) +
    scale_color_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right_panel) expression("Daily N"[2] * "O Flux (g N ha"^-1 * " d"^-1 * ")") else "",
      limits = c(0, plot_top_flux),
      sec.axis = sec_axis(
        ~ (plot_top_flux - .) / precip_scale_factor,
        name = if (is_right_panel) "Precipitation (mm)" else ""
      )
    ) +
    labs(x = "") +
    theme_publication(base_size = 10) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.line.x = element_blank(),
      axis.text.y = if (is_right_panel) element_blank() else element_text(),
      axis.ticks.y = if (is_right_panel) element_blank() else element_line(),
      axis.line.y = if (is_right_panel) element_blank() else element_line(),
      axis.title.y = if (is_right_panel) element_blank() else element_text(margin = margin(r = 8)),
      axis.text.y.right = if (is_right_panel) element_text(color = "#377EB8") else element_blank(),
      axis.ticks.y.right = if (is_right_panel) element_line() else element_blank(),
      axis.title.y.right = if (is_right_panel) element_text(color = "#377EB8") else element_blank(),
      legend.position = "none"
    ) +
    add_panel_label(label_id)

  p
}

create_n2o_cum_panel <- function(year_val, is_right_panel = FALSE, label_id) {
  cum_flux_year <- filter(cum_flux_summary, year == year_val)
  fert_year <- filter(fert_events, year == year_val)
  x_lims <- as.Date(paste0(year_val, c("-01-01", "-12-31")))

  p <- ggplot(cum_flux_year, aes(x = date, y = cum_flux_mean, color = Treatment, fill = Treatment)) +
    geom_ribbon(
      aes(
        ymin = pmax(cum_flux_mean - cum_flux_ci, 0),
        ymax = cum_flux_mean + cum_flux_ci
      ),
      alpha = 0.15,
      linewidth = 0
    ) +
    geom_line(linewidth = 1) +
    geom_vline(
      data = fert_year,
      aes(xintercept = fertilizer_date),
      linetype = "dashed",
      color = "black",
      alpha = 0.7
    ) +
    scale_color_manual(values = treatment_colors) +
    scale_fill_manual(values = treatment_colors) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b", limits = x_lims) +
    scale_y_continuous(
      name = if (!is_right_panel) expression("Cumulative N"[2] * "O Flux (kg N ha"^-1 * ")") else "",
      limits = c(0, global_max_cum_flux)
    ) +
    labs(x = "Month") +
    theme_publication(base_size = 10) +
    theme(
      axis.text.y = if (is_right_panel) element_blank() else element_text(),
      axis.ticks.y = if (is_right_panel) element_blank() else element_line(),
      axis.line.y = if (is_right_panel) element_blank() else element_line(),
      axis.title.y = if (is_right_panel) element_blank() else element_text(margin = margin(r = 8)),
      legend.position = "none"
    ) +
    add_panel_label(label_id)

  p
}

# Difference and annual summary panels
n2o_diff_span <- diff(range(daily_n2o_diff$diff_wcc_minus_sorghum, na.rm = TRUE))
if (!is.finite(n2o_diff_span) || n2o_diff_span == 0) n2o_diff_span <- 1
n2o_diff_anno <- daily_n2o_diff |>
  group_by(year) |>
  summarise(y = max(diff_wcc_minus_sorghum, na.rm = TRUE), .groups = "drop") |>
  left_join(n2o_diff_stats, by = "year") |>
  filter(significant) |>
  mutate(y = y + n2o_diff_span * 0.08)

p4c <- ggplot(daily_n2o_diff, aes(x = factor(year), y = diff_wcc_minus_sorghum)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "grey92", color = "black") +
  geom_jitter(
    aes(color = factor(year)),
    width = 0.12,
    alpha = 0.7,
    size = 1.4,
    show.legend = FALSE
  ) +
  geom_text(
    data = n2o_diff_anno,
    aes(x = factor(year), y = y, label = letter),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.8
  ) +
  scale_color_manual(values = c("2023" = "#E41A1C", "2024" = "#377EB8")) +
  labs(
    x = "Year",
    y = expression("Daily N"[2] * "O Difference (g N ha"^-1 * " d"^-1 * ", WCC - Sorghum)")
  ) +
  theme_publication(base_size = 10) +
    add_panel_label("c)")

n2o_letter_offset <- diff(range(annual_n2o$annual_cum_n2o, na.rm = TRUE))
if (!is.finite(n2o_letter_offset) || n2o_letter_offset == 0) n2o_letter_offset <- 0.05
n2o_letter_pos <- annual_n2o |>
  group_by(Treatment) |>
  summarise(y = max(annual_cum_n2o, na.rm = TRUE), .groups = "drop") |>
  left_join(n2o_letters, by = "Treatment") |>
  mutate(y = y + n2o_letter_offset * 0.08)

p4f <- ggplot(annual_n2o, aes(x = Treatment, y = annual_cum_n2o, fill = Treatment)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.5, color = "black") +
  geom_jitter(
    aes(shape = factor(year)),
    width = 0.12,
    size = 1.8,
    alpha = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = n2o_letter_pos,
    aes(x = Treatment, y = y, label = letter),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.8
  ) +
  scale_fill_manual(values = treatment_colors) +
  labs(
    x = "",
    y = expression("Annual Cumulative N"[2] * "O Flux (kg N ha"^-1 * ")"),
    subtitle = paste0("One-way ANOVA p = ", ifelse(anova_n2o_p < 0.001, "< 0.001", sprintf("%.3f", anova_n2o_p)))
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "none") +
    add_panel_label("f)")

# Assemble six-panel Figure 4
p4a <- create_n2o_daily_panel(2023, is_right_panel = FALSE, label_id = "a)")
p4b <- create_n2o_daily_panel(2024, is_right_panel = TRUE, label_id = "b)")
p4d <- create_n2o_cum_panel(2023, is_right_panel = FALSE, label_id = "d)")
p4e <- create_n2o_cum_panel(2024, is_right_panel = TRUE, label_id = "e)")

figure4 <- (p4a | p4b | p4c) /
  (p4d | p4e | p4f) +
  patchwork::plot_layout(guides = "collect", widths = c(1, 1, 0.95)) &
  theme(legend.position = "top", legend.justification = "center")

n2o_diff_stats
anova_n2o_p
figure4
```

# Biomass Yields (Figure 5)

**Scientific Context:** A critical question for cover crop adoption is whether the additional biomass production impacts the primary cash crop yield. We measured end-of-season biomass for both sorghum and cereal rye to determine if the cover crop caused a "yield drag" on sorghum production.

**Note:** The sorghum biomass source files are stored in the researcher's Box folder and not included in this repository. The code below demonstrates the analysis workflow. Users with access to the complete dataset can update the file paths accordingly.
Biomass is reported as oven-dried (0% moisture) equivalent dry matter.

```{r figure5-biomass}
sorghum_2023_path <- "/Users/jaredflater/Library/CloudStorage/Box-Box/CABBI/Data/MeasuredData/SABR/SmallDrainagePlots/Biomass/2023GrowingSeason/DerivedDatasets/chopper_load_weights_yields.xlsx"
sorghum_2024_path <- "/Users/jaredflater/Library/CloudStorage/Box-Box/CABBI/Data/MeasuredData/SABR/SmallDrainagePlots/Biomass/2024GrowingSeason/RecordingDataSheets/sorg_chopper_harvest_load_weights.xlsx"

plot_size_ha <- 0.145
conversion_factor <- 0.907185
treatment_levels <- c("Sorghum", "Sorghum + WCC")

# Moisture-adjustment factor used to report dry-equivalent sorghum biomass
moisture_df <- tibble(
  Plot = c(2, 7, 9, 15),
  Moisture = c(0.749262537, 0.766325316, 0.731502423, 0.767241379)
)
avg_moisture <- mean(moisture_df$Moisture, na.rm = TRUE)

sorghum_2023 <- read_excel(sorghum_2023_path) |>
  group_by(Plot) |>
  summarise(total_biomass_ton = sum(`Net Weight`, na.rm = TRUE), .groups = "drop") |>
  mutate(
    year = 2023,
    Plot = as.character(Plot),
    treatment = case_when(
      Plot %in% c("2", "3", "5", "15") ~ "Sorghum",
      TRUE ~ "Sorghum + WCC"
    ),
    biomass_mg_ha = ((total_biomass_ton * conversion_factor) / plot_size_ha) * (1 - avg_moisture)
  ) |>
  select(year, Plot, treatment, biomass_mg_ha)

sorghum_2024 <- read_excel(sorghum_2024_path) |>
  filter(Study == "Drainage") |>
  group_by(Plot) |>
  summarise(total_biomass_ton = sum(`Net Weight (t)`, na.rm = TRUE), .groups = "drop") |>
  mutate(
    year = 2024,
    Plot = as.character(Plot),
    treatment = case_when(
      Plot %in% c("2", "3", "5", "15") ~ "Sorghum",
      TRUE ~ "Sorghum + WCC"
    ),
    biomass_mg_ha = ((total_biomass_ton * conversion_factor) / plot_size_ha) * (1 - avg_moisture)
  ) |>
  select(year, Plot, treatment, biomass_mg_ha)

sorghum_plot_data <- bind_rows(sorghum_2023, sorghum_2024) |>
  mutate(
    year = factor(year, levels = c(2023, 2024)),
    treatment = factor(treatment, levels = treatment_levels)
  )

rye_data <- bind_rows(
  read_excel("data/BiomassData_Rye_20230522.xlsx") |> mutate(year = 2023),
  read_excel("data/BiomassData_Rye_20240515.xlsx") |> mutate(year = 2024)
) |>
  mutate(
    year = factor(year, levels = c(2023, 2024)),
    Plot = as.character(`Plot ID`),
    treatment = factor("Sorghum + WCC", levels = treatment_levels),
    biomass_mg_ha = (`Dry Mass + Dry Bag Mass` - `Dry Bag Mass`) * 0.01
  ) |>
  select(year, Plot, treatment, biomass_mg_ha)

sorghum_stats <- sorghum_plot_data |>
  group_by(year, treatment) |>
  summarise(
    mean_sorghum = mean(biomass_mg_ha, na.rm = TRUE),
    sd_sorghum = sd(biomass_mg_ha, na.rm = TRUE),
    n = n(),
    se_sorghum = sd_sorghum / sqrt(n),
    ci_sorghum = qt(0.975, df = pmax(n - 1, 1)) * se_sorghum,
    .groups = "drop"
  )

rye_stats <- rye_data |>
  group_by(year, treatment) |>
  summarise(
    mean_rye = mean(biomass_mg_ha, na.rm = TRUE),
    sd_rye = sd(biomass_mg_ha, na.rm = TRUE),
    n = n(),
    se_rye = sd_rye / sqrt(n),
    ci_rye = qt(0.975, df = pmax(n - 1, 1)) * se_rye,
    .groups = "drop"
  )

sorghum_tests_by_year <- sorghum_plot_data |>
  group_by(year) |>
  summarise(
    p_value = {
      if (n_distinct(treatment) < 2) {
        NA_real_
      } else {
        tryCatch(t.test(biomass_mg_ha ~ treatment)$p.value, error = function(e) NA_real_)
      }
    },
    .groups = "drop"
  )

sorghum_letters <- sorghum_stats |>
  left_join(sorghum_tests_by_year, by = "year") |>
  group_by(year) |>
  mutate(
    letter = case_when(
      is.na(p_value) ~ "a",
      p_value >= 0.05 ~ "a",
      mean_sorghum == max(mean_sorghum, na.rm = TRUE) ~ "a",
      TRUE ~ "b"
    )
  ) |>
  ungroup()

letter_offsets <- sorghum_letters |>
  group_by(year) |>
  summarise(offset = max(mean_sorghum + ci_sorghum, na.rm = TRUE) * 0.08, .groups = "drop")

sorghum_letters <- sorghum_letters |>
  left_join(letter_offsets, by = "year") |>
  mutate(y_label = mean_sorghum + ci_sorghum + offset)

sorghum_stats <- sorghum_stats |>
  mutate(fill_key = as.character(treatment))

rye_overlay <- rye_stats |>
  mutate(fill_key = "Rye overlay")

fill_palette <- c(
  "Sorghum" = "#D8D97AFF",
  "Sorghum + WCC" = "#95C36EFF",
  "Rye overlay" = "#F5DEB3"
)

rye_layer <- if (requireNamespace("ggpattern", quietly = TRUE)) {
  ggpattern::geom_col_pattern(
    data = rye_overlay,
    aes(x = treatment, y = mean_rye, fill = fill_key),
    pattern = "stripe",
    pattern_fill = "grey35",
    pattern_colour = "grey35",
    pattern_angle = 45,
    pattern_density = 0.45,
    alpha = 0.95,
    color = "black",
    width = 0.32,
    inherit.aes = FALSE
  )
} else {
  geom_col(
    data = rye_overlay,
    aes(x = treatment, y = mean_rye, fill = fill_key),
    alpha = 0.95,
    color = "black",
    width = 0.32,
    inherit.aes = FALSE
  )
}

facet_panel_labels <- tibble(
  year = factor(c(2023, 2024), levels = c(2023, 2024)),
  treatment = factor("Sorghum", levels = treatment_levels),
  y = Inf,
  label = c("a)", "b)")
)

final_biomass_plot <- ggplot() +
  geom_col(
    data = sorghum_stats,
    aes(x = treatment, y = mean_sorghum, fill = fill_key),
    color = "black",
    width = 0.7
  ) +
  geom_errorbar(
    data = sorghum_stats,
    aes(
      x = treatment,
      ymin = pmax(0, mean_sorghum - ci_sorghum),
      ymax = mean_sorghum + ci_sorghum
    ),
    width = 0.18,
    inherit.aes = FALSE
  ) +
  rye_layer +
  geom_errorbar(
    data = rye_overlay,
    aes(
      x = treatment,
      ymin = pmax(0, mean_rye - ci_rye),
      ymax = mean_rye + ci_rye
    ),
    width = 0.1,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = sorghum_letters,
    aes(x = treatment, y = y_label, label = letter),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.8
  ) +
  geom_text(
    data = facet_panel_labels,
    aes(x = treatment, y = y, label = label),
    inherit.aes = FALSE,
    hjust = panel_label_hjust,
    vjust = panel_label_vjust,
    fontface = panel_label_face,
    size = panel_label_size
  ) +
  facet_wrap(~year, nrow = 1) +
  scale_fill_manual(
    values = fill_palette,
    breaks = c("Sorghum", "Sorghum + WCC", "Rye overlay"),
    name = "Biomass component"
  ) +
  labs(
    x = "",
    y = expression("Oven-dried Biomass (Mg ha"^-1 * ")")
  ) +
  theme_publication(base_size = 10) +
  theme(
    legend.position = "top",
    legend.justification = "center"
  )

print(final_biomass_plot)
ggsave(
  "SubmissionFiguresTables/final_biomass_stacked.png",
  final_biomass_plot,
  width = 174,
  height = 174,
  units = "mm",
  dpi = 600,
  bg = "white"
)

sorghum_tests_by_year
figure5 <- final_biomass_plot
```

# Environmental Damage Costs (Table 3)

**Scientific Context:** To evaluate the broader environmental implications of nitrogen losses, we calculated the "Environmental Damage Cost" (EDC) by monetizing N₂O emissions and nitrate leaching using damage cost values from Preza-Fontes et al. This provides an economic perspective on the environmental benefits of cover cropping.

```{r table3-environmental-costs}
# Prepare N2O plot-level data (averaging row/interrow)
# First, create interpolated flux data for annual totals
interpolate_daily_flux <- function(data) {
  plot_combinations <- data |>
    distinct(plot, rowvs_interrow, treatment, year)
  
  purrr::map_dfr(seq_len(nrow(plot_combinations)), function(i) {
    combo <- plot_combinations[i, ]
    
    subset_data <- data |>
      filter(
        plot == combo$plot,
        rowvs_interrow == combo$rowvs_interrow,
        treatment == combo$treatment,
        year == combo$year
      ) |>
      arrange(date)
    
    start_date <- as.Date(paste0(combo$year, "-01-01"))
    end_date <- as.Date(paste0(combo$year, "-12-31"))
    all_days <- seq(start_date, end_date, by = "day")
    
    if (nrow(subset_data) >= 2 && !all(is.na(subset_data$mean_flux))) {
      valid_data <- subset_data[!is.na(subset_data$mean_flux), ]
      
      if (nrow(valid_data) >= 2) {
        interpolated_flux <- approx(
          x = as.numeric(valid_data$date),
          y = valid_data$mean_flux,
          xout = as.numeric(all_days),
          rule = 1
        )$y
      } else {
        interpolated_flux <- rep(NA_real_, length(all_days))
      }
    } else {
      interpolated_flux <- rep(NA_real_, length(all_days))
    }
    
    tibble(
      plot = combo$plot,
      rowvs_interrow = combo$rowvs_interrow,
      treatment = combo$treatment,
      year = combo$year,
      date = all_days,
      interpolated_flux = interpolated_flux
    )
  })
}

# Summarize daily flux by plot/date
daily_flux_summary <- flux_data |>
  mutate(plot = as.numeric(plot)) |>
  group_by(plot, rowvs_interrow, treatment, year, date) |>
  summarise(mean_flux = mean(gnha_day_no_negative, na.rm = TRUE), .groups = "drop") |>
  arrange(plot, rowvs_interrow, treatment, year, date)

# Apply interpolation
interpolated_flux_data <- interpolate_daily_flux(daily_flux_summary)

# Calculate plot-level N2O (average row/interrow, then sum for year)
n2o_plot <- interpolated_flux_data |>
  mutate(plot = as.numeric(plot)) |>
  group_by(plot, treatment, year, date) |>
  summarise(daily_plot_flux = mean(interpolated_flux, na.rm = TRUE), .groups = "drop") |>
  group_by(plot, treatment, year) |>
  summarise(
    n2o_kg_ha = sum(daily_plot_flux, na.rm = TRUE) / 1000, 
    .groups = "drop"
  )

# Prepare leaching plot-level data
leach_plot <- leaching_data |>
  mutate(plot = as.numeric(plot)) |>
  group_by(plot, treatment, year) |>
  summarise(
    leaching_kg_ha = (max(cumulative_n_loss_mg, na.rm = TRUE) / 1e6) / plot_area_ha,
    .groups = "drop"
  ) |>
  filter(is.finite(leaching_kg_ha), leaching_kg_ha >= 0)

# Join N2O and leaching, calculate EDC
plot_economics <- n2o_plot |>
  inner_join(leach_plot, by = c("year", "plot", "treatment")) |>
  mutate(edc_dollars = (n2o_kg_ha * 16.18) + (leaching_kg_ha * 18.54))

# Calculate 2-year totals for continuous systems (Sorghum treatments)
continuous_2year <- plot_economics |>
  filter(year %in% c(2023, 2024), treatment %in% c("Sorghum", "Sorghum + Rye")) |>
  group_by(plot, treatment) |>
  summarise(
    n2o_kg_ha = sum(n2o_kg_ha, na.rm = TRUE),
    leaching_kg_ha = sum(leaching_kg_ha, na.rm = TRUE),
    edc_dollars = sum(edc_dollars, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  ) |>
  filter(count == 2) |>
  select(-count)

# Calculate 2-year statistics for rotation systems (Corn, Soy)
rotation_annual_stats <- plot_economics |>
  filter(year %in% c(2023, 2024), treatment %in% c("Corn", "Soy")) |>
  group_by(treatment, year) |>
  summarise(
    mu_n2o = mean(n2o_kg_ha, na.rm = TRUE), var_n2o = var(n2o_kg_ha, na.rm = TRUE),
    mu_leach = mean(leaching_kg_ha, na.rm = TRUE), var_leach = var(leaching_kg_ha, na.rm = TRUE),
    mu_edc = mean(edc_dollars, na.rm = TRUE), var_edc = var(edc_dollars, na.rm = TRUE),
    .groups = "drop"
  )

rotation_2year_stats <- rotation_annual_stats |>
  group_by(treatment) |>
  summarise(
    final_mu_n2o = sum(mu_n2o), final_sd_n2o = sqrt(sum(var_n2o)),
    final_mu_leach = sum(mu_leach), final_sd_leach = sqrt(sum(var_leach)),
    final_mu_edc = sum(mu_edc), final_sd_edc = sqrt(sum(var_edc)),
    .groups = "drop"
  )

# Create synthetic plots for rotation systems
rotation_synthetic_plots <- rotation_2year_stats |>
  group_by(treatment) |>
  group_modify(~ {
    n_sim <- 3
    make_dist <- function(n, mu, sd) {
      x <- rnorm(n)
      if(sd > 0) x_final <- ((x - mean(x)) / sd(x)) * sd + mu else x_final <- rep(mu, n)
      return(pmax(0, x_final))
    }
    tibble(
      plot = paste0("Syn_", 1:n_sim),
      n2o_kg_ha = make_dist(n_sim, .x$final_mu_n2o, .x$final_sd_n2o),
      leaching_kg_ha = make_dist(n_sim, .x$final_mu_leach, .x$final_sd_leach),
      edc_dollars = make_dist(n_sim, .x$final_mu_edc, .x$final_sd_edc)
    )
  }) |>
  mutate(plot = as.numeric(plot)) |>
  ungroup()

# Combine real + synthetic
final_2year_data <- bind_rows(continuous_2year, rotation_synthetic_plots) |>
  mutate(year = "2023-2024")

# Master dataset for statistics
stats_input_data <- bind_rows(
  plot_economics |> mutate(year = as.character(year)),
  final_2year_data
)

# Statistical function with log transformation where needed
calc_stat_column <- function(data, col_name, is_money = FALSE) {
  # Use log transformation for leaching and costs
  if (col_name %in% c("leaching_kg_ha", "edc_dollars")) {
    f <- as.formula(paste("log10(", col_name, " + 1) ~ treatment"))
  } else {
    f <- as.formula(paste(col_name, "~ treatment"))
  }
  
  fit <- lm(f, data = data)
  
  # Get letters based on transformed model
  lets <- tryCatch({
    emmeans(fit, ~ treatment) |>
      cld(adjust = "tukey", Letters = letters, alpha = 0.05) |>
      as_tibble() |>
      mutate(letter = trimws(.group)) |>
      select(treatment, letter)
  }, error = function(e) tibble(treatment = unique(data$treatment), letter = ""))
  
  # Calculate arithmetic mean/SD for display
  data |>
    group_by(treatment) |>
    summarise(mean = mean(.data[[col_name]], na.rm = TRUE), 
              sd = sd(.data[[col_name]], na.rm = TRUE), .groups = "drop") |>
    left_join(lets, by = "treatment") |>
    mutate(
      mean_str = formatC(signif(mean, 3), digits = 3, format = "fg", flag = "#"),
      sd_str = formatC(signif(sd, 3), digits = 3, format = "fg", flag = "#"),
      mean_str = sub("\\.$", "", mean_str),
      sd_str = sub("\\.$", "", sd_str),
      result_str = if(is_money) 
        sprintf("$%s ± %s (%s)", mean_str, sd_str, letter) 
      else 
        sprintf("%s ± %s (%s)", mean_str, sd_str, letter)
    ) |>
    select(treatment, result_str)
}

# Generate table body
table_body <- stats_input_data |>
  group_by(year) |>
  group_modify(~ {
    n2o <- calc_stat_column(.x, "n2o_kg_ha")
    leach <- calc_stat_column(.x, "leaching_kg_ha")
    edc <- calc_stat_column(.x, "edc_dollars", is_money = TRUE)
    n2o |> rename(N2O = result_str) |>
      left_join(leach |> rename(Leach = result_str), by = "treatment") |>
      left_join(edc |> rename(Cost = result_str), by = "treatment")
  }) |>
  ungroup()

# Source of variation (ANOVA p-values)
aov_n2o <- summary(aov(n2o_kg_ha ~ as.factor(year) * treatment, data = plot_economics))[[1]]
aov_leach <- summary(aov(log10(leaching_kg_ha+1) ~ as.factor(year) * treatment, data = plot_economics))[[1]]
aov_edc <- summary(aov(log10(edc_dollars+1) ~ as.factor(year) * treatment, data = plot_economics))[[1]]

format_p <- function(p) ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))

source_var <- tibble(
  year = "Source of Variation",
  treatment = c("Treatment", "Year", "Year × Treatment"),
  N2O = c(format_p(aov_n2o[2, "Pr(>F)"]), format_p(aov_n2o[1, "Pr(>F)"]), format_p(aov_n2o[3, "Pr(>F)"])),
  Leach = c(format_p(aov_leach[2, "Pr(>F)"]), format_p(aov_leach[1, "Pr(>F)"]), format_p(aov_leach[3, "Pr(>F)"])),
  Cost = c(format_p(aov_edc[2, "Pr(>F)"]), format_p(aov_edc[1, "Pr(>F)"]), format_p(aov_edc[3, "Pr(>F)"]))
)

# Finalize table
trt_order <- c("Corn", "Soy", "Sorghum", "Sorghum + Rye")
final_table <- bind_rows(table_body, source_var) |>
  mutate(
    order_helper = case_when(
      year == "2023" ~ 1, year == "2024" ~ 2,
      year == "2023-2024" ~ 3, year == "Source of Variation" ~ 4
    ),
    treatment_factor = factor(treatment, levels = c(trt_order, "Treatment", "Year", "Year × Treatment"))
  ) |>
  arrange(order_helper, treatment_factor) |>
  select(Year = year, Treatment = treatment, `N2O Loss` = N2O, `N Leaching` = Leach, `Env. Damage Cost` = Cost)

# Create GT table
table3 <- final_table |>
  gt() |>
  tab_header(title = md("**Table 3. Annual nitrogen losses and environmental costs**")) |>
  cols_label(
    `N2O Loss` = md("N<sub>2</sub>O Loss<br>(kg N ha<sup>-1</sup>)"),
    `N Leaching` = md("N Leaching<br>(kg N ha<sup>-1</sup>)"),
    `Env. Damage Cost` = md("Total Environmental Cost<br>(USD ha<sup>-1</sup>)")
  ) |>
  cols_align(align = "center", columns = -Treatment) |>
  tab_style(
    style = cell_borders(sides = "top", color = "gray", weight = px(2)),
    locations = cells_body(rows = Year == "Source of Variation")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = Year == "Source of Variation")
  ) |>
  tab_footnote(
    footnote = "Values are arithmetic mean ± SD (3 significant figures). Letters indicate Tukey-adjusted groups derived from log-transformed data for leaching/costs and non-transformed data for N₂O (α=0.05). Costs based on Preza-Fontes et al.: N₂O ($16.18/kg N) + Leaching ($18.54/kg N). Two-year totals for Corn/Soy estimated using variance propagation of annual crop phases."
  )

table3
```

# Summary

This reproducible companion document has walked through the complete analytical workflow for the manuscript "Integrating Winter Cover Crops With Biomass Sorghum Can Reduce Nitrogen Losses." The code presented here generates all figures and tables in the published paper, providing full transparency and enabling reproduction of our findings.

## Key Findings Demonstrated:

1.  **Weather variability** between study years (2023 dry, 2024 near-normal)
2.  **Soil microclimate effects** of cover crops on temperature and moisture
3.  **No nitrogen immobilization** detected in soil inorganic N pools
4.  **Substantial reduction** in nitrate leaching with cover crops
5.  **Low N₂O emissions** with no consistent treatment effect
6.  **No yield penalty** from cover crop integration
7.  **Significant economic value** from reduced environmental damage costs

All data processing steps, statistical analyses, and visualizations are shown explicitly, allowing readers to understand exactly how results were derived and to adapt methods for their own research.

------------------------------------------------------------------------

*Document generated: `r Sys.Date()`*

*R version: `r R.version.string`*

```{r export-figures-tables}
#| echo: false
#| message: false
#| warning: false
#| results: asis

# Make HTML figures render larger in the output document.
if (knitr::is_html_output()) {
  cat(
    "<style>
      .cell-output-display img { max-width: 100% !important; width: 100% !important; height: auto; }
    </style>"
  )
}

# Save all ggplot/patchwork figures and gt tables at one-page size.
page_width_in <- 11
page_height_in <- 8.5
dpi <- 300

dir.create("figures", showWarnings = FALSE)
dir.create("tables", showWarnings = FALSE)

objs <- mget(ls(envir = .GlobalEnv), envir = .GlobalEnv)

safe_name <- function(x) gsub("[^A-Za-z0-9_]+", "_", x)

plot_names <- names(objs)[vapply(objs, function(x) inherits(x, c("ggplot", "patchwork")), logical(1))]
for (nm in plot_names) {
  out_file <- file.path("figures", paste0("figure_", safe_name(nm), ".png"))
  ggplot2::ggsave(
    filename = out_file,
    plot = objs[[nm]],
    width = page_width_in,
    height = page_height_in,
    units = "in",
    dpi = dpi,
    limitsize = FALSE
  )
}

table_names <- names(objs)[vapply(objs, function(x) inherits(x, "gt_tbl"), logical(1))]
for (nm in table_names) {
  out_file <- file.path("tables", paste0("table_", safe_name(nm), ".png"))
  gt::gtsave(
    data = objs[[nm]],
    filename = out_file,
    vwidth = page_width_in * dpi,
    vheight = page_height_in * dpi
  )
}
```
