---
title: "Table S2: Cereal Rye Cover Crop Biomass"
author: "Jared Flater et al."
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
execute:
  warning: false
  message: false
  echo: true
---

# Overview

Summary statistics for cereal rye (cover crop) dry biomass at termination,
reported in Mg ha⁻¹ (oven-dry equivalent).

**Unit conversion note:** Samples were collected from 1 m² quadrats (4 reps per
plot, 4 plots per year → n = 16 per year). Dry mass is recorded in grams. The
conversion from g m⁻² to Mg ha⁻¹ is:

$$1 \text{ g m}^{-2} \times \frac{1 \text{ Mg}}{10^6 \text{ g}} \times
\frac{10{,}000 \text{ m}^2}{\text{ha}} = 0.01 \text{ Mg ha}^{-1}$$

So `biomass_mg_ha = net_dry_mass_g * 0.01` is the correct conversion.

# Setup

```{r setup}
library(tidyverse)
library(readxl)
library(gt)
```

# Load & Process Data

```{r load-data}
rye_biomass <- bind_rows(
  read_excel("data/BiomassData_Rye_20230522.xlsx") |> mutate(year = 2023),
  read_excel("data/BiomassData_Rye_20240515.xlsx") |> mutate(year = 2024)
) |>
  mutate(
    year          = factor(year, levels = c(2023, 2024)),
    Plot          = as.character(`Plot ID`),
    # Samples from 1 m² quadrats; dry mass in grams → Mg ha⁻¹
    # g/m² × 0.01 = Mg/ha
    biomass_mg_ha = (`Dry Mass + Dry Bag Mass` - `Dry Bag Mass`) * 0.01
  ) |>
  select(year, Plot, biomass_mg_ha)
```

# Raw Data Check

```{r raw-check}
rye_biomass |>
  group_by(year) |>
  summarise(
    n         = n(),
    mean_g_m2 = mean(biomass_mg_ha / 0.01),   # back-convert to g/m² for inspection
    mean_Mg_ha = mean(biomass_mg_ha),
    min_Mg_ha = min(biomass_mg_ha),
    max_Mg_ha = max(biomass_mg_ha)
  )
```

# Summary Statistics

```{r summary-stats}
rye_biomass_summary <- rye_biomass |>
  group_by(year) |>
  summarise(
    n     = n(),
    mean  = mean(biomass_mg_ha, na.rm = TRUE),
    sd    = sd(biomass_mg_ha, na.rm = TRUE),
    se    = sd / sqrt(n),
    min   = min(biomass_mg_ha, na.rm = TRUE),
    max   = max(biomass_mg_ha, na.rm = TRUE),
    .groups = "drop"
  )

rye_biomass_summary
```

# Table S2

```{r table-s2}
table_s2 <- rye_biomass_summary |>
  gt() |>
  cols_label(
    year = "Year",
    n    = "n",
    mean = "Mean",
    sd   = "SD",
    se   = "SE",
    min  = "Min",
    max  = "Max"
  ) |>
  fmt_number(columns = c(mean, sd, se, min, max), decimals = 2) |>
  tab_header(
    title    = md("**Table S2. Cereal Rye (Cover Crop) Biomass**"),
    subtitle = md("Biomass in Mg ha⁻¹, measured at termination")
  ) |>
  cols_align(align = "center", columns = everything())

table_s2
```

# Save

```{r save}
gtsave(table_s2, "SubmissionFiguresTables/Table_S2_Cover_Crop_Biomass.rtf")
```
